---
title: CRAN updates
author: Lluís Revilla Sancho
date: '2022-05-28'
slug: cran-updates
categories:
  - CRAN
  - r
tags:
  - R
authors: []
description: ''
draft: no
editor_options:
  chunk_output_type: console
featured: no
image:
  caption: ''
  focal_point: ''
subtitle: ''
summary: ''
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.retina = 2)
```


There are many great things in base R, one of them is the tools package. 
Looking into it I found some internal functions that access freely available files with information about CRAN packages.
These private functions are at the CRANtools.R file.

```{r}
# issues <- tools:::CRAN_check_issues()
packages <- tools:::CRAN_package_db()
# alias <- tools:::CRAN_aliases_db()
archive <- tools:::CRAN_archive_db()
current <- tools:::CRAN_current_db()
# rdxrefs <- tools:::CRAN_rdxrefs_db()
```


```{r}
packages$Published <- as.Date(packages$Published, format = "%Y-%m-%d", tz = "UTC")
packages$Packaged_by <- gsub(".*; (.*)", "\\1", packages$Packaged)
packages$Packaged <- as.POSIXct(packages$Packaged, 
                                format = "%Y-%m-%d %H:%M:%S", tz = "UTC")

p <- packages |> 
  mutate(p =  as.Date(Packaged),
         diff = round(difftime(p, Published, units = "days"), 0)) |> 
  mutate(with_archives = Package %in% names(archive)) |> 
  filter(diff <= 0) |> # 3 Packages built after published!
  mutate(diff = abs(diff)) |> 
  filter(diff <= 50) # Few packages waited several months
```

## Acceptance rate

```{r}
p3 <- p |> 
  group_by(p) |> 
  summarize(m = median(diff),
            n = n())
library("ggplot2")
p3 |> 
  ggplot(aes(p, n)) +
  # geom_point(aes(size = n)) +
  geom_smooth(span = 1/24) +
  scale_y_continuous( limits = c(0, NA), expand = expansion(0, add = 0.5)) +
  scale_x_date(date_breaks = "years", date_labels = "%Y") +
  labs(y = "Packages ", 
       x = element_blank(),
       title = "Daily packages accepted in CRAN") +
  theme_minimal()
```

```{r}
library("lubridate")
p |> 
  count(p = floor_date(p, "month")) |> 
  ggplot(aes(p, n)) +
  # geom_point(aes(size = n)) +
  geom_smooth(span = 1/12) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(0, add = 0.5), 
                      position = "right") +
  scale_x_date(date_breaks = "years", date_labels = "%Y") +
  labs(y = "Packages ", 
       x = element_blank(),
       title = "Monthly packages accepted in CRAN") +
  theme_minimal()
```


```{r}
p2 |> 
  mutate(nt = cumsum(n)) |> 
  ggplot(aes(p, nt)) +
  # geom_point(aes(size = n)) +
  geom_smooth(span = 1/24) +
  geom_hline(yintercept = nrow(packages), col = "black") +
  scale_y_continuous( limits = c(0, NA), expand = expansion(0, add = 0.5)) +
  scale_x_date(date_breaks = "years", date_labels = "%Y") +
  labs(y = "Packages ", 
       x = element_blank(),
       title = "Daily packages accepted in CRAN") +
  theme_minimal()
```



```{r}
yesno <- function(x) {
  k <- x
  k[x] <- "Yes"
  k[!x] <- "No"
  k
}
ggplot(p) +
  geom_point(aes(p, Published, col = as.numeric(diff))) +
  scale_color_viridis_c(trans = "log10")
```

## Developer delays

Difference between packages and build.

```{r}
library("ggpattern")
ggplot(p) +
  geom_histogram(aes(as.numeric(diff), fill = yesno(with_archives))) +
  # facet_wrap(~yesno(with_archives), scales = "free") +
  scale_y_log10(expansion(), limits = c(1, NA)) +
  scale_x_continuous(expand = expansion()) +
  labs(y = "Packages", 
       x = "Time (days)",
       title = "Time from build to publication",
       fill = "First time?") +
  theme_minimal() +
  theme(legend.position = c(0.7, 0.7))
```


```{r}
p2 <- p |> 
  group_by(p, with_archives) |> 
  summarize(m = median(diff),
            n = n()) |> 
  ungroup()

p2 |> 
  ggplot(aes(p, m, col = yesno(with_archives))) +
  # geom_point(aes(size = n)) +
  geom_smooth(span = 1/24) +
  scale_y_continuous( limits = c(0, NA), expand = expansion()) +
  scale_x_date(date_breaks = "years", date_labels = "%y") +
  labs(y = "Time (days)", 
       x = element_blank(),
       col = "Archives?",
       title = "Time between building and in CRAN") +
  theme_minimal()
```

## Archives

The archives of CRAN is a true treasure too.

```{r}
archive_df <- do.call("rbind", archive)
archives <- vapply(archive, nrow, numeric(1))
pkg <- rep(names(archive), times = archives)
archive_df$package <- pkg

version <- gsub(".*_(.*)\\.tar\\.gz$", "\\1", rownames(archive_df))
archive_df$version <- version
archive_df$version[archive_df$version == archive_df$package] <- NA
archive_df$status <- "archived"

# Make sure they match
current$package <- rownames(current)
current$version <- packages$version[match(current$package, packages$package)]
current$status <- "available"
all_packages <- rbind(archive_df, current)
library("dplyr")
all_packages2 <- all_packages[, c("Package", "mtime", "Version", "uname", "grname", "size")] |> 
  mutate(date = as.Date(mtime)) |> 
  select(-mtime) |> 
  arrange(package, date) |> 
  group_by(package) |> 
  mutate(trelative = difftime(date, min(date), units = "weeks"),
         tprevious = t_relative - lag(trelative, default = 0)
         n = n(),
         release = 1:n()) |> 
  ungroup()
rownames(all_packages2) <- NULL
```

```{r}
count(all_packages2, package, sort = TRUE, name = "Releases") |> 
  head(20)
count(all_packages2, uname, sort = TRUE)
```


```{r}

a <- archive_df2 |> 
  group_by(Package) |> 
  filter(trelative != 0) |> 
  filter(trelative == min(trelative)) |>
  ungroup()
all_packages2 |> 
  filter(release == 2) |> 
  ggplot() +
  geom_histogram(aes(trelative)) +
  labs(x = "Time (weeks)",
       y = "Packages",
       title = "Time since first release") +
  scale_x_continuous(expand = expansion()) +
  scale_y_log10(expand = expansion()) +
  theme_minimal()
  
all_packages2 |> 
  mutate(versions = as.character(release)) |>
  ggplot() +
  geom_point(aes(x = as.numeric(trelative), y = release, col = package)) +
  guides(col = "none") +
  labs(y = "Archived versions", 
       x = "Time since first release (weeks)", 
       title = "Time since the first releast to the second archival") +
  scale_x_continuous(expand = expansion(), limits = c(0, NA)) +
  theme_minimal() 
```


```{r}
all_packages2 |> 
  ggplot() +
  geom_line(aes(x = date, y = release, group = package))
  guides(col = "none") +
  labs(y = "Archived versions", 
       x = "Time since first release (weeks)", 
       title = "Time since the first releast to the second archival") +
  theme_minimal() 
```


```{r}
all_packages2 |> 
  ggplot() +
  # geom_point(aes(date, trelative), alpha = 0.5) +
  geom_smooth(aes(date, trelative), span = "1 month") +
  scale_x_date(date_breaks = "years", date_labels = "%Y") +
  labs(x = element_blank(), 
       y = "Time since first archival (days)",
       title = "Tendency of time to update the package") +
  theme_minimal()
```


```{r}

```

## Oldest package

Given the archives and the current packages on CRAN.
Which are the oldest packages?

```{r}
packages |> 
  ggplot() +
  geom_histogram(aes(p))
```

## Best conserved package

Looking at packages with just 1 release but still one package which are those?

```{r}
never_archived <- all_packages2 |> 
  group_by(package) |> 
  filter(max(release) == 1) |> 
  ungroup() |> 
  pull(package)
never_archived <- never_archived[never_archived %in% current$Package]

direct_dep <- package_dependencies(never_archived, db = p)
deps <- lengths(direct_dep)
recurs_dep <- package_dependencies(never_archived, db = p, recursive = TRUE)
r_deps <- lengths(recurs_dep)
r_deps <- r_deps[names(deps)]
deps_all <- data.frame(direct_deps = deps,
                       r_deps)
deps_all |> 
  count(direct_deps, r_deps) |> 
  ggplot() + 
  geom_point(aes(direct_deps, r_deps, col = n)) +
  # geom_abline(slope = 1, intercept = 0) +
  scale_color_continuous(trans = "log10") +
  labs(x = "Direct dependencies",
       y = "Recursive dependencies",
       title = "Dependenecies of packages never archived",
       col = "Packages") +
  theme_minimal()
```

Now that we know that there is a wide range of packages never archived/updated maybe those packages are recent

```{r}
packages_k <- intersect(never_archived, p$Package)
p |> 
  filter(Package %in% packages_k) |> 
  count(p) |> 
  ggplot() +
  geom_point(aes(p, n))
```

This shows the date of the packages in relation to their recursive dependencies. 

### Dependencies make unstable packages

There are much to say about packages and their dependencies but CRAN test them. 


### Reproducibility

<details>
```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info()
```
</details>
