---
title: CRAN updates
author: Lluís Revilla Sancho
date: '2022-05-28'
slug: cran-updates
categories:
  - CRAN
  - r
tags:
  - R
authors: []
description: ''
draft: no
editor_options:
  chunk_output_type: console
featured: no
image:
  caption: ''
  focal_point: ''
subtitle: ''
summary: ''
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, cache= TRUE, fig.retina = 2)
Sys.setlocale("LC_ALL", "C")
```

There are many great things in base R, one of them is the [tools package](https://search.r-project.org/R/refmans/tools/html/00Index.html).

Looking into the source code and exploring how it is used by CRAN I found some internal functions that access freely available files with information about CRAN packages.
These private functions are at the CRANtools.R file.

```{r internal-functions}
packages <- tools:::CRAN_package_db()
current <- tools:::CRAN_current_db()
archive <- tools:::CRAN_archive_db()
# issues <- tools::CRAN_check_issues()
# alias <- tools:::CRAN_aliases_db()
# rdxrefs <- tools:::CRAN_rdxrefs_db()
```

I will focus on the first three functions as they have information about current packages and previous packages on CRAN.

First of all a very brief exploration of what is in this packages

```{r package-head}
packages[1, 1:10]
```

Packages has similar information as `available.packages()` but with many more columns with published date, reverse dependencies, X-CRAN-Comment, who packaged...
Also note that all this packages are not filtered to match R version, OS_type, subarch and duplicates (I learned about this filtering while reading the great documentation of available.packages and also finding some mentions online). 

The current database has different data, it has the package size, dates of modification, which I assume is date added to CRAN and user name of who last modified it.
This is the same information returned by [`file.info`](https://search.r-project.org/R/refmans/base/html/file.info.html)

```{r current-head}
current[1, 1:10]
```


The archive database returns the same information, but as you might guess by the name it doesn't provide information about current packages but for packages not longer available by default.

```{r archive-head}
archive[[1]]
```

The date matches that available on the [Old sources](https://cran.r-project.org/src/contrib/Archive/A3/).
However, as I was not sure of the information I asked on [R-devel](https://stat.ethz.ch/pipermail/r-devel/2022-May/081770.html) but recieved no answer. 


Now that we know what is there let's explore the data. 
Some ideas, and questions to explore:

- When were the current [packages accepted on CRAN?](#accepted)
- There is the dates of all packages accepted on CRAN (except those removed): [Are more packages accepted now than in previous years?](#releases)
- There is the date when a package is build and when it is published: How long does it take between a [developer builds the package till it is published on CRAN?](#delays)
- Using the user information; who is [accepting packages on CRAN?](#admins)
- Given the dates, is there some [seasonal patterns in CRAN acceptance?](#season)
- What are the [**oldest** packages on CRAN?](#oldest) 
- What are the [**biggest** packages on CRAN?](#biggest) 
- Are there difference in [size between old and new packages?](#size) 
- Given the dates and versions of packages, [are dependencies related with newer packages or updates?](#dependencies)

This is only using these files wihout cross checking with other information available about CRAN (like [PACKAGES.in](https://llrs.dev/post/2021/12/07/reasons-cran-archivals/) or [submission records](https://llrs.dev/post/2021/01/31/cran-review/)).
Using together all this data will provide a much complete picture and might help answer many more questions.

Let me know if you explore this data with a different question or the same and you find something different/interesting!

```{r libraries}
library("dplyr")
library("ggplot2")
library("tidyr")
library("ggrepel")
library("lubridate")
library("tools")
library("forcats")
library("flextable")
library("rversions")
```

First of all some preparation; as it 

```{r}
df <- data.frame(major = integer(), minor = integer(), patch = integer())
r_versions <- r_versions() |> 
  mutate(date = as.Date(date),
         strcapture("([0-9])\\.([0-9]*)\\.?([0-9]*)", 
                    version, proto = df),
         equal = major == lag(major, default = TRUE) & minor == lag(minor, default = TRUE))
```

## Accepted packages {#accepted}

[CRAN was set up in 1997](https://stat.ethz.ch/pipermail/r-announce/1997/000001.html) (Right after the r-testers mailing list was split in [three mailing list](https://stat.ethz.ch/pipermail/r-announce/1997/000000.html)), since then there have been many additions of packages but how fast have been packages accepted? We will use the packages database to answer this. 

First of all some clean up to have all the dates in UTC (not that it will affect much but  better to have uniform data), store all the version of packages and add some columns to make it easy relate it with the other files.


```{r packages-setup}
packages$Published <- as.Date(packages$Published, format = "%Y-%m-%d",
                              tz = "UTC")
packages$Packaged_by <- gsub(".*; (.*)", "\\1", packages$Packaged)
packages$Packaged <- as.POSIXct(packages$Packaged, 
                                format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
packages2 <- packages |> 
  mutate(date =  as.Date(Packaged),
         diff = round(difftime(Published, date, units = "days"), 0)) |> 
  mutate(with_archives = Package %in% names(archive)) |> 
  distinct(Package, .keep_all = TRUE)
```

According to this file the oldest current package added was in 2010:

```{r daily-cran}
minor_changes0 <- filter(r_versions, !equal)
minor_changes <- filter(minor_changes0, date >= min(packages2$date, 
                                                    na.rm = TRUE))

packages2 |> 
  count(date) |> 
  ggplot() +
  # geom_point(aes(size = n)) +
  geom_text_repel(aes(x = date, y = 25, label = version),
                  data = minor_changes, nudge_x = 0, angle = 90) +
  geom_smooth(aes(date, n), span = "1 month") +
  scale_y_continuous(expand = expansion(0, add = 0.5)) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_x_date(date_breaks = "years", date_labels = "%Y") +
  labs(y = "Packages ", 
       x = element_blank(),
       title = "Daily packages accepted on CRAN") +
  theme_minimal()
```

The daily rate of acceptance has increased from less than 10 a day till 2020 to more than 30 this year 2022.
If we summarize that information for month we see the same, but the little bump in 2020 disappears but we see other patterns:

```{r monthly-cran}
packages2 |> 
  count(month = floor_date(date, "month")) |> 
  ggplot(aes(month, n)) +
  # geom_point(aes(size = n)) +
  geom_smooth(span = 1/12) +
  geom_text_repel(aes(x = date, y = 600, label = version), 
                  data = minor_changes, nudge_x = 0, angle = 90) +
  scale_y_continuous(expand = expansion(0, add = 0.5), position = "right") +
  coord_cartesian(ylim = c(0, NA),
                  xlim = c(as.Date("2010-01-01"), NA)) +
  scale_x_date(date_breaks = "years", date_labels = "%Y") +
  labs(y = "Packages ", 
       x = element_blank(),
       title = "Monthly packages accepted on CRAN") +
  theme_minimal()
```

Instead of just one bump we see some waves with less packages on CRAN accepted late in the year and an increase of packages the first months of the year.
We'll explore this [temporal patterns later](#seasonal).

If we look at the accumulated packages on CRAN we see an exponential growth:

```{r cran-cumsum}
packages2 |> 
  group_by(date) |> 
  count() |> 
  ungroup() |> 
  mutate(nt = cumsum(n)) |> 
  ggplot(aes(date, nt)) +
  # geom_point(aes(size = n)) +
  geom_text_repel(aes(x = date, y = 10000, label = version),
                  data = minor_changes, nudge_x = 0, angle = 90) +
  geom_smooth(span = "1 week") +
  # geom_hline(yintercept = nrow(packages), col = "black") +
  scale_y_continuous(expand = expansion(0, add = 0.5)) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_x_date(date_breaks = "years", date_labels = "%Y") +
  labs(y = "Packages ", 
       x = element_blank(),
       title = "Cumulative sum of packages accepted on CRAN") +
  theme_minimal()
```

In conclusions, more packages are being accepted now than ever in the past.

## Developer delays {#delays}

Next question is: how fast is the process for developers of packages? 

We can compare the Published date with the build date to answer this.
The build date is done by the developer to create the tar.gz file submitted to CRAN while the Published date is set by CRAN once the packages is accepted on CRAN.

A first exploration of this dates:

```{r cran-built}
#' Convert TRUE, FALSE to Yes, No for easier reading
#' @param x Logical vector
yesno <- function(x) {
  k <- x
  k[x] <- "Yes"
  k[!x] <- "No"
  k
}

ggplot(packages2) +
  geom_point(aes(date, Published, col = as.numeric(diff))) +
  scale_color_viridis_c(trans = "log10") +
  coord_cartesian(ylim = c(as.Date(min(packages2$Published)), NA),
                  xlim = c(as.Date(min(packages2$Date)), NA)) +
  scale_x_date(expand = expansion(), date_breaks = "2 year", date_labels = "%Y") +
  scale_y_date(expand = expansion(), date_breaks = "2 year", date_labels = "%Y") +
  theme_minimal() +
  labs(x = "Built",
       y = "Published",
       col = "Difference (days)",
       title = "Time between building a package and being published")
```

We can see some point that it was build much later than published (published in 2017 but built in 2017), this might indicate that the CRAN team had to rebuilt the package a problem with the date settings of the computer or other causes.

To better visualize the differences we can focus on only those with some positive difference between publish date and build.
I will also compare if there is some difference with new packages and packages that were already on CRAN:

```{r cran-delays}
packages2 |> 
  filter(as.numeric(diff) >= 0) |> 
  ggplot() +
  geom_histogram(aes(as.numeric(diff), 
                     fill = fct_relevel(yesno(with_archives), c("Yes", "No"))),
                 position = "identity", alpha = 0.5, binwidth = 7) +
  # facet_wrap(~yesno(with_archives), scales = "free_x") +
  scale_y_log10(expand = expansion()) +
  scale_x_continuous(expand = expansion()) +
  coord_cartesian(xlim = c(0, NA), ylim = c(1, NA)) +
  labs(y = "Packages", 
       x = "Time (days)",
       title = "Time from build to publication",
       fill = "First time?") +
  theme_minimal() +
  theme(legend.position = c(0.7, 0.7))
```

There doesn't seem to be much difference between date of building and date of publication according to if it is the first release or not.
The times are short and usually well below 50 days and too short to notice in this scale.
This doesn't mean that the packages first time accepted on CRAN didn't had previous submissions, but when they finally built and pass the checks and manual review they are handled as with the other packages already on CRAN.

However, this might have changed with time:

```{r cran-delays2}
packages2 |> 
  filter(as.numeric(diff) >= 0) |> 
  group_by(date, with_archives) |> 
  summarize(m = median(diff),
            n = n()) |> 
  ungroup() |> 
  ggplot() +
  # geom_point(aes(size = n)) +
  geom_smooth(aes(date, m, col = yesno(!with_archives), 
             linetype = yesno(!with_archives), weight = n), span = "1 month") +
  geom_text_repel(aes(x = date, y = 10, label = version),
                  data = minor_changes, nudge_x = 0, angle = 90) +
  scale_y_continuous(expand = expansion()) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y", 
               expand = expansion(add = 0.5)) +
  coord_cartesian(ylim =  c(0, NA)) +
  labs(y = "Time (days)", 
       x = element_blank(),
       col = "New?",
       linetype = "New?",
       title = "Time between building and on CRAN") +
  theme_minimal() +
  theme(legend.position = c(0.7, 0.7))
```

Here we clearly see that those already on CRAN are handled faster than new packages which take longer to be published.
This is consistent with the manual review process (For more information see [this other blog post](https://llrs.dev/post/2021/01/31/cran-review/)).
It means also that there is a huge variation of time about how packages are handled.
However this seems to be reducing: while in 2010 it took around 2 weeks nowadays it takes less than a week and getting closer to time an already on CRAN package takes: a 1 day of median time  between a package being built and appearing on CRAN that takes for existing packages.
Maybe the delay in 2020 was due the huge volume of submissions CRAN received.

<details>
<summary>More packages mean further delays? No, but I'm not sure </summary>

```{r cran-reasons}
packages2 |> 
  filter(as.numeric(diff) >= 0) |> 
  group_by(date, with_archives) |> 
  summarize(m = median(diff),
            n = n()) |> 
  ungroup() |> 
  ggplot() +
  # geom_point(aes(size = n)) +
  geom_smooth(aes(n, m, col = yesno(with_archives), 
             linetype = yesno(with_archives), weight = n)) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(y = "Time (days)", 
       x = "Packages accepted the same day",
       col = "Archives?",
       linetype = "Archives?",
       title = "Delay relationship with packages accepted the same day") +
  theme_minimal() +
  theme(legend.position = c(0.7, 0.7))
```

As expected we see a lot of variation on the delay of packages going through manual review.
Surprisingly it seems that the more new packages accepted the same day the less delay there is. 
I think this just means that when reviewers get to it several packages might be approved. 
This might also mean that packages have already been built several times before finally being accepted and now the errors, warnings and notes have been solved.

For packages already accepted comparing days of delay is not informative as they typically take less than a day.
However, comparing hours of delay would be misleading as it can simply be the developer taking some time to double check before submission or a time zone difference.

</details>


## Archives

As we have seen there are some files about the archives of CRAN.
These include information about date of modification (moving/editing) and user who did it and of course name and sometimes version of the package.
These archives are the great treasure of CRAN even if we don't have the same information about them as current packages on CRAN.

Note that I'm not sure that this archives contains the full record of packages, some initial packages might be missing and I'm aware of some packages removed by CRAN which do not longer appear on this records. 

Nevertheless this should provide an accurate picture of packages available. 
Also as there is no information when a package is archived (here, [there is on PACKAGES.in](https://llrs.dev/post/2021/12/07/reasons-cran-archivals/)), we might slightly overestimate the packages available at a given moment.

```{r archives-setup}
archive_df <- do.call("rbind", archive)
archives <- vapply(archive, nrow, numeric(1))
pkg <- rep(names(archive), times = archives)
archive_df$package <- pkg

version <- gsub(".*_(.*)\\.tar\\.gz$", "\\1", rownames(archive_df))
archive_df$version <- version
archive_df$version[archive_df$version == archive_df$package] <- NA
archive_df$status <- "archived"

# Make sure they match
current$package <- rownames(current)
current$version <- packages$Version[match(current$package, packages$Package)]
current$status <- "available"
all_packages <- rbind(archive_df, current)

# Arrange dates and data
all_packages$mtime <- with_tz(all_packages$mtime, tzone = "UTC")
keep_columns <- c("package", "mtime", "version", "uname", "size", "status")
all_packages2 <- all_packages[, keep_columns] |> 
  mutate(date = as.Date(mtime)) |> 
  select(-mtime) |> 
  arrange(package, date) |> 
  group_by(package) |> 
  mutate(trelative = difftime(date, min(date), units = "weeks"),
         tprevious = trelative - lag(as.numeric(trelative), default = 0),
         n = n(),
         release = 1:n(),
         available = yesno(any(status == "available"))) |> 
  ungroup()
```

Remember the plot about [acceptance of packages on CRAN?](#accepted)
That plot only looked at current packages available, let's check it with all the archive:

```{r}
all_packages2 |> 
   group_by(date) |> 
  count() |> 
  ungroup() |> 
  mutate(nt = cumsum(n)) |> 
  ggplot() +
  # geom_point(aes(size = n)) +
  geom_smooth(aes(date, nt), span = "1 week") +
  geom_text_repel(aes(x = date, y = 100000, label = version),
                  data = minor_changes0, nudge_x = 0, max.overlaps = 15, angle = 90) +
  # geom_hline(yintercept = nrow(packages), col = "black") +
  scale_y_continuous(expand = expansion(0, add = 0.5)) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(y = "Packages ", 
       x = element_blank(),
       title = "Accumulative packages on CRAN archive") +
  theme_minimal()
```

To have an overview of the data we can see some output
If we look at some basics tally to find which packages have more releases:

```{r archives-release}
count(all_packages2, package, sort = TRUE, name = "Releases") |> 
  head(20) |> 
  flextable() |> 
  autofit()
```

Surprisingly there are packages with more than 200 versions on CRAN!

And which CRAN amdin has been moving more packages:

```{r archives-managers}
count(all_packages2, uname, sort = TRUE) |> 
  flextable() |> 
  autofit()
```

Not surprisingly the three top users is dominated by known names associated with CRAN, followed by the root user, perhaps there is an automated script run as root. 
What surprised me more are finding some names, that I associate less with CRAN.
Although Zileis also signs the R-journal sections about CRAN, there are other surnames from people that is not on the R Core. 

```{r packages-available}
all_packages2 |> 
  distinct(package, available) |> 
  count(available, name = "packages", sort = TRUE) |> 
  mutate(proportion = scales::percent(packages/max(packages))) |> 
  flextable() |> 
  autofit()
```

The number of packages currently on CRAN `r nrow(current)` and those not on CRAN are around a 15% of that.

Given all this different versions of packages how big are all the packages on CRAN?

### CRAN size

CRAN source packages are approximately:

```{r}
k <- sum(all_packages2$size)
class(all_packages2$size) <- "object_size"
class(k) <- "object_size"
format(k, units = "auto")
```

This doesn't include binaries for multiple architectures and OS, but it is a good indicator of the source file and if the package has some considerable amount of data.

Looking back to the size of the packages along time we can see this pattern:

```{r}
all_packages2 |> 
  group_by(date, available, status) |> 
  summarise(s = median(size),
            n = n()) |> 
  ungroup() |> 
  ggplot() +
  geom_smooth(aes(date, s, col = available, linetype = status, weight = n),
              method = "gam") +
  geom_text_repel(aes(x = date, y = 500000, label = version),
                  data = minor_changes0, nudge_x = 0, max.overlaps = 15, angle = 90) +
  labs(x = element_blank(),
       y = "Package size (B)",
       col = "Package on CRAN?",
       linetype = "Version available?",
       title = "Median size of packages in CRAN") +
  scale_y_continuous(expand = expansion(), limits = c(0, NA)) +
  scale_x_date(expand = expansion()) +
  theme_minimal() +
  theme(legend.position = c(0.2, 0.8))
```

Packages available on CRAN are smaller than those no longer on CRAN. 
Even versions of packages on CRAN that got archived are usually bigger than current versions. 
Median size of packages is increasing quickly. 

```{r}
all_packages2 |> 
  group_by(release) |> 
  summarise(s = median(size), n = n()) |> 
  ungroup() |> 
  ggplot() +
  geom_smooth(aes(release, s, weight = n)) +
  scale_y_log10() +
  geom_hline(yintercept = 5*1000^2) +
  labs(x = "Release number",
       y = "Size (B)",
       title = "Package size with release number")
```

Typically packages increase their size with new releases up to when they reach 50 releases.

```{r eval=FALSE, include=FALSE}
aps <- all_packages2 |> 
  group_by(package) |> 
  summarise(m = median(size),
            n = n()) |> 
  ungroup() |> 
  arrange(-m)
```


## Releases

As CRAN is not static and there are new releases in a continuous way let's explore other releases.

```{r cran-turnover}
type_updates <- all_packages2 |> 
  group_by(year = floor_date(date, "year"), package) |> 
  summarise(type = case_when(any(release == 1) & any(release > 1) ~ "new & update",
                             any(release > 1) ~ "update",
                             release == 1 ~ "new",
                             TRUE ~ "update"),
            n = n()) |> 
  ungroup() |> 
  distinct(.keep_all = TRUE) |> 
  group_by(year) |> 
  count(type, name = "packages") |> 
  mutate(prop = packages/sum(packages)) |> 
  ungroup()

type_updates |> 
  ggplot() +
  geom_col(aes(year, packages, fill = type), position = "stack") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y", 
               expand = expansion()) +
  scale_y_continuous(expand = expansion()) +
  theme_minimal() +
  labs(title = "Packages news on CRAN",
       x = element_blank(),
       y = element_blank(),
       fill = "Type") +
  theme(legend.position = c(0.2, 0.75), plot.title.position = "plot")
```

There is an increase number of work on CRAN to update and add new packages.
I counted as new & updated those packages that were inlcuded on CRAN and in the same year they were updated.
This is to keep in mind that [CRAN policies](https://cran.r-project.org/web/packages/policies.html#Submission) says: 

> Once a package is established (which may take several rounds), “no more than every 1–2 months” seems appropriate. 

To it is kind of expected for new packages to take some time to establish themselves. 

Note that this does not include packages removed from CRAN.

Also note that in 2021 there was the first reduction in packages approvals (around 160). 
This was also noted in the [R journal: changes on CRAN](https://journal.r-project.org/archive/2021-2/cran.pdf) mentioning a reduction in submissions.
It is also worth mentioning the big increase in packages approved for inclusion on CRAN in 2020 (effect of the COVID pandemic?).
Perhaps some packages that would be submitted later got sooner to CRAN?

```{r cran-proportion}
ggplot(type_updates) +
  geom_col(aes(year, prop, fill = type)) +
  scale_y_continuous(labels = scales::label_percent(), expand = expansion(), n.breaks = 10, limits = c(0, 1)) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y",
               expand = expansion()) +
  theme_minimal() +
  labs(title = "Updates of existing packages on CRAN",
       subtitle = "Percentage of packages",
       x = element_blank(),
       y = element_blank(),
       fill = "Type")
```

Very early on, since 2000, the most frequent action on CRAN is updating a package.
This has remained more or less stable with time close to 80% if we take into account also the new packages that are updated the same year. 

This is consistent with language updates and the increasing requirements to keep a package on CRAN.

Given that updates is the most pressing matter to the CRAN team, let's explore them.

Percentage of packages updated on CRAN: 
```{r}
CRAN_till_then <- all_packages2 |> 
  arrange(date) |> 
  group_by(date, package) |> 
  mutate(type = case_when(any(release > 1) ~ "update",
                             release == 1 ~ "new",
                             TRUE ~ "update")) |> 
  ungroup() |> 
  arrange(date) |> 
  mutate(CRAN = cumsum(type == "new"),
         actions = 1:n(),
         updates_cum = cumsum(type == "update")/actions,
         new_cum = CRAN/actions,
         )
CRAN_till_then |>
  ggplot() +
  geom_smooth(aes(date, updates_cum), linetype = 2) +
  geom_smooth(aes(date, new_cum), col = "orange")  +
  scale_y_continuous(expand = expansion(), labels = scales::percent_format()) +
  labs(x = element_blank(),
       y = "CRAN updates",
       title = "Acumulative percentage of actions") +
  theme_minimal()
CRAN_till_then |>
  group_by(date) |> 
  summarise(updates_cran = sum(type == "update")/max(CRAN)) |> 
  ggplot() +
  geom_smooth(aes(date, updates_cran), linetype = 2) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = element_blank(),
       y = "% of CRAN",
       title = "Percentage of packages on CRAN updated") +
  theme_minimal()
```

It seems that 0.2% of packages of CRAN is updated daily!
There has been a slow down on packages updated on CRAN since the beginning.

## Updates

Updates are new releases to CRAN, looking at the second update:

```{r second-release}
all_packages2 |> 
  filter(release == 2) |> 
  ggplot() +
  # geom_vline(xintercept = 52*c(1:15), linetype = 2) +
  geom_histogram(aes(trelative), binwidth = 26) +
  labs(x = "Time (weeks)",
       y = "Packages",
       title = "Time since first release") +
  scale_x_continuous(expand = expansion(), breaks = 52*c(1:19)) +
  scale_y_log10(expand = expansion()) +
  theme_minimal()
```

There is a wide range of time between the first release and the next one.
Those that are still on CRAN show any difference to those that are not on CRAN?

```{r second-release-status}
all_packages2 |> 
  filter(release == 2) |> 
  ggplot() +
  # geom_vline(xintercept = 52*c(1:15), linetype = 2) +
  geom_histogram(aes(trelative, 
                     fill = fct_relevel(available, "Yes", "No")), 
                 binwidth = 26, position = "identity", alpha = 0.5) +
  labs(x = "Time (weeks)",
       y = "Packages",
       fill = "On CRAN?",
       title = "Time since first release") +
  scale_x_continuous(expand = expansion(), n.breaks = 18) +
  scale_y_log10(expand = expansion()) +
  theme_minimal() +
  theme(legend.position = c(0.7, 0.7))
```

There doesn't seem to be much difference between packages on CRAN and archived.
Perhaps given the low number of packages no longer on CRAN it is surprising that there are so many updates to later being removed.
This indicates a lot of effort from the package developers that is later no longer available to the users.
The causes of this might be multiple, perhaps they do not have enough time to keep up with the "[publication quality](https://cran.r-project.org/web/packages/policies.html)" standard of CRAN, the package is superseeded by other tools, or there aren't good experiences about updating packages.

```{r releases-times}
all_packages2 |> 
  filter(release <= 6 & release != 1) |> 
  ggplot() +
  # geom_vline(xintercept = 52*c(1:15), linetype = 2) +
  geom_histogram(aes(tprevious, 
                     fill = fct_relevel(available, "Yes", "No")), 
                 binwidth = 26, position = "identity", alpha = 0.5) +
  facet_wrap(~release, scales = "free_x") +
  labs(x = "Time (weeks)",
       y = "Packages",
       fill = "On CRAN?",
       title = "Time since previous release",
       subtitle = "By number of release") +
  scale_x_continuous(expand = expansion()) +
  scale_y_log10(expand = expansion()) +
  theme_minimal() +
  theme(legend.position = c(0.8, 0.3))
```

This might be obvious but packages currently not on CRAN are not updated as much as those that are.
But note that they are only \~20% of the packages on CRAN, so proportionally removed packages updated more.
Wit this data there isn't any explanation we can make about why aren't packages updated more.

```{r last-release}
all_packages2 |> 
  group_by(package) |> 
  filter(release == max(release) & release != 1) |> 
  ggplot() +
  # geom_vline(xintercept = 52*c(1:13), linetype = 2) +
  geom_histogram(aes(tprevious, 
                     fill = fct_relevel(status, c("available", "archived"))), 
                 binwidth = 26, position = "identity") +
  labs(x = "Time (weeks)",
       y = "Packages",
       fill = "Status",
       title = "Time since previous release") +
  scale_x_continuous(expand = expansion()) +
  scale_y_log10(expand = expansion()) +
  theme_minimal()
```

```{r}
prop_packages <- all_packages2 |> 
  count(release, available) |>
  group_by(release) |> 
  mutate(prop_release = n/sum(n)) |> 
  ungroup() |> 
  group_by(available) |> 
  mutate(prop_available = n/max(n)) |> 
  ungroup()
prop_packages |> 
  ggplot() +
  geom_tile(aes(available, release, fill = n)) +
  # facet_wrap(~available, scales = "free_x") +
  scale_fill_viridis_c(trans = "log10") +
  scale_y_continuous(expand = expansion()) +
  theme_minimal() +
  labs(x = "On CRAN?",
       y = "Releases",
       fill = "Packages",
       title = "Package updates") +
  theme(legend.position = c(0.2, 0.8))
```

By the gradation of colors and length there is a clear sense that packages on CRAN are updated much more and have more releases (even if those pacakges with more than 150 releases are very few).

```{r}
prop_packages |> 
  ggplot() +
  geom_line(aes(release, prop_available, col = available, linetype = available)) +
  scale_x_continuous(expand = expansion()) +
  scale_y_continuous(expand = expansion(), labels = scales::label_percent()) +
  labs(x = "Release",
       y = "Proportion",
       linetype = "On CRAN?",
       col = "On CRAN?",
       title = "Packages on CRAN are updated more."
       ) +
  theme_minimal() +
  coord_cartesian(xlim = c(0, 50)) +
  theme(legend.position = c(0.7, 0.7))
```

From this plot we can see that packages on CRAN make update more than those who are no longer on CRAN. 
Maybe packages removed are only updated on other platforms (Bioconductor, Github, R-forge), or are no longer maintained.

Similarly the speed of release of each version might be important

```{r releases-speed}
all_packages2 |> 
  mutate(versions = as.character(release)) |>
  ggplot() +
  geom_line(aes(x = as.numeric(trelative), y = release, col = package, 
                group = package)) +
  guides(col = "none") +
  labs(y = "Versions", 
       x = "Time since first release (weeks)", 
       title = "Time since the first release to the second archival") +
  scale_x_continuous(expand = expansion(), limits = c(0, NA)) +
  theme_minimal() 
```

```{r releases-speed2}
all_packages2 |> 
  filter(release >= 2) |> 
  ggplot() +
  geom_count(aes(x = as.numeric(trelative), as.numeric(tprevious), col = release)) +
  labs(y = "Time sine previous release (weeks)", 
       x = "Time since first release (weeks)", 
       col = "Releases",
       size = "Packages",
       title = "Time between releases") +
  scale_x_continuous(expand = expansion(add = 1), limits = c(0, NA)) +
  scale_y_continuous(expand = expansion(add = 1), limits = c(0, NA)) +
  theme_minimal() 
```

Looking at time since first release and time since previous release we don't see any pattern.
Some packages have few updates but distributed in time while some seem to have more constant distributions but there isn't a clear pattern. 

```{r}
rle_weeks <- function(x, y) {
  x <- x[y > 8] # Wait two months to have established package
  rle <- rle(x < 4)
  rle$lengths[rle$values ] >= 2
}

packages_updates <- all_packages2 |> 
  filter(n > 1,
         available == "Yes") |> 
  group_by(package) |> 
  summarise(d = min(date),
            max = as.numeric(difftime(Sys.Date(), d), units = "weeks"),
            releases = unique(n),
            m = releases/max,
            trouble2 = rle_weeks(tprevious, trelative)) |> 
  ungroup() |> 
  filter(max > 8) |> 
  mutate(troubles = yesno(m >= 1/8),
         troubles2 = yesno(trouble2))

ggplot(packages_updates) +
  geom_smooth(aes(d, m)) +
  theme_minimal() +
  scale_y_continuous(
    "Updates per week",
    sec.axis = sec_axis(~ . / 4, name = "Updates per month")
  ) +
  labs(x = element_blank(),
       title = "")
```

Developers lately seem to have to update packages more frequently. 
This plot is packages 2 months old on CRAN and the number of releases in their lifetime.
Clearly now there are some packages that need to update if faster.

If we split between those that have more troubles (more than 1 updates per 8 weeks as in code), we see that this is happening to all packages recently accepted: 

```{r}
ggplot(packages_updates) +
  geom_smooth(aes(d, m, col = troubles, linetype = troubles)) +
  geom_text_repel(aes(x = date, y = 0.1, label = version),
                  data = minor_changes0, nudge_x = 0, angle = 90) +
  theme_minimal() +
  scale_y_continuous(
    "Updates per week",
    sec.axis = sec_axis(~ . / 4, name = "Updates per month")
  ) +
  labs(x = element_blank(),
       title = "New packages require more updates")
ggplot(packages_updates) +
  geom_smooth(aes(d, m, col = troubles2, linetype = troubles2)) +
  geom_text_repel(aes(x = date, y = 0.1, label = version),
                  data = minor_changes0, nudge_x = 0, angle = 90) +
  theme_minimal() +
  scale_y_continuous(
    "Updates per week",
    sec.axis = sec_axis(~ . / 4, name = "Updates per month")
  ) +
  labs(x = element_blank(),
       title = "New packages require more updates")
packages_updates |> 
  mutate(troubles_all = troubles2 == "Yes" & troubles == "Yes") |> 
ggplot() +
  geom_smooth(aes(d, m, col = troubles_all, linetype = troubles_all)) +
  geom_text_repel(aes(x = date, y = 0.1, label = version),
                  data = minor_changes0, nudge_x = 0, angle = 90) +
  theme_minimal() +
  scale_y_continuous(
    "Updates per week",
    sec.axis = sec_axis(~ . / 4, name = "Updates per month")
  ) +
  labs(x = element_blank(),
       title = "New packages require more updates")
```

This could indicate problems following CRAN policies, keeping up to date with changes in R code and checks.
Despite troubles keep pushing to keep packages on CRAN.
Some succeed some don't and do not longer submit new releases.


```{r releases-dates}
all_packages2 |> 
  ggplot() +
  geom_line(aes(x = date, y = release, group = package)) +
  guides(col = "none") +
  labs(y = "Archived versions", 
       x = "Time since first release (weeks)", 
       title = "Time since the first releast to the second archival") +
  theme_minimal() 
```


```{r}
all_packages2 |> 
  ggplot() +
  # geom_point(aes(date, trelative), alpha = 0.5) +
  geom_smooth(aes(date, trelative), span = "1 month") +
  scale_x_date(date_breaks = "years", date_labels = "%Y") +
  labs(x = element_blank(), 
       y = "Time since first archival (days)",
       title = "Tendency of time to update the package") +
  theme_minimal()
```


```{r}
a <- all_packages2 |> filter(n !=1) |> group_by(package) |> summarise(d = (max(date) -min(date)), m = d/max(n),  n = max(n))
ggplot(a, aes(d, n)) + geom_point() +
  geom_abline(slope = c(1/30, 1/15), intercept = 0, col = "red")
```

```{r}
block <- all_packages2 |> 
  filter(release != 1) |> 
  group_by(month = floor_date(date, "halfyear"),
           release) |> 
  summarise(t = median(tprevious),
            n = n()) |> 
  ungroup() 

block |> 
  group_by(release) |> 
  summarize(n_dates = n_distinct(month),
            n_median = median(n)) |> 
  ungroup() |> 
  pull(n_dates) |> 
  summary()

block |> 
  group_by(release) |> 
  summarize(n_dates = n_distinct(month),
            n_median = median(n)) |> 
  ungroup() |> 
  filter(n_dates > 20) |> 
  pull(release) |> 
  max()
```


```{r}
block |> 
  filter(release <= 6) |> 
  ggplot() +
  geom_smooth(aes(month, n, size = n, col = as.factor(release),
                  linetype = as.factor(release), weight = n), 
              alpha = 0.25)  +
  scale_x_date(expand = expansion(), date_breaks = "2 year", date_labels = "%Y") +
  scale_y_continuous(expand = expansion()) +
  coord_cartesian(ylim = c(0, NA)) +
  labs(x = element_blank(),
       y = "Packages",
       title = "Updates",
       col = "Releases",
       linetype = "Releases") +
  theme_minimal()
```


```{r}
block |> 
  filter(release <= 6) |> 
  ggplot() +
  geom_smooth(aes(month, t, size = n, col = as.factor(release),
                  linetype = as.factor(release), weight = n), 
              alpha = 0.25)  +
  scale_x_date(expand = expansion(), date_breaks = "2 year", date_labels = "%Y") +
  scale_y_continuous(expand = expansion()) +
  coord_cartesian(ylim = c(0, NA)) +
  labs(x = element_blank(),
       y = "Median time since previous release (weeks)",
       title = "Speed of updates",
       col = "Releases",
       linetype = "Releases") +
  theme_minimal()
  
```

The reasons of this longer updates might be multiple, it might be that packages have become better and require less updates to keep up with the more strict CRAN rules.
It could also mean that maintainers are no longer updating packages and abandoning them. 
This has been explore in a previous post about [packages archived by CRAN](https://llrs.dev/post/2021/12/07/reasons-cran-archivals/)  (using another file)

## Seasonal effects {#season}

TODO: really use seasonal effects like this one https://stackoverflow.com/questions/46797884/can-i-estimate-a-time-varying-seasonal-effect-in-r-with-gamm
or this one
https://www.idrisstsafack.com/post/arma-models-with-r-the-ultimate-practical-guide-with-bitcoin-data
OR: https://www.datascienceinstitute.net/blog/time-series-decomposition-in-r

```{r}
dates_approval <- all_packages2 |> 
  mutate(year = year(date),
         month = month(date),
         week = week(date),
         yday = yday(date),
         mday = mday(date),
         wday = wday(date, week_start = 1, label = FALSE),
         type = ifelse(release == 1, "new", "update"))

d <- c(1:7)
lab_names <- lubridate::wday(d, week_start = 1, label = TRUE)
names(lab_names) <- as.character(d)
dmonth_breaks <- seq(0, 31, by = 7)
dmonth_breaks[1] <- 1

dates_approval |> 
  count(type, year, month) |> 
  ggplot() +
  geom_smooth(aes(month, n, col = type, linetype = type)) +
  theme_minimal()
```


```{r}
dates_approval |>  
  count(type, year, week) |> 
  ggplot() +
  geom_smooth(aes(week, n, col = type, linetype = type)) +
  theme_minimal()
```


```{r}
dates_approval |>  
  count(type, year, month, week, wday) |> 
  ggplot() +
  geom_smooth(aes(wday, n, col = type, linetype = type), method = "glm") +
  coord_cartesian(ylim = c(0, 21)) +
  scale_x_continuous(expand = expansion(), 
                     labels = lab_names, position = "top")
```


```{r}
dates_approval |> 
  count(type, year, month, mday) |> 
  ggplot() +
  geom_smooth(aes(mday, n, col = type, linetype = type)) +
  coord_cartesian(ylim = c(0, 21)) +
  scale_x_continuous(expand = expansion(), breaks = dmonth_breaks,
                       position = "top") +
  scale_y_continuous(expand = expansion())
```


```{r}
dates_approval |> 
  count(type, year, yday) |> 
  ggplot() +
  geom_smooth(aes(yday, n, col = type, linetype = type)) + 
  annotate(geom = "text", x = 100, y = 12, label = "April") +
  annotate(geom = "text", x = 200, y = 12, label = "July") +
  annotate(geom = "text", x = 300, y = 12, label = "November") +
  coord_cartesian(ylim = c(0, 21))
```


At the end of the year, beginning of the next there the submission queue is closed, [for example in 2017](https://stat.ethz.ch/pipermail/r-devel/2017-December/075215.html).
I also [collected](https://github.com/llrs/blogR/issues/57) other dates when the submission queue was closed.
There doesn't seem to be a regular site to announce this, sometimes is on the home page of CRAN sometimes is announced on the mailing lists. 

This doesn't include some [unexpected downtime](https://stat.ethz.ch/pipermail/r-devel/2022-May/081751.html), which recently happened for the first time in my short exposure. 


### Admins

CRAN is not run by bots, some volunteers with their own work and hobbies 

```{r}
people <- dates_approval |> 
  count(uname) |> 
  filter(n > 1000) |> 
  pull(uname)
dates_approval |> 
  filter(uname %in% people) |>
  group_by(uname) |> 
  summarise(min = min(date),
            max = max(date),
            n = n()) |> 
  arrange(min)
```

We can see that there are three users active for more than 10 years

```{r}
dates_approval |> 
  count(year, uname) |> 
  ggplot() +
  geom_line(aes(year, n, 
                col = fct_other(uname, keep = people, other_level = "other"))) +
  labs(col = "Users") +
  theme_minimal() +
  theme(legend.position = c(0.3, 0.7))
  scale_colour_viridis_d()
```


There are some patterns here on people volunteering on CRAN.
Uwe Ligges has been doing the shear work of accepting and updating the packages since 2015.
Initially it was done by other volunteers, but lately they don't seem to perform this action.
Notable is the Kurt Hornik case were he did most of the work around 2010, later reduced the activity and came back before 2015 to stop shortly after.
Probably he was doing the same under the root username 
Brian Ripely was very active between 2010 and 2015.

Previous Volunteers


```{r}
dates_approval |> 
  count(year, month, uname) |> 
  ggplot() +
  geom_smooth(aes(month, n, col = fct_other(uname, keep = people, other_level = "other"))) +
  labs(col = "Users")
```


```{r}
dates_approval |>  
  count(year, week, uname) |> 
  ggplot() +
  geom_smooth(aes(week, n, col = fct_other(uname, keep = people, other_level = "other"))) +
  labs(col = "Users")
  facet_wrap(~type, scales = "free")
```


```{r}
dates_approval |>  
  count(year, month, week, wday, uname) |> 
  ggplot() +
  geom_smooth(aes(wday, n,  col = fct_other(uname, keep = people, other_level = "other")), method = "glm") + 
  labs(col = "Users")
```


```{r}
dates_approval |> 
  count(year, month, mday, uname) |> 
  ggplot() +
  geom_smooth(aes(mday, n, col = fct_other(uname, keep = people, other_level = "other"))) +
  labs(col = "Users") +
  coord_cartesian(ylim = c(0, NA)) +
  scale_x_continuous(expand = expansion(), breaks = dmonth_breaks,
                       position = "top") +
  scale_y_continuous(expand = expansion())
```


```{r}
dates_approval |> 
  count(year, yday, uname) |> 
  ggplot() +
  geom_smooth(aes(yday, n, col = fct_other(uname, keep = people, other_level = "other"))) + 
  labs(col = "Users") +
  annotate(geom = "text", x = 100, y = 12, label = "April") +
  annotate(geom = "text", x = 200, y = 12, label = "July") +
  annotate(geom = "text", x = 300, y = 12, label = "November") +
  coord_cartesian(ylim = c(0, NA))
```


### Holidays

Given the height rate of package added and updated in CRAN let's see days without acceptance of packages. 
First some transformation:

```{r}
d1 <- min(all_packages2$date)
d2 <- max(all_packages2$date)
s <- seq(from = d1, to = d2, by = 1)
holiday <- s[!s %in% unique(all_packages2$date)]
holiday <- data.frame(holiday = holiday) |> 
  mutate(year = year(holiday),
         month = month(holiday),
         week = week(holiday),
         yday = yday(holiday),
         mday = mday(holiday),
         wday = wday(holiday, week_start = 1, label = FALSE))
```

Then some plots by week:

```{r}
holiday |>  
  count(year, week) |> 
  ggplot() +
  geom_smooth(aes(week, n)) +
  labs(title = "Weeks without packages updates",
       y = "Years")
```

Similarly we can see the days of the year without additions:

```{r}
holiday |>  
  count(year, yday) |> 
  ggplot() +
  geom_count(aes(yday, n))  +
  labs(title = "Weeks without packages updates",
       y = "Years")
```

Clearly at the end and beginning of the year there is a tendency to not accept packages.
Which is usually when the CRAN closes the submission queue for holidays. 

## Oldest package {#oldest}

Given the archives and the current packages on CRAN.
Which are the oldest packages?
I'm not sure if this date is the archival date or date of the first publication. 
It could be that the release date was earlier and the list might be a bit off. 

```{r cran-oldest}
all_packages2 |> 
  filter(available == "Yes") |> 
  filter(release == 1) |> 
  arrange(date) |> 
  select(package, version, size, date) |> 
  head(10)
```

Which are the oldest packages without any update?

```{r cran-oldest-conserved}
all_packages2 |> 
  filter(available == "Yes") |> 
  group_by(package) |> 
  filter(max(release) == 1) |> 
  ungroup() |> 
  arrange(date) |> 
  select(package, version, size, date) |> 
  head(10)
```

These are the oldest 10 packages released and without any further release on CRAN. 
They are around 10 years old.

But the packages that have been longest on CRAN, be it the first release or not are these:

```{r cran-longest3}
all_packages2 |> 
  filter(available == "Yes" & status == "available") |> 
  arrange(date) |> 
  select(package, version, size, date, release) |> 
  head(10)
```

A couple of the top 10 repeats from those without any update!
They don't fall far from the oldest packages!

Let's show the initial releases of packages:

```{r cran-initial-releases}
never_archived <- all_packages2 |> 
  filter(available == "Yes") |> 
  group_by(package) |> 
  filter(max(release) == 1) |> 
  ungroup() |> 
  pull(package)

all_packages2 |> 
  filter(available == "Yes" & release == 1) |> 
  group_by(month = floor_date(date, "month"),
           same = package %in% never_archived) |> 
  count() |> 
  ungroup() |> 
  ggplot(aes(month, n, fill = yesno(same))) +
  geom_col(position = "identity", alpha = 0.5) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  theme_minimal() +
  labs(fill = "Updated?",
       x = element_blank(),
       y = "Packages",
       title = "First archival or release date") +
  theme(legend.position = c(0.3, 0.8))
```


## Dependencies current packages {#dependencies}

Looking at packages with just 1 release but still one package which are those?

This shows the date of the packages in relation to their recursive dependencies.

```{r}
direct_dep <- tools::package_dependencies(packages2$Package, 
                                          db = packages2)

base_packages <- rownames(installed.packages(priority = "base"))
deps0 <- vapply(direct_dep, function(x){sum(!x %in% base_packages)}, numeric(1L))
deps <- lengths(direct_dep)
recurs_dep <- tools::package_dependencies(packages2$Package, 
                                          db = packages2, recursive = TRUE)
r_deps0 <- vapply(recurs_dep, function(x){sum(!x %in% base_packages)}, numeric(1L))


r_deps <- lengths(recurs_dep)
r_deps <- r_deps0[names(deps0)]
deps_all <- data.frame(direct_deps = deps0,
                       r_deps)
deps_all$Package <- rownames(deps_all)
deps_all |> 
  ggplot() +
  geom_count(aes(direct_deps, r_deps, col = after_stat(n))) +
  scale_size(trans = "log10") +
  scale_color_viridis_c(trans = "log10", begin = 0, end = 1, direction = -1) +
  scale_y_log10() +
  labs(x = "Direct", y = "Indirect",
       size = "Packages",
       col = "Packages",
       title = "Relationship between direct and indirect dependencies") +
  theme_minimal()
```


```{r}
packages2 |> 
  merge(deps_all, by = "Package", all = TRUE) |> 
  mutate(archived = yesno(Package %in% never_archived)) |> 
  select(date, direct_deps, r_deps, archived) |> 
  pivot_longer(c(direct_deps, r_deps), names_to = "dependency") |> 
  mutate(dependency = ifelse(dependency == "direct_deps", "direct", "recursive")) |> 
  ggplot() +
  geom_smooth(aes(date, value, col = archived, linetype = dependency), span = 1/24) +
  scale_y_log10(limits = c(1, NA), expand = expansion()) +
  scale_x_date(expand = expansion(), date_breaks = "2 years",
               date_labels = "%Y") +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(x = element_blank(),
       col = "Updated?",
       y = "Dependencies",
       linetype = "Type of dependencies",
       title = "More dependencies lead to archive?",
       subtitle = "Packages dependencies, without base packages by date published on CRAN.") +
  theme(legend.position = c(0.3, 0.8))
```


The oldest packages on CRAN which were never updated have similar direct dependencies to those with archives but with more indirect dependencies.  
More recent packages have more direct dependencies and even more indirect dependencies. 
This can also lead to a different lecture, to avoid problems packages reduce their (direct) dependencies. 

Without knowing the history of dependencies of CRAN or what actually lead to being archived it might be hard to solve this question. 

### Date

When I asked online about more question Jonathan Carroll suggested to look up if we are using newer packages or more of the older ones:

```{r echo=FALSE}
blogdown::shortcode('tweet', '1533744129695330304')
```

```{r}
dates_deps <- function(dependencies, dates, base_packages = NULL) {
  if (is.null(base_packages)) {
    base_packages <- rownames(installed.packages(priority = "base"))
  }
  deps <- dependencies[!dependencies %in% base_packages]
  d <- dates[deps]
  if (length(d) == 0) {
    return(data.frame(min = NA_Date_, median = NA_Date_, max = NA_Date_, n = 0))
  }
  data.frame(min = min(d, na.rm = TRUE), 
             median = median(d, na.rm = TRUE), 
             max = max(d, na.rm = TRUE),
             n = length(d))
}

first_release_dates <- all_packages2 |> 
  group_by(package) |> 
  summarise(date = min(date))
dates <- first_release_dates$date
names(dates) <- first_release_dates$package
l <- lapply(direct_dep, dates_deps, dates, base_packages)
# First date of releas of a package and current date of publication of packages
deps_dates <- do.call("rbind", l) 
deps_dates$package <- names(direct_dep)
deps_dates <- left_join(deps_dates, packages2, by = c("package" = "Package"))

deps_dates |> 
  pivot_longer(cols = c(min, median, max)) |> 
  filter(!is.na(value)) |> 
  ggplot() +
  geom_smooth(aes(date, value, weight = n, col = name))+
  geom_abline(slope = 1, intercept = 0) +
  labs(x = "Published date",
       y = "Dependencies date",
       title = "Min, median and max date of direct direct dependencies") +
  theme_minimal()
```


```{r}
deps_dates |> 
  pivot_longer(cols = c(min, median, max)) |> 
  mutate(difftime = as.numeric(difftime(date, value, units = "weeks"))) |> 
  mutate(difftime = if_else(is.na(difftime), 0, difftime)) |> 
  ggplot() +
  geom_smooth(aes(date, difftime, col = name)) +
  geom_abline(slope = 1/52, intercept = 0) +
  scale_y_continuous(sec.axis = sec_axis( ~ ./52, 
                                          name = "Dependencies years of difference")) +
  scale_x_continuous(expand = expansion()) +
  labs(x = "Published date",
       y = "Dependencies weeks of difference",
       title = "Min, median and max date of direct dependencies") +
  theme_minimal()
```


```{r}
deps_dates |> 
  pivot_longer(cols = c(min, median, max)) |> 
  mutate(difftime = as.numeric(difftime(date, value, units = "weeks"))) |> 
  ggplot() +
  geom_smooth(aes(date, difftime, col = name, linetype = yesno(with_archives))) +
  geom_abline(slope = 1/52, intercept = 0) +
  scale_y_continuous(sec.axis = sec_axis( ~ ./52, 
                                          name = "Dependencies years of difference")) +
  scale_x_date(expand = expansion()) +
  facet_wrap(~ifelse(with_archives, "New", "Already on CRAN"), scales = "free_x") +
  labs(x = "Published date",
       y = "Dependencies weeks of difference",
       linetype = "Already on CRAN?",
       col = "Dependencies date",
       title = "Min, median and max date of direct dependencies") +
  theme_minimal() +
  theme(panel.spacing = unit(1, "lines"))
```


There is around a median of 10 year gap between publishing a package and getting a package depending on it according to the first release of packages currently on CRAN.
At earliest there is a median of 6 year gap and sometimes dependencies are up to 16 years older than the new package.
For reference a new packages today depending, importing or enhancing dplyr, which first release was in 2014-01-29, is around 8 years and depending on the first packages on CRAN means at most 25 years of release differences.

If we take the dependency with the less difference on each day we see the fastest dependency tend to be 2 years

```{r}
deps_dates |> 
  pivot_longer(cols = c(min, median, max)) |> 
  mutate(difftime = as.numeric(difftime(date, value, units = "weeks"))) |> 
  filter(name == "max") |> 
  group_by(date, with_archives) |> 
  summarise(value = min(date, na.rm = TRUE),
            difftime = min(difftime, na.rm = TRUE)) |> 
  ungroup() |> 
  ggplot() +
  geom_smooth(aes(date, difftime, linetype = yesno(with_archives))) +
  geom_abline(slope = 1/52, intercept = 0) +
  scale_y_continuous(sec.axis = sec_axis( ~ ./52, 
                                          name = "Dependencies years of difference")) +
  scale_x_date(expand = expansion()) +
  # facet_wrap(~ifelse(with_archives, "New", "Already on CRAN"), scales = "free_x") +
  labs(x = "Published date",
       y = "Dependencies weeks of difference",
       linetype = "Already on CRAN?",
       col = "Dependencies date",
       title = "Min, median and max date of direct dependencies") +
  theme_minimal() +
  theme(panel.spacing = unit(1, "lines"), legend.position = c(0.3, 0.4))
```


### Looking by day

The plots above are about the packages published that date.
It can be confusing but it is not the median time difference of the day.

If we summarise the time difference we can do the difference by min value, max value and median.

<details><summary>If we take the minimum difference on each day:</summary>


```{r}
deps_dates |> 
  pivot_longer(cols = c(min, median, max)) |> 
  mutate(difftime = as.numeric(difftime(date, value, units = "weeks"))) |> 
  group_by(date, name, with_archives) |> 
  summarise(value = min(date, na.rm = TRUE),
            difftime = min(difftime, na.rm = TRUE)) |> 
  ungroup() |> 
  ggplot() +
  geom_smooth(aes(date, difftime, col = name, linetype = yesno(with_archives))) +
  geom_abline(slope = 1/52, intercept = 0) +
  scale_y_continuous(sec.axis = sec_axis( ~ ./52, 
                                          name = "Dependencies years of difference")) +
  scale_x_date(expand = expansion()) +
  facet_wrap(~ifelse(with_archives, "New", "Already on CRAN"), scales = "free_x") +
  # facet_zoom(x = date >= as.Date("2021/01/01"),
  #            xlim = c(as.Date("2021/01/01"), Sys.Date())) +
  labs(x = "Published date",
       y = "Dependencies weeks of difference",
       linetype = "Already on CRAN?",
       col = "Dependencies date",
       title = "Min, median and max date of direct dependencies") +
  theme_minimal() +
  theme(panel.spacing = unit(1, "lines"))
```
</details>

<details><summary>If we take the maximum difference on each day:</summary>


```{r}
deps_dates |> 
  pivot_longer(cols = c(min, median, max)) |> 
  mutate(difftime = as.numeric(difftime(date, value, units = "weeks"))) |> 
  group_by(date, name, with_archives) |> 
  summarise(value = max(date, na.rm = TRUE),
            difftime = max(difftime, na.rm = TRUE)) |> 
  ungroup() |> 
  ggplot() +
  geom_smooth(aes(date, difftime, col = name, linetype = yesno(with_archives))) +
  geom_abline(slope = 1/52, intercept = 0) +
  scale_y_continuous(sec.axis = sec_axis( ~ ./52, 
                                          name = "Dependencies years of difference")) +
  scale_x_date(expand = expansion()) +
  facet_wrap(~ifelse(with_archives, "New", "Already on CRAN"), scales = "free_x") +
  # facet_zoom(x = date >= as.Date("2021/01/01"),
  #            xlim = c(as.Date("2021/01/01"), Sys.Date())) +
  labs(x = "Published date",
       y = "Dependencies weeks of difference",
       linetype = "Already on CRAN?",
       col = "Dependencies date",
       title = "Min, median and max date of direct dependencies") +
  theme_minimal() +
  theme(panel.spacing = unit(1, "lines"))
```

</details>

With the median daily difference:

```{r}
deps_dates |> 
  pivot_longer(cols = c(min, median, max)) |> 
  mutate(difftime = as.numeric(difftime(date, value, units = "weeks"))) |> 
  group_by(date, name, with_archives) |> 
  summarise(value = median(date, na.rm = TRUE),
            difftime = median(difftime, na.rm = TRUE),
            n = n()) |> 
  ungroup() |> 
  ggplot() +
  geom_smooth(aes(date, difftime, col = name, linetype = yesno(with_archives),
                  weight = n)) +
  geom_abline(slope = 1/52, intercept = 0) +
  scale_y_continuous(sec.axis = sec_axis( ~ ./52, 
                                          name = "Dependencies years of difference")) +
  scale_x_date(expand = expansion()) +
  facet_wrap(~ifelse(with_archives, "New", "Already on CRAN"), scales = "free_x") +
  # facet_zoom(x = date >= as.Date("2021/01/01"),
  #            xlim = c(as.Date("2021/01/01"), Sys.Date())) +
  labs(x = "Published date",
       y = "Dependencies weeks of difference",
       linetype = "Already on CRAN?",
       col = "Dependencies date",
       title = "Daily time difference of direct dependencies") +
  theme_minimal() +
  theme(panel.spacing = unit(1, "lines"))
```

We can see that in 2012 new packages tended to depend on newer packages than now, while now they tend to depend around 12 years and close to 6 years at least.
Reminder that this is only for those with dependencies outside base packages.

Looking at the number of packages without dependencies we see a constant number o

```{r}
deps_dates |> 
  filter(n == 0) |> 
  count(date, with_archives) |> 
  ggplot() +
  geom_smooth(aes(date, n, 
                  col = yesno(!with_archives), linetype = yesno(!with_archives)),
              span = "1 month") +
  # facet_wrap(~with_archives, scales = "free") +
  scale_y_continuous(expand = expansion(add = c(NA_integer_, NA_integer_))) +
  scale_x_date(expand = expansion()) +
  labs(col = "New on CRAN?", linetype = "New on CRAN?", y = "Packages",
       title = "Packages without dependencies") +
  theme_minimal() +
  theme(legend.position = c(0.2, 0.7))
```



```{r eval=FALSE, include=FALSE}
deps_dates |> 
  pivot_longer(cols = c(min, median, max)) |> 
  mutate(difftime = as.numeric(difftime(date, value, units = "weeks"))) |> 
  group_by(name) |>
  summarise(weeks = median(difftime, na.rm = TRUE),
            months = weeks/4,
            years = weeks/52) |> 
  ungroup()
deps_dates |> 
  pivot_longer(cols = c(min, median, max)) |> 
  mutate(difftime = as.numeric(difftime(date, value, units = "weeks"))) |> 
  group_by(with_archives, name) |>
  summarise(weeks = median(difftime, na.rm = TRUE),
            months = weeks/4,
            years = weeks/52) |> 
  ungroup()
```


### Reproducibility

<details>

```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info()
```

</details>

