---
title: Leadershin in open source projects
author: Lluís Revilla Sancho
date: '2022-05-27'
slug: leadership-open-source
categories:
  - OSS
tags:
  - concepts
authors: []
description: ''
draft: no
editor_options:
  chunk_output_type: console
featured: no
image:
  caption: ''
  focal_point: ''
subtitle: ''
summary: ''
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.retina = 2)
```

This is the second part of the previous post about [concepts around open source/free software](https://llrs.dev/post/2022/03/16/concepts-around-open-source-free-software/).

Since last book, I also read [Uncurled](https://un.curl.dev) which recommended ["The Making and maintenance of open source software"](https://www.amazon.com/Working-Public-Making-Maintenance-Software/dp/0578675862) (I will short it to MMOSS) and "[Producing Open Source Software](https://producingoss.com/)" (POSS), which provide first hand experiences about how those maintainers handled different projects.

This post is about projects, growth and management of the project (I don't want to make them about R but I will also use my experience with different project in R to reflect on these ideas).

## Projects

It is common to say managing open source projects.
While it is true that open source/free software is a project with limited resources it doesn't have a fixed end date.
I don't think they fit the criteria of the [PMBOK](https://en.wikipedia.org/wiki/Project_Management_Body_of_Knowledge "Project Management Body of Knowledge").

Software projects run from the usefulness for the maintainer and the feedback received.
If a maintainer does no longer use a piece of software and doesn't receive any feedback they might choose to not update the public version.

This explains why it is so common to look up for activity on a "project".
If a developer/maintainer hasn't commited or reacted in a long time the "project" is no longer active.
The code might still be used, it might have more users, but it is no longer a project.

Another characteristics of projects is that there is a leadership.
It might be a team or an individual like a benevolent dictator for life as explained by POSS.
This leadership comes with communication.
It is not enough to have commit access or a foundation but to communicate and steward the community towards a direction or following some guidelines.

These leaders are those who can steward with contributor growth, support, ease of participation and technical improvements.
Particularly POSS mentions the distinction between commit access and maintainer.
Considering also maintainers those that organize conferences, set up infrastructure and participate even if not committing code.

All the activities of the maintainers help with the project status and growth.

## Growth?

There seems to be in the MMOSS book a tendency to only mention growth, which is very hard to measure.
Just fo give an example focusing on users: Is LAPACK user base growing or not?
I would say it is, because R uses it and more people is using R.
But there isn't, to my knowledge, a growth on users claiming to be experts on LAPACK, although there are several LAPACK optimizations.

About the contributors growth MMOSS also mentions that (long term) contributors tend to lurk and just chime in when they can propose something of value to the project or given the room.
Casual contributors post when they have an issue/bug while using it but they are not interested on the long term health of the software.
Presumably on the classification on table 1 of the [previous post](https://llrs.dev/post/2022/03/16/concepts-around-open-source-free-software/) refers to both.
As acknowledge by the author it is hard to measure users and contributors.

Probably the best indicatior is the relationship between users and contributors.
But the books always focus on the very positive outcome growth, not stagnation or reduction of users/contributors.
I think it is important to acknowledge this happens too and find the causes to avoid them.

On python the support of python 2.7 had to be extended multiple times due to slow uptake of python 3.x.
It might have similar effects than the upgrade from perl4 to perl5, which (besides many other reasons) left perl outside the curriculum of the universities, so without many new users.

## Code characteristics

MMOSS mentions motivation as a characteristic of developing code.
This might be a reason why OSS is usually seen as dominated by rich males.
Usually intrinsic motivation to give away time and code

### Reproducibility

<details>

```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info()
```

</details>
