---
title: CRAN maintained packages
author: Lluís Revilla Sancho
date: '2023-05-03'
slug: cran-maintained-packages
categories:
  - CRAN
tags:
  - CRAN
  - packages
authors:
  - admin
description: 'Looking at XML and RCurl impact'
draft: no
editor_options:
  chunk_output_type: console
featured: no
image:
  caption: ''
  focal_point: ''
subtitle: ''
summary: 'Some analysis on the RCurl and XML dependencies.'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.retina = 2, message = FALSE, cache = TRUE)
library("knitr")
knit_print.data.frame = function(x, ...) {
  res = paste(c('', '', kable(x)), collapse = '\n')
  asis_output(res)
}
knit_print.tbl_df = function(x, ...) {
  res = paste(c('', '', kable(x)), collapse = '\n')
  asis_output(res)
}
knit_print.matrix = function(x, ...) {
  res = paste(c('', '', kable(as.data.frame(x))), collapse = '\n')
  asis_output(res)
}
# register the method
registerS3method("knit_print", "data.frame", knit_print.data.frame)
registerS3method("knit_print", "tbl_df", knit_print.tbl_df)
registerS3method("knit_print", "matrix", knit_print.matrix)
```

The role of package managers in software is paramount for developers.
In R the CRAN team provides a platform to tests and host packages.
This means ensuring that R dependencies are up to date and software required by some packages are also available in CRAN.

This helps testing \~20000 packages frequently (daily for most packages) in several architectures and R versions.
In addition, they test updates for compatibility with the dependencies and test and review new packages.

Most of the work with packages is automated but often requires human intervention ([50% of the submisions](https://journal.r-project.org/news/RJ-2022-4-cran/#cran-package-submissions)).
Another consuming activity is keeping up packages abandoned by their original maintainers.

While newer packages are [archived from CRAN often](https://llrs.dev/post/2021/12/07/reasons-cran-archivals/), some old packages were adopted by CRAN.
The [CRAN team](https://cran.r-project.org/CRAN_team.htm) is [looking for help](https://mastodon.social/@henrikbengtsson/110186925898457474) maintining those.

In this post I'll explore the packages maintained by CRAN.

# CRAN in packages

```{r cran_authors}
packages_db <- as.data.frame(tools::CRAN_package_db())
cran_author <- grep("CRAN Team", x = packages_db$Author, ignore.case = TRUE)
cran_authorsR <- grep("CRAN Team", x = packages_db$`Authors@R`, ignore.case = TRUE)
CRAN_TEAM_mentioned <- union(cran_author, cran_authorsR)
unique(packages_db$Package[CRAN_TEAM_mentioned])
```

In some of these package the CRAN team appears as contributors because they provided help/code to fix bugs:

-   [geiger](https://cran.r-project.org/package=geiger)
-   [fMultivar](https://cran.r-project.org/package=fMultivar)
-   [fBasics](https://cran.r-project.org/package=fBasics)
-   [udunits2](https://cran.r-project.org/package=udunits2)

In others they are the maintainers:

-   [XML](https://cran.r-project.org/package=XML)
-   [RCurl](https://cran.r-project.org/package=RCurl)
-   [RJSONIO](https://cran.r-project.org/package=RJSONIO)

From these three packages RJSONIO is the newest (first release in 2010) and requires less updates (lately 1 or 2 a year).
However, in 2022 RCurl and XML required 4 and 5 updates respectively.
I will focus on these packages as these are the ones they are looking for new maintainers.

# RCurl and XML

## Circular dependency

Both XML and RCurl depend on each other.

We can see that the packages are direct dependencies of one of their direct dependencies!
How can be that?
If we go the the [RCurl](https://cran.r-project.org/package=RCurl) website we see in "Suggests: XML", and in the [XML](https://cran.r-project.org/package=XML) website the RCurl is there too.
This circular dependency is allowed because they have each other in Suggests.

A first step to reduce any possible problem would be to separate them.
This would make it easier understanding which package is worth prioritizing and possible missteps will have less impact.

If we look at [XML source code for RCurl we find](https://github.com/search?q=repo%3Acran%2FXML%20RCurl&type=code) some code in `inst/` folder.
If these two cases were removed the package could remove its dependency to RCurl.

Similarly, if we look at [RCurl source code for XML we find](https://github.com/search?q=repo%3Acran%2FRCurl%20XML&type=code) some code in `inst/` folder and in some examples.
If these three cases were removed the package could remove its dependency to XML.

RCurl has been [more stable](https://diffify.com/R/RCurl/1.95-4.9/1.98-1.12) than XML, which have seen [new functions added and one removed](https://diffify.com/R/XML/3.98-1.7/3.99-0.14) since CRAN is maintaining it.

## Relevant data

We will look at 4 sets of data for each pacakge: [dependencies], [releases], [maintainers] and [downloads].

### Dependencies

Both packages have some system dependencies which might make the maintenance harder.
In addition they have a large number of dependencies.
We can gather the dependencies in CRAN and Bioconductor software packages:

```{r deps}
library("tools")
# Look up only software dependencies in Bioconductor
options(repos = BiocManager::repositories()[c("BioCsoft", "CRAN")])
ap <- available.packages()
all_deps <- package_dependencies(c("RCurl", "XML"), 
                                 reverse = TRUE, db = ap, which = "all")
all_unique_deps <- unique(unlist(all_deps, FALSE, FALSE))
first_deps <- package_dependencies(all_unique_deps, db = ap, which = "all")
first_deps_strong <- package_dependencies(all_unique_deps, db = ap, which = "strong")
strong <- sapply(first_deps_strong, function(x){any(c("XML", "RCurl") %in% x)})
deps_strong <- package_dependencies(all_unique_deps, recursive = TRUE, 
                                 db = ap, which = "strong")
first_rdeps <- package_dependencies(all_unique_deps, 
                                   reverse = TRUE, db = ap, which = "all")
deps_all <- package_dependencies(all_unique_deps, recursive = TRUE, 
                                 db = ap, which = "all")
```

They have `r length(all_unique_deps)` direct dependencies (and 8 more in annotation packages in Bioconductor: recount3, ENCODExplorerData, UCSCRepeatMasker, gDNAinRNAseqData, qdap, qdapTools, metaboliteIDmapping and curatedBreastData).

These two packages with their dependencies are used one way or another by around 20000 packages (about 90% of CRAN and Bioconductor)!
If these packages fail the impact on the community will be huge.

To reduce the impact of the dependencies we should look up the direct dependencies.
But we also looked at the reverse dependencies to asses the impact of the package in the other packages. 

Know which are these, and who maintain them will help decide what is the best course of action.

### Releases

A first approach is looking into the number of releases and dates to asses if the package has an active maintainer or not:

```{r releases}
archive <- tools:::CRAN_archive_db()[all_unique_deps]
packages <- tools::CRAN_package_db()
library("dplyr")
library("BiocPkgTools")
fr <- vapply(archive, function(x) {
  if (is.null(x)) {
    return(NA)
  }
  as.Date(x$mtime[1])
}, FUN.VALUE = Sys.Date())
fr_bioc <- biocDownloadStats("software") |> 
  filter(Package %in% all_unique_deps) |> 
  firstInBioc() |> 
  pull(Date, name = Package)
first_release <- c(as.Date(fr[!is.na(fr)]), as.Date(fr_bioc))[all_unique_deps]
last_update <- packages$Published[match(all_unique_deps, packages$Package)]
releases <- vapply(archive, NROW, numeric(1L)) + 1
```

We only have information about CRAN packages:\
Bioconductor has two releases every year, and while the maintainers can release patched versions of packages between them that information is not stored (or easily retrieved, they are still available in the [git server](https://code.bioconductor.org)).

Even if Bioconductor maintainers didn't modify the package the version number increases with each release.
But the version update in the git doesn't propagate to users automatically unless their checks pass.
For all these reasons it doesn't make sense to count releases of packages in Bioconductor.

### Maintainers

Now that we know which packages are more active, we can look up for the people behind it.
This way we can prioritize working with maintainers that are known to be active[^1].

[^1]: the `maintainer` function only works for installed packages, and I don't have all these packages installed.

```{r maintainers}
maintainers <- packages_db$Maintainer[match(all_unique_deps, packages_db$Package)]
maintainers <- trimws(gsub("<.+>", "", maintainers))
```

Once again, the Bioconductor repository doesn't provide a file to gather this kind of data.

### Downloads

Another variable we can use are the downloads from users of said packages.
Probably, packages more downloaded are used more and a breaking change on them will have impact on more people than other packages.

```{r downloads}
library("cranlogs")
acd <- cran_downloads(intersect(all_unique_deps, packages_db$Package), 
                          when = "last-month")
cran_pkg <- summarise(acd, downloads = sum(count), .by = package)
loc <- Sys.setlocale(locale = "C")
bioc_d <- vapply(setdiff(all_unique_deps, packages_db$Package), function(x){
  pkg <- pkgDownloadStats(x)
  tail(pkg$Nb_of_downloads, 1)
  }, numeric(1L))
bioc_pkg <- data.frame(package = names(bioc_d), downloads = bioc_d)
downloads <- rbind(bioc_pkg, cran_pkg)
rownames(downloads) <- downloads$package
dwn <- downloads[all_unique_deps, ]
```

The logs are provided by the global mirror of CRAN (sponsored by Rstudio).  
The Bioconductor infrastructure which provides total number of downloads and number of downloads from distinct IPs [^3].

[^3]: Both logs only count those of their repository and not from other mirrors or approaches (RSPM, bspm, r2u, ....).

## Analysis

We collected the data that might be relevant.
Now, we can start looking all the data gathered:

```{r all-together}
repo <- vector("character", length(all_unique_deps))
ap_deps <- ap[all_unique_deps, ]
repo[startsWith(ap_deps[, "Repository"], "https://bioc")] <- "Bioconductor"
repo[!startsWith(ap_deps[, "Repository"], "https://bioc")] <- "CRAN"
deps <- data.frame(package = all_unique_deps,
                   direct_dep_XML = all_unique_deps %in% all_deps$XML,
                   direct_dep_RCurl = all_unique_deps %in% all_deps$RCurl,
                   first_deps_n = lengths(first_deps),
                   deps_all_n = lengths(deps_all),
                   first_rdeps_n = lengths(first_rdeps),
                   first_deps_strong_n = lengths(first_deps_strong), 
                   deps_strong_n = lengths(deps_strong),
                   direct_strong = strong, 
                   releases = releases,
                   strong = strong, 
                   first_release = first_release,
                   last_release = last_update,
                   maintainer = maintainers,
                   downloads = dwn$downloads,
                   repository = repo) |> 
  mutate(type = case_when(direct_dep_XML & direct_dep_RCurl ~ "both",
                          direct_dep_XML ~ "XML",
                          direct_dep_RCurl ~ "RCurl"))
rownames(deps) <- NULL
head(deps)
```

I added some numbers and logical values that might help exploring this data.

We will look at the [packages dependencies between RCurl and XML](#distribution-dependencies), some plots to have a [quick view](#overview)

### Distribution dependencies 

Let's see how many packages depend in each of them:

```{r direct-deps}
deps |> 
  summarise(Packages = n(), deps = sum(first_deps_n),
            q25 = quantile(deps_all_n, probs = 0.25),
            mean_all = mean(deps_all_n),
            q75 = quantile(deps_all_n, probs = 0.75),
            .by = c(direct_dep_XML, direct_dep_RCurl)) |> 
  arrange(-Packages)
```

There are \~40 more packages depending on XML than to RCurl and just `r sum(deps$direct_dep_XML & deps$direct_dep_RCurl)` to both of them.

### Overview

We can plot some variables to get a quick overview of the packages:

```{r plot1}
#| fig.cap = "Direct dependencies vs downloads. Many pakcages have up to 50 packages and most have below 1000 downloads in a month."
library("ggplot2")
library("ggrepel")
deps_wo <- filter(deps, !package %in% c("XML", "RCurl"))
deps_wo |> 
  ggplot() +
  geom_point(aes(first_deps_n, downloads, shape = type)) +
  geom_text_repel(aes(first_deps_n, downloads, label = package),
                  data = filter(deps_wo, first_deps_n > 40 | downloads > 10^5)) +
  theme_minimal() +
  scale_y_log10(labels = scales::label_log()) +
  labs(title = "Packages and downloads", 
       x = "Direct dependencies", y = "Downloads", size = "Packages")
```

There is an outlier on \@ref(fig:plot1), the mlr package has more than 10k downloads and close to 120 direct dependencies, but down to less than 15 strong dependencies !

```{r plot2}
#| fig.cap = "Dependencies vs packages that depend on them. "
deps_wo |> 
  ggplot() +
  geom_point(aes(first_deps_n, first_rdeps_n, shape = type)) +
  geom_text_repel(aes(first_deps_n, first_rdeps_n, label = package),
                  data = filter(deps_wo, first_deps_n > 60 | first_rdeps_n > 50)) +
  theme_minimal() +
  scale_y_log10(labels = scales::label_log()) +
  labs(title = "Few dependencies but lots of dependents",
    x = "Direct dependencies", y = "Depend on them", size = "Packages")
```

In general though, the packages that have more dependencies have less direct dependencies.


```{r plot3}
#| fig.cap = "Direct strong dependencies vs downloads. Many pakcages have more than 20 direct imports."
library("ggplot2")
library("ggrepel")
deps_wo <- filter(deps, !package %in% c("XML", "RCurl"))
deps_wo |> 
  ggplot() +
  geom_vline(xintercept = 20, linetype = 2) +
  geom_point(aes(first_deps_strong_n, downloads, shape = repository)) +
  geom_text_repel(aes(first_deps_strong_n, downloads, label = package),
                  data = filter(deps_wo, first_deps_strong_n > 20 | downloads > 10^5)) +
  theme_minimal() +
  scale_y_log10(labels = scales::label_log()) +
  labs(title = "Packages and downloads", 
       x = "Direct strong dependencies", y = "Downloads", shape = "Repository")
```

One observable effect is that many packages do not comply with current CRAN rules of having 20 strong dependencies (as [described in R-internals](https://cran.r-project.org/doc/manuals/r-devel/R-ints.html#index-_005fR_005fCHECK_005fEXCESSIVE_005fIMPORTS_005f)). 
This suggests that these CRAN packages are old or that this limit is not checked in packages updates. 

```{r plot-maintainers}
#| fig.cap = "Looking at maintainers and the number of downloads they have."
data_maintainers <- deps_wo |> 
  filter(!is.na(maintainer)) |> 
  summarize(n = n(), downloads = sum(downloads), .by = maintainer)
data_maintainers |> 
  ggplot() +
  geom_point(aes(n, downloads)) +
  geom_text_repel(aes(n, downloads, label = maintainer),
                  data = filter(data_maintainers, n > 2 | downloads > 10^4)) +
  scale_y_log10(labels = scales::label_log()) +
  scale_x_continuous(breaks = 1:10, minor_breaks = NULL) +
  theme_minimal() +
  labs(title = "CRAN maintainers that depend on XML and RCurl",
       x = "Packages", y = "Downloads")
```

Most maintainer have few packages, some highly used packages but some have many packages relatively highly used. 

### Finding important packages

We can use a PCA to find which packages are more important.


```{r pca-all}
#| fig.cap = "PCA of all packages."
cols_pca <-  c(4:7, 15)
pca_all <- prcomp(deps_wo[, cols_pca], scale. = TRUE, center = TRUE)
summary(pca_all)
pca_data <- cbind(pca_all$x, deps_wo)
ggplot(pca_data) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_point(aes(PC1, PC2, col = repository, shape = repository)) +
  geom_text_repel(aes(PC1, PC2, label = package), 
                  data = filter(pca_data, abs(PC1) > 2 | abs(PC2) > 2)) +
  theme_minimal() +
  theme(axis.text = element_blank()) +
  labs(title = "PCA of the numeric variables")
```

We can see in the first PCA some packages that have many downloads and/or depend on many packages.
The second one are packages with many dependencies, as explained by `rotation`:

```{r pca-rotation}
pca_all$rotation[, 1:2]
```

But more important is that are packages that are named in \@ref(fig:pca-all), there is the RUnit package, markdown and rgeos that have high number of downloads and many package depend on them one way or another. 

However we can focus on packages that without RCurl or XML wouldn't work:

```{r pca-strong}
#| fig.cap = "PCA of packages with strong dependency to XML or RCurl."
pca_strong <- prcomp(deps_wo[deps_wo$strong, cols_pca], 
                     scale. = TRUE, center = TRUE)
summary(pca_strong)
pca_data_strong <- cbind(pca_strong$x, deps_wo[deps_wo$strong, ])
ggplot(pca_data_strong) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_point(aes(PC1, PC2, col = repository, shape = repository)) +
    geom_text_repel(aes(PC1, PC2, label = package), 
                  data = filter(pca_data_strong, abs(PC1) > 2 | abs(PC2) > 2)) +
  theme_minimal() +
  theme(axis.text = element_blank()) +
  labs(title = "Important packages depending on XML and RCurl", 
       subtitle = "PCA of numeric variables of strong dependencies",
       col = "Repository", shape = "Repository")
```

The main packages that depend on XML and RCurl are from Biocondcutor, followed by mlr and rlist.
rlist has as dependency XML and only uses 3 functions from it.
mlr uses 5 different functions from XML.

```{r pca-weak}
#| fig.cap = "Packages with weak dependency to XML or RCurl."
pca_weak <- prcomp(deps_wo[!deps_wo$strong, cols_pca], 
                   scale. = TRUE, center = TRUE)
summary(pca_weak)
pca_data_weak <- cbind(pca_weak$x, deps_wo[!deps_wo$strong, ])
ggplot(pca_data_weak) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_point(aes(PC1, PC2, col = type, shape = type)) +
  geom_text_repel(aes(PC1, PC2, label = package), 
                  data = filter(pca_data_weak, abs(PC1)> 2 | abs(PC2) > 2)) +
  theme_minimal() +
  theme(axis.text = element_blank()) +
  labs(title = "PCA of packages in CRAN", col = "Type", shape = "Type")
```


```{r pca-cran}
#| fig.cap = "PCA of packages on CRAN."
keep <- deps_wo$repository == "CRAN" & deps_wo$strong
pca_cran <- prcomp(deps_wo[keep, cols_pca], 
                     scale. = TRUE, center = TRUE)
summary(pca_cran)
pca_data_strong <- cbind(pca_cran$x, deps_wo[keep, ])
ggplot(pca_data_strong) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_point(aes(PC1, PC2, col = type, shape = type)) +
    geom_text_repel(aes(PC1, PC2, label = package), 
                  data = filter(pca_data_strong, abs(PC1) > 2 | abs(PC2) > 2)) +
  theme_minimal() +
  theme(axis.text = element_blank()) +
  labs(title = "Packages in CRAN", 
       col = "Type", shape = "Type")
```



```{r pca-bioc}
#| fig.cap = "PCA of packages on Bioconductor."
keep <- deps_wo$repository == "Bioconductor"  & deps_wo$strong
pca_bioc <- prcomp(deps_wo[keep, cols_pca], 
                     scale. = TRUE, center = TRUE)
summary(pca_bioc)
pca_data_strong <- cbind(pca_bioc$x, deps_wo[keep, ])
ggplot(pca_data_strong) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_point(aes(PC1, PC2, col = type, shape = type)) +
    geom_text_repel(aes(PC1, PC2, label = package), 
                  data = filter(pca_data_strong, abs(PC1) > 2 | abs(PC2) > 2)) +
  theme_minimal() +
  theme(axis.text = element_blank()) +
  labs(title = "Packages in Bioconductor", 
       subtitle = "PCA of numeric variables of strong dependencies",
       col = "Type", shape = "Type")
```

GenomeInfoDb is the package that seems more important that only uses the `RCurl::getURL` function.


## Outro

[^2]: I recently found this as opposite of introduction/intro. 

I wanted to explore a bit how these packages got into this position [^2].

```{r deps-time}
#| fig.cap = "First release of packages in relation to the maintenance by CRAN of XML and RCurl."
deps |> 
  filter(strong) |> 
  ggplot() +
  geom_vline(xintercept = as.Date("2013-06-15"), linetype = 2) +
  geom_point(aes(first_release, downloads, col = type, shape = type, 
                 size = first_deps_strong_n)) +
  geom_label(aes(first_release, downloads, label = package),
             data = filter(deps, package %in% c("XML", "RCurl")), show.legend = FALSE) +
  theme_minimal() +
  scale_y_log10(labels = scales::label_log()) +
  annotate("text", x = as.Date("2014-6-15"), y = 5*10^5, 
           label = "CRAN maintained", hjust = 0) +
  labs(x = "Release date", y = "Downloads", 
       title = "More packages added after CRAN maintenance than before",
       subtitle = "Release date and downloads",
       col = "Depends on", shape = "Depends on", size = "Direct strong dependencies") 
```

Almost the CRAN team have been maintaining these packages longer than the previous maintainer(s?).

Next, we look at the dependencies added after CRAN started maintaining them

```{r period}
summarize(deps_wo,
          before = sum(first_release <= as.Date("2013-06-15"), na.rm = TRUE), 
          later = sum(first_release > as.Date("2013-06-15"), na.rm = TRUE),
          .by = type)
```

More packages have been released after CRAN is maintaining it than before.
Maybe packages authors trusted the CRAN team for their dependencies or there was no other alternative for the functionality.
This might also be explained by the expansion of CRAN (and Bioconductor) with more packages being added each day.
However, this places further pressure in the CRAN team to maintain those packages. Removing this burden might free more time for them or to dedicate to CRAN.

A replacement for XML could be [xml2](https://cran.r-project.org/package=xml2), first released in 2015 (which uses the same system dependency libxml2).  
A replacement for RCurl could be [curl](https://cran.r-project.org/package=curl), first released at the end of 2014 (which uses the same system dependency libcurl).

Until their release there were no other replacement for these packages (if there are other packages, please let me know). 
It is not clear to me if those packages at their first release could replace XML and RCurl. 

This highlight the importance of correct replacement of packages in the community.
Recent examples are the efforts taken by the [spatial community](https://r-spatial.org/) led by Roger Bivand, Edzer Pebesma.
Where packages have been carefully designed and planned to replace older packages that are going to be retired soon.


# Recomendations

As a final recommendations I think:

-   Disentangle the XML and RCurl circular dependency.
-   Evaluate if the xml2 and curl packages provides enough functionality to replace XML and RCurl respectively. 
If not see what should be added to these packages or how to develop alternative packages to fill the gap if needed.  
Maybe a helping documentation about the alternative from XML and RCurl could be written to ease the transition and evaluate if the functionality is covered by these packages. 
- Contact package maintainers to replace the functionality they currently depend on XML and RCurl as seen in \@ref(fig:plot-maintainers) and the maintainers of packages seen in figures \@ref(fig:pca-all), \@ref(fig:pca-strong), \@ref(fig:pca-cran), and \@ref(fig:pca-bioc).
-   Set deprecation warnings on the XML and RCurl packages.
-   Archive XML and RCurl packages in CRAN.

This might take years of moving packages around but I am confident that once the word is out, package developers will avoid XML and RCurl and current maintainers that depend on them will replace them.  

**Update**:

On 2024/01/22 the [CRAN team asked for a maintainer of XML](https://stat.ethz.ch/pipermail/r-package-devel/2024q1/010359.html)

### Reproducibility

<details>

```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 80)
sessioninfo::session_info()
```

</details>
