---
title: rOpenSci submissions
author: Lluís Revilla Sancho
date: '2020-09-02'
slug: ropensci-submissions
categories:
  - r
  - rOpenSci
tags:
  - r
  - reviews
  - GitHub
authors: []
description: 'Checking rOpenSci review process from the issues. Authors, editors, topics, speed of the review are compared with last post about Bioconductor review process.'
editor_options:
  chunk_output_type: console
featured: no
image:
  caption: ''
  focal_point: ''
subtitle: 'Checking rOpenSci review process'
summary: 'Comparing rOpenSci review process to the Bioconductor review process. Most important differences are external reviewers and build on external machines as well as a longer review time.'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, warning = FALSE, message = FALSE, 
                      echo = FALSE, cache = FALSE)
```

```{r downloading, eval=FALSE,cache=FALSE}
library("socialGH")
repo <- "ropensci/software-review"
gi <- get_issues(repo)
i <- unique(gi$id)
# gt <- get_timelines(repo) there's a limit of 40000 so we cannot use it.
gt <- lapply(i, get_timelines, repository = repo)
gt2 <- do.call(rbind, gt)
library("dplyr")
gi2 <- gi %>% 
  select(-n_comments) %>% 
  mutate(actor = poster, event = "created")
column <- intersect(colnames(gt2), colnames(gi2))
g <- rbind(gi2[, colnames(gt2)], gt2) %>% 
  arrange(id, created) %>% 
  group_by(id) %>% 
  mutate(event_n = 1:n(),
         event = unlist(event, FALSE, FALSE),
         state = ifelse(event_n == 1, list("opened"), state))
pr <- g %>% 
  ungroup() %>% 
  filter(event %in% c("merged", "committed")) %>% 
  pull(id)
g2 <- filter(g, !id %in% pr)
saveRDS(g2, file = "static/20200902_github_rOpenSci_data.RDS")
```

Following [last post on Bioconductor](https://llrs.dev/2020/07/bioconductor-submissions-reviews/) I wanted to analyze another venue where code reviews are made: [rOpenSci](https://rOpenSci.org).
There are some [other analysis](https://ropensci.org/blog/2018/04/26/a-satrday-ct-series/) of the the reviews made by rOpenSci themselves:

> The [first post](https://ropensci.org/blog/2018/04/26/rectangling-onboarding/) in the series will explain how I rectangled onboarding. The [second post](https://ropensci.org/blog/2018/05/03/onboarding-is-work/) will give some clues as to how to quantify the work represented by rOpenSci onboarding. The [third and last post](https://ropensci.org/blog/2018/05/10/onboarding-social-weather/) will use tidy text analysis of onboarding threads to characterize the social weather of onboarding.


rOpenSci review process has some of the differences:  

One can ask before hand if the package submitted fits the scope of rOpenSci. 
In addition the review is done by two volunteers usually not affiliated with rOpenSci but with an editor that has the final decision
Also there is an encouragement on a dialog and iteration to improve the package.
Lastly, the build process is perform by third parties (Github actions, travis, appveyor ...) and the results are not reported on the issue.

Despite all these differences following the same methods I downloaded the data on 2020/09/02 to analyze similarly the submissions.

```{r read}
init <- readRDS(here::here("static", "20200902_github_rOpenSci_data.RDS"))
library("tidyverse")
theme_set(theme_minimal())
get_element <- function(x, name) {
  if (!is.null(names(x))) {
    getElement(x, name)
  } else {
    NA_character_
  }
} 
init <- init %>% 
  filter(id != 26) %>%
  mutate(reviewer = vapply(assignee, get_element, name = "user", character(1)),
         actor = vapply(actor, get_element, name = "user", character(1L)))
```



```{r by-issue}
reviewers <- function(assigners, unassigners) {
  ta <- table(assigners)
  tu <- table(unassigners)
  y <- 0
  n <- sum(ta) - sum(tu)
  reviewers <- vector("character", n)
  for(reviwer in names(ta)) {
    x <- ta[reviwer]-tu[reviwer]
    if (x >= 1 | is.na(x)){
      y <- y + 1
      reviewers[y] <- reviwer
    }
  }
  reviewers
}

by_issue <- init %>% 
  group_by(id) %>% 
  summarize(time_window = difftime(max(created), min(created), units = "days"),
            events = n(), 
            diff_users = n_distinct(actor),
            diff_events = n_distinct(event),
            assignments = sum(event %in% "assigned"),
            reassigned = any(event %in% "unassigned"),
            assigners = list(reviewer[event %in% "assigned"]),
            unassigners = list(reviewer[event %in% "unassigned"]),
            editors = list(reviewers(unlist(assigners, FALSE, FALSE), 
                                    unlist(unassigners, FALSE, FALSE))),
            editor_comments = sum(event == "commented" & 
                                      actor %in% unlist(editors), na.rm = TRUE),
            closers = list(actor[event == "closed"]),
            openers = list(actor[event == "reopened"]),
            closed = sum(event == "closed") >= sum(event == "reopened") & any(event == "closed"),
            closer = list(setdiff(unlist(closers, FALSE, FALSE), 
                                    unlist(openers, FALSE, FALSE))),
            labels_added = list(label[event == "labeled"]),
            labels_removed = list(label[event == "unlabeled"]),
            labels_final = list(setdiff(unlist(labels_added, FALSE, FALSE), 
                                    unlist(labels_removed, FALSE, FALSE))),
            check_labels = all(unlist(labels_final, FALSE, FALSE) %in%
                                 label[event == "created"]),
            submitter = actor[event == "created"]
            
) %>% 
  mutate(n_reviewers = lengths(editors),
         n_closers = lengths(closer))

by_issue1 <- init %>% 
  count(event) %>% 
  filter(event != "created") %>% 
  pivot_wider(values_from = n, names_from = event, values_fill = 0) %>% 
  nest_by(.key = "event")
by_issue2 <- init %>% 
  count(actor) %>% 
  pivot_wider(values_from = n, names_from = actor, values_fill = 0) %>% 
  nest_by(.key = "actor")

by_issue <- by_issue %>% 
  inner_join(by_issue1) %>% 
  inner_join(by_issue2)
```


```{r by-user}
by_user <- init %>% 
  group_by(actor) %>% 
  summarize(
    actions = n(),
    issues_participated = n_distinct(id),
    issues = list(unique(id)),
    events_participated = n_distinct(event),
  ) %>% 
  mutate(is_editor = actor %in% unlist(by_issue$editors, FALSE, FALSE))

by_user1 <- init %>% 
  group_by(actor) %>% 
  count(event) %>% 
  pivot_wider(values_from = n, names_from = event, values_fill = 0) %>% 
  nest_by(.key = "event")
by_user2 <- init %>% 
  group_by(actor) %>% 
  count(id) %>% 
  pivot_wider(values_from = n, names_from = id, values_fill = 0) %>% 
  nest_by(.key = "ids")

by_user <- by_user %>% 
  inner_join(by_user1) %>% 
  inner_join(by_user2)
```

```{r first-plot}
init %>% 
  ggplot(aes(created, id, col = fct_lump(event, 7))) +
  geom_point() +
  labs(x = element_blank(), y = "Issue", col = "Event",
       title = "Issues opened on rOpenSci") +
  theme_minimal()
```

We can see that there are fewer issues opened on rOpenSci but also the timeline is shorter. 
One difference is that there seems to be more events after the initial activity on the issues.

If we look at the editors involved on the issues we can see who has commented more and in more issues

```{r by-users}
library("ggrepel")
editors <- by_user %>% 
  filter(is_editor) %>% 
  distinct(actor) %>% 
  pull(actor)
by_user %>% 
  filter(is_editor) %>%
  unnest(event) %>% 
  filter(commented != 0) %>% 
  ggplot() + 
  geom_point(aes(actions, issues_participated)) +
  geom_text_repel(aes(actions, issues_participated, label = actor)) +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") +
  labs(size = "Users", col = "Different events", x = "Comments", y = "Issues",
       title = "Editors involvement") +
  theme_minimal()
```

It seems like there are 10 comments per each issue an editor is involved with.
Of course, this will change for people submitting software:

```{r by-user2}
by_user %>% 
  filter(!is_editor & actor != "bioc-issue-bot" & !is.na(actor)) %>%
  unnest(event) %>% 
  filter(commented != 0) %>% 
  ggplot() + 
  geom_count(aes(actions, issues_participated)) +
  scale_color_viridis_d() + 
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") +
  geom_text_repel(aes(actions, issues_participated, label = actor), 
                data = . %>% filter(issues_participated > 10 | actions > 100)) +
  labs(size = "Users", col = "Different events", x = "Comments", y = "Issues",
       title = "Users involvement") +
  theme_minimal()
```

Here I kept the same threshold to show the names of users. 
There are more names, indicating more people commenting and involved on the issues.
I'm surprised to appear on this graph because I haven't been much involved with rOpenSci.
Stefanie Butland as the community manager of rOpenSci comments more and is involved in more issues than regular users. 

Now that we know who is involved comenting, what about the other events?

```{r events-view}
init %>% 
  group_by(event) %>% 
  summarize(n = n(),
            diff_issues = n_distinct(id),
            diff_author = n_distinct(actor)) %>% 
  ggplot() +
  geom_abline(slope = 1, col = "grey") +
  geom_text_repel(aes(diff_issues, diff_author, label = event, col = n, size = n)) +
  scale_y_continuous(trans = "log10") +
  scale_x_continuous(trans = "log10", limits = c(1, NA)) +
  scale_color_binned(trans = "log10", guide = "bins") +
  scale_size(trans = "log10") +
  guides(size = FALSE) +
  labs(title = "Events", x = "Number of different issues", 
       y = "Number of different users", label = "Events", size = "Events",
       col = "Events") +
  theme_minimal()
```

First thing, more people are mentioned between them than on Bioconductor. 
Also there seems to be a lot more of cross-references than on Bioconductor.

```{r second-plot}
init %>% 
  count(event) %>% 
  filter(!event == "created") %>% 
  ggplot() +
  geom_tile(aes(id, fct_reorder(event, n, sum) , col = n)) +
  scale_color_viridis_c(trans = "log10") +
  labs(x = "Issue", y = element_blank(), col = "Times", 
       title = "Events per issue") +
  theme_minimal()
```

The issues do have a highly social component with lots of mentions, subscriptions and comments.
It is generally rare to unsubscribe from an issue or to have the issue added to a project.

```{r events-time}
init %>% 
  count() %>% 
  ggplot() +
  geom_histogram(aes(n), binwidth = 5) +
  labs(x = "Events", y = "Issues", title = "Events per issue") +
  theme_minimal()
```

Here we don't see a clear pattern on the events, but most issues have few events (median events of `r round(median(table(init$id)))`).

```{r events-time2}
init %>% 
  summarize(time = difftime(created[created == max(created)], created[event == "created"],
                            units = "days"),
            n = n(),
            slope = n/as.numeric(time)) %>% 
  ggplot(aes(time, n)) +
  geom_point(aes(col = slope)) +
  scale_color_continuous(trans = "log10") +
    labs(y = "Events", x = "Time (days)", 
       col = "Events per days", 
       title = "Number of events and time")  +
  theme(axis.text.x = element_text())
```

It is fairly similar to Bioconductor. 
We can see that issues have fewer events than on Bioconductor. 
This is because there isn't any bot replying with each update to the repository neither a report of the build with every version increment. 

```{r events-time3}
init %>% 
  summarize(time = difftime(created[created == max(created)], created[event == "created"],
                            units = "days"),
            n = n(),
            slope = n/as.numeric(time)) %>% 
  ungroup() %>% 
  filter(time <= median(time)) %>% 
  ggplot(aes(time, n)) +
  geom_point(aes(col = slope)) +
  scale_color_continuous(trans = "log10") +
  labs(y = "Events", 
       x = "Time (days)", 
       col = "Events per days", 
       title = "Number of events and time",
       subtitle = "A zoom to the fastest half") +
  scale_x_continuous(breaks = 1:7*14, labels = function(x) {paste(x/7, "weeks")}) +
  theme_minimal()
```

Most issues have all the events on the first weeks.

# Editor

As mentioned the editor is assigned to find the reviewers and have the final decision on the package.

```{r assignments}
by_issue %>% 
  ggplot() +
  geom_count(aes(as.factor(n_reviewers), assignments, col = reassigned, 
                 shape = reassigned)) +
  labs(x = "Final editor", y = "Assigned", title = "Editors", 
       col = "Reassigned?", shape = "Reassigned?", size = "Submissions") +
  scale_color_brewer(labels = c("No", "Yes"), type = "qual") +
  scale_shape(labels = c("No", "Yes")) +
  scale_size(trans = "log10")
```

There seem to be few assigned users, and in some cases there are like two users assigned.
This might be because the reviewers are also assigned that issue or something else.

```{r submission-acceptance}
revi <- by_issue %>% 
  filter(lengths(editors) != 0)

reviwer_didnt_close <- revi %>% 
  filter(!is.na(closer),
         closer %in% unlist(editors, FALSE, FALSE)) %>% 
  pull(id)

author_closed <- revi %>% 
  filter(!is.na(closer), submitter %in% closer) %>% 
  pull(id)

revi_sum <- revi %>% 
  filter(!id %in% author_closed,
         n_reviewers == 1) %>% 
  mutate(editor = unlist(editors, FALSE, FALSE),
         editor_commented = ifelse(editor_comments != 0, "commented", "not commented")) %>% 
  group_by(editor, editor_commented) %>% 
  count() %>% 
  group_by(editor) %>% 
  mutate(perc = n/sum(n)) %>% 
  arrange(editor, editor_commented)

revi %>% 
  ggplot() +
  geom_histogram(aes(editor_comments, fill = n_reviewers), binwidth = 1) +
  labs(x = element_blank(), y = "Editor comments", 
       title = "Comments from editors") + 
  theme_minimal()
```

Editors are fairly more involved on the issues commenting around 11 times.

```{r events-days}
trelative <- function(x) {
  created <- x$created
  event <- x$event
  start <- created[event == "created"]
  k <- event == "closed"
  if (any(k)){
    closing <- created[which.max(k)]
  } else {
    closing <- max(created)
  }
  
  o <- difftime(created[!is.na(created)], start, units = "days")
  as.numeric(o)
}
full <- init %>% 
  nest_by() %>% 
  summarize(time_relative = trelative(data), created = data$created) %>% 
  inner_join(init) %>% 
  unique() %>% 
  mutate(presubmission = any(grepl("[Pp]re-?[Ss]ubmission", title))) %>% 
  mutate(presubmission = ifelse(presubmission, "Presubmission", "Submission"))
full %>% 
  arrange(id, created, event) %>% 
  mutate(event_n = 1:n()) %>% 
  filter(event_n <= event_n[which.max(event == "closed")]) %>% 
  ggplot() +
  geom_line(aes(time_relative, event_n, col = id, group = id)) +
  facet_wrap(~presubmission, scales = "free") +
  labs(title = "Events along time till closed",
       x = "Days", y = element_blank(), col = "Issue") + 
  theme_minimal()
```

We can see that one presubmission issue took longer to decide than most of the submissions!
In general submissions show the same behavior as in Bioconductor:
Most of them have lots of events on a relatively short period of time. 
Some submissions take longer to close and remain without events for long periods of time.

```{r events-user-distribution}
p1 <- ggplot(by_issue) +
  geom_bar(aes(as.factor(diff_events))) +
  labs(y = "Issues", x = element_blank(), 
       title = "Different events in the issue")
p2 <- ggplot(by_issue) +
  geom_bar(aes(as.factor(diff_users))) +
  labs(y = element_blank(), x = element_blank(), 
       title = "Different users involved in the issue")
library("patchwork")
p1 + p2
```

Here are some differences on the number of different events per issue and different users involved.
There are more users and events involved with issues on rOpenSci than on Bioconductor.
This is partially expected as the review are done by more people but I expected a similar pattern of events per issue.

```{r actor-event-types}
by_issue %>% 
  mutate(diff_actors = factor(diff_users, levels = seq_len(max(diff_users)))) %>% 
  ggplot() +
  geom_count(aes(diff_actors, as.factor(diff_events))) +
  labs(x = "Users", y = "Events", title = "Users involved and different events",
       size = "Issues") + 
  scale_radius() +
  scale_x_discrete(drop = FALSE) +
  theme_minimal()
```

We can see that users and events are more distributed. 


```{r actors-events}
by_issue %>% 
  mutate(diff_actors = factor(diff_users, levels = seq_len(max(diff_users)))) %>% 
  group_by(diff_actors) %>% 
  count(events) %>% 
  ggplot() +
  geom_jitter(aes(as.factor(diff_actors), events, size = n), 
             height = 0) +
  labs(x = "Users", y = "Events", size = "Issues",
       title = "Users involved on the issues and events") +
  scale_x_discrete(drop = FALSE) + 
  scale_size(breaks = c(seq(0, 300, by = 50)))
```

As in Bioconductor the more people involved more events per issue.

# Who does each action ?

We can look now at who performs what, we know there are `r nrow(by_user)` participants:

```{r who}
top_events <- 35
by_user %>%
  arrange(-actions) %>% 
  top_n(top_events, actions) %>% 
  select(actor, event) %>% 
  unnest(event) %>% 
  pivot_longer(names_to = "event", cols = created:unlabeled, values_to = "n") %>% 
  filter(n != 0) %>% 
  ungroup() %>% 
  ggplot() + 
  geom_tile(aes(fct_reorder2(event, -n, actor), 
                fct_reorder(actor, n, .fun = sum), 
                fill = n)) +
  scale_fill_viridis_c(trans = "log10", expand = expansion()) +
  labs(title = "Events by users", y = element_blank(), x = element_blank(),
       fill = "Events") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
  # coord_flip()
```

The top 35 people involved is dominated by editors.
Surprisingly editors do also cross-reference other issues.

If we look at how many comments there are on each issue per author and editor we can get a sense of how much work it takes for editors:

```{r comments-plot}
comments <- init %>%
  mutate(n = sum(event %in% "assigned"),
         assigners = list(reviewer[event == "assigned"]),
         unassigners = list(reviewer[event == "unassigned"]),
         reasigned  = any(event %in% "unassigned"),
         reviewers = list(reviewers(unlist(assigners, FALSE, FALSE), 
                                    unlist(unassigners, FALSE, FALSE))),
         reviewers_n = lengths(assigners)-lengths(unassigners),
         creator = unique(actor[event == "created"])) %>% 
  filter(actor != "bioc-issue-bot", event == "commented", reviewers_n >= 1) %>% 
  group_by(id) %>% 
  summarise(speaking = n_distinct(actor), 
            comments = n(), 
            reviewer = sum(actor %in% unlist(reviewers[1])),
            author = sum(actor == creator),
            mtmorgan = sum(actor == "mtmorgan"),
            other = comments - reviewer - author)
comments %>% 
  ggplot() +
  geom_abline(slope = 1, col = "grey") +
  geom_count(aes(author, reviewer)) +
  labs(title = "Comments", x = "Authors", y = "Editor(s)", size = "Issues") +
  theme_minimal() 
```

Here we can see that there are more comments from authors and editors.

```{r closing}
closed_c <- full %>% 
  mutate(comment_n = cumsum(event == "commented")) %>% 
  filter(event == "closed") %>% 
  filter(event_n == max(event_n)) %>% 
  select(id, time_relative, comment_n, event_n)

closed_c <- mutate(closed_c, Closed = TRUE)
aftr <- full %>% 
  filter(event %in% c("commented", "created") & actor != "bioc-issue-bot") %>% 
  mutate(comment_n = cumsum(event == "commented" & actor != "bioc-issue-bot"),
         Closed = FALSE) %>% 
  select(id, time_relative, comment_n, event_n, Closed) %>% 
  rbind(closed_c) %>% 
  group_by(id) %>% 
  mutate(Closed = event_n > ifelse(any(Closed), event_n[Closed], Inf))
aftr %>% 
  group_by(id) %>% 
  count(Closed) %>% 
  filter(n_distinct(Closed) >= 2) %>% 
  pivot_wider(names_from = Closed, values_from = n) %>% 
  ggplot() + 
  geom_count(aes(`TRUE`, `FALSE`)) +
  labs(title = "Comments before and after being closed",
       size = "Issues", x = "Comments after being closed",
       y = "Comments before being closed")
```

Some issues have fairly large number of comments after being closed.
Perhaps discussing other alternatives or comparing to other existing software.

# Mentions

There are many more mentions than on Bioconductor.
So what do these people do when added to the conversation ?

```{r mentions}
mentioned <- init %>% 
  ungroup() %>% 
  filter(event == "mentioned") %>% 
  distinct(actor) %>% 
  pull()
non_editors <- mentioned[!mentioned %in% editors]
creators <- init %>% 
  filter(event == "created")
init %>% 
  filter(actor %in% non_editors,
         !event %in% c("mentioned", "subscribed")) %>% 
  anti_join(creators) %>% 
  count(event) %>% 
  ggplot() +
  geom_tile(aes(id, fct_reorder(event, n, sum), col = n)) +
  labs(x = "Issue", y = "Event", col = "Times", 
       title = "What do people mentioned?")
```

People mentioned bring their experience by commenting, and cross-referencing some other issues. 

# Labels

rOpenSci uses labels much more than Bioconductor, with a total of 65 different labels.
It also uses them to mark on which step of the review process each issue is. 
We can see that it got expanded with time, from initially just three labels to the current 6-7 labels:

```{r labels-general}
labels <- full %>% 
  ungroup() %>% 
  filter(event == "labeled") %>% 
  mutate(label = unlist(label),
         label = case_when(
           label == "reviewer-requested" ~ "2/seeking-reviewer(s)",
           label == "seeking-reviewers" ~ "2/seeking-reviewer(s)",
           label == "2/seeking-reviewers" ~ "2/seeking-reviewer(s)",
           label == "3/reviewers-assigned" ~ "3/reviewer(s)-assigned",
           label == "4/review-in-awaiting-changes" ~ "4/review(s)-in-awaiting-changes",
           label == "review-in-awaiting-changes" ~ "4/review(s)-in-awaiting-changes",
           label == "changes-in-awaiting-response" ~ "4/review(s)-in-awaiting-changes",
           label == "5/awaiting-reviewer-response" ~ "5/awaiting-reviewer(s)-response",
           label == "approved" ~ "6/approved",
           label == "topic:linquistics" ~ "topic:linguistics",
           TRUE ~ label
         ))
```


```{r labels_plot_overview}
ord_label <- c("0/presubmission",
               "1/editor-checks",  
               "2/seeking-reviewer(s)", "3/reviewer(s)-assigned", 
               "4/review(s)-in-awaiting-changes",
               "5/awaiting-reviewer(s)-response", 
               "6/approved", 
               "TIMEOUT", "ERROR", 
               "WARNINGS", "OK")
labels %>% 
  group_by(id) %>% 
  count(label, sort = TRUE) %>% 
  filter(label %in% ord_label) %>% 
  ggplot() +
  geom_tile(aes(id, label, fill = n)) +
  labs(x = "Issue", y = element_blank(), title = "Labels related to the review process",
       fill = "Times")
```

But labels are also used to indicate to which is the topic of each package:

```{r labels_topic}
labels %>% 
  group_by(id) %>% 
  count(label, sort = TRUE) %>% 
  filter(grepl(pattern = "^topic", label)) %>% 
  mutate(label = gsub(pattern = "topic:", replacement = "", x = label)) %>% 
  ggplot() +
  geom_tile(aes(id, fct_reorder(label, n, .fun = sum), fill = n)) +
  labs(x = "Issue", y = element_blank(), title = "Topics", fill = "Times")
```

Surprisingly the focus seems to be on data-access followed by geospatial and data-extraction:

```{r topics-freq}
labels %>% 
  group_by(id) %>% 
  count(label, sort = TRUE) %>% 
  filter(grepl(pattern = "^topic", label)) %>% 
  mutate(Topic = gsub(pattern = "topic:", replacement = "", x = label)) %>% 
  ungroup() %>% 
  count(Topic, sort = TRUE) %>% 
  slice_head(n = 10) %>% 
  knitr::kable()
```

Labels are also used to classify topics (and initially if it had an editor assigned)

```{r labels-other}
labels %>% 
  group_by(id) %>% 
  count(label, sort = TRUE) %>% 
  filter(!grepl(pattern = "^topic", label) & !label %in% ord_label) %>% 
  mutate(label = gsub(pattern = "topic:", replacement = "", x = label)) %>% 
  ggplot() +
  geom_tile(aes(id, fct_reorder2(label, n, n), fill = n)) +
  labs(x = "Issue", y = element_blank(), title = "Other topics", fill = "Times")
```

Other labels assigned, seem to be for packages or some other related to the process. 
Surprisingly there have been few submissions to MEE and JOSS according to the "pub:" labels. 

Looking at the labels for each step can compute the median time required to get them:

```{r labels-step}
labels %>% 
  filter(label %in% ord_label) %>% 
  group_by(label) %>% 
  summarize(days = median(time_relative)) %>% 
  filter(label != "0/presubmission") %>% 
  mutate(d = days -lag(days),
         time = round(if_else(is.na(d), days, d), 1),
         `Total days` = round(cumsum(time), 1)) %>% 
  select(label, time, `Total days`) %>% 
  knitr::kable(col.names = c("Step", "Median days", "Total days"))
```

We see that the slowest step is awaiting reviewers response, right after awaiting changes from the authors.
According to this the review process is long and it takes around 3 months to get a package approved.
However, if we look for each issue how much time does it take to get the next label we see another picture:

```{r labels_ind}
lab_wide <- labels %>% 
  ungroup() %>% 
  filter(label %in% ord_label[2:7]) %>% 
  pivot_wider(id_cols = id, values_from = time_relative, names_from = label,
              values_fn = last) %>% 
  select(id, ord_label[2:7])
colnames(lab_wide)[2:7] <- as.character(c(1:6))
res <- lab_wide %>% 
  summarize(
    s1 = median(`1`, na.rm = TRUE),
    s2 = median(`2` - `1`, na.rm = TRUE),
    s3 = median(`3` - `2`, na.rm = TRUE),
    s4 = median(`4` - `3`, na.rm = TRUE),
    s5 = median(`5` - `4`, na.rm = TRUE),
    s6 = median(`6` - `5`, na.rm = TRUE))
colnames(res) <- ord_label[2:7]
res %>% 
  pivot_longer(cols = ord_label[2:7]) %>% 
  mutate(`Median days` = round(value, 1),
         `Total days` = round(cumsum(value), 1)) %>% 
  select(name, `Median days`, `Total days`) %>% 
  knitr::kable()
```

Now the longest step is changing the packages with the feedback from the reviewers, followed by awaiting the reviewers comments but that is almost half of the time it was. 
The other times do not change much.
The total time is still around 2 months, which is the double of what it takes to get a  package accepted in Bioconductor.

# Conclusions

Following the steps/labels the review process is similar enough to be able to compare the review process of [Bioconductor](https://bioconductor.org) and [rOpenSci](https://ropensci.org).
Not having information about the build and check status of packages makes harder to compare some steps and the stage of the package upon submission. 
On some early submissions it was editor's duties to review the packages like in Bioconductor.
It was later abandoned in favor of two external reviewers.
The reviews on the rOpenSci are handled by more people which makes submission to take longer (probably also because usually there are two reviewers) and 
because there might be a clarification of changes and a dialog after the first review.

In general it takes longer to get the packages approved than on Bioconductor. 


### Reproducibility

<details>
```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info()
```
</details>
