---
title: "Exploring CRAN's files: part 3"
subtitle: 'References: links between pages, packages and base R'
author: Lluís Revilla Sancho
date: '2024-09-23'
slug: cran-files-3
categories:
  - CRAN
  - r
tags:
  - r
  - packages
  - cran-files
authors:
  - admin
description: ''
draft: no
editor_options:
  chunk_output_type: console
featured: yes
image:
  caption: ''
  focal_point: ''
summary: 'Exploring references and help pages of base R packages and CRAN packages.'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.retina = 2, cache = TRUE, include = TRUE, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, eval = TRUE)
Sys.setlocale("LC_ALL", "C")
Sys.setenv(R_CRAN_WEB = "https://cloud.r-project.org")
```


In the blog post of CRAN files, I [initially explored the database](https://llrs.dev/post/2022/07/23/cran-files-1/) (mostly around `CRAN_package_db()`). On the [second post](https://llrs.dev/post/2022/07/28/cran-files-2/) I moved to analyze the archive around `CRAN_archive_db()`.

Recently at useR!2024, Kurt Hornik offered to export data from CRAN. After an email [several new functions were exported](https://developer.r-project.org/blosxom.cgi/R-devel/NEWS/2024/08/22#n2024-08-22) and introduced to base R.

In that presentation and in a previous exchange with Kurt, he explained his project about providing the HTML manual pages of all CRAN packages.
One of the problems with this projects is providing links to the right pages of the packages (others are making it accessible for all users).

In R-devel now there are test that ensure links to other pages are in the right format [^1].
In this post we we'll explore links between documentation pages, what manuals say about them and what is the state of base R and the CRAN packages.

[^1]: By setting `_R_CHECK_XREFS_NOTE_MISSING_PACKAGE_ANCHORS_=true` as environmental variable.

# Introduction

Help pages are defined in R documentation files, commonly with the `.Rd` extension.
Each documentation file can have multiple topics.
A topic is defined in R documentation format with `\alias{topic}`, so once can use interchangeably alias and topics.
Links, or cross-references (xref for short) [should be to an alias](https://cran.r-project.org/doc/manuals/r-devel/R-exts.html#Cross_002dreferences), not to files (allowing to move topics between help pages)[^4].

[^4]: But using links to files is still fine.

One can define links with `\link{alias}` or `\link[=alias]{name}` if the link is to a different place than the name [^2]. 
In the second form the content within the square bracket is known as anchor.

There are two other forms accepted: `\link[pkg]{alias}` and `\link[pkg:alias]{name}` which do not use the `=` inside the anchor [^3]. 
These last two forms are only used in the HTML format. 
Packages referred to on these square brackets should be declared in the DESCRIPTION file, in the ‘Depends’, ‘Imports’, ‘Suggests’ or ‘Enhances’ fields.

One important thing is that links are case sensitive: `?Complex` leads to a different help page than `?complex`. 

[^2]: There is also the `\linkS4class{abc}` that expands to `\link[=abc-class]{abc}` for S4 classes.

[^3]: So there are three names for (almost) the same thing: topic, alias and anchor. Anchors do not need to start with `=` but should be resolvable.

With this in mind we can check to which packages have more manual pages, have more cross-references, or have an overview help page (those with *pkgname*-package.Rd).
We will also check which documentation pages have more links, are more linked to, or needs its links fixed.

# Base R

Before looking to CRAN packages we can start by looking to R base packages.

Previously there wasn't a way to know the links and aliases in base R. Since recently (~2024/08/20) we can download them (on R-devel till it is released on R 4.5) with `tools::base_aliases_db()` and `tools:::base_rdxrefs_db()`.

```{r base}
aliases_base <- tools::base_aliases_db()
xrefs_base <- tools:::base_rdxrefs_db()
```

We retrieve both so that we can compare and explore better the packages.

## Aliases

```{r}
#| label: alias2df
alias2df <- function(x){
  l <- lapply(x, function(x) {
    data.frame(Source = rep(names(x), lengths(x)),
               Target = unlist(x, FALSE, FALSE))
  })
  aliasesDF <- do.call(rbind, l)
  aliasesDF$Package <- rep(names(l), vapply(l, NROW, numeric(1L)))
  rownames(aliasesDF) <- NULL
  aliasesDF[, c("Source", "Target", "Package"), drop = FALSE]
}

aliasesBase <- alias2df(aliases_base)
library("dplyr")
```

Looking at the aliases present on base R we find, there are `r dplyr::n_distinct(aliasesBase$Source)` help pages with `r dplyr::n_distinct(aliasesBase$Target)` topics/aliases.

```{r}
#| label: base-dup-files
dup_files <- aliasesBase |> 
  select(-Target) |> 
  distinct() |> 
  summarise(.by = Source, 
            dup_pages = n(),
            packages = paste(Package, collapse = ", ")) |> 
  arrange(-dup_pages) |> 
  filter(dup_pages > 1L) |> 
  head()
```

There are some help files with the same name in base R that can be accessed via their targets (via `?Target`). 
Those files are `r dup_files$Source`.  
To distinguish between those files one need to check the title(s) and content of the pages.

Similarly there are multiple help pages of different packages with the same topics:

```{r}
#| label: base-dup-targets
dtBase <- aliasesBase |> 
  summarise(.by = Target, n = n(), Package = paste0(Package, collapse = ", ")) |> 
  filter(n > 1) |> 
  arrange(Target, Package) |> 
  select(Alias = Target, Packages = Package)
dtBase
```

Most of them are in base and other packages, from the methods package to tools, grDevices, graphics, and utils.

To access them directly one needs to use `help()`: `help("clipboard", package = "utils")` otherwise one will need to choose them from the menu it will pop up from using `?clipboard`.
This works to access any other help page too.

If we explore

```{r}
#| label: base-dup-pages
uniq_target_pages <- aliasesBase |> 
  filter(.by = Source, !Target %in% dup_files$Alias) |> 
  filter(.by = c(Package, Source), n()  == 1) |> 
  select(Package, Target, Source) |> 
  group_by(Target, Package) |> 
  arrange(Target, Package, Source) |> 
  ungroup() |> 
  select(Target, Package, Source)
```

Looking at the help pages are `r nrow(uniq_target_pages)` that only have one alias. For example, `.BasicFunsList`, `.Platform` or `.Last.value`. 
Some of these are datasets or very specific help pages


```{r}
#| label: base-package.rd
#| eval: interactive()
aliasesBase |> 
  filter(grepl("-package.Rd", Source)) |> 
  filter(.by = Package, sum(Target == Package, 
                            grepl(unique(paste0(Package, "-package")), Target)) != 2) |> 
  select(Package, Source, Alias = Target)
```

Last, we can check the topics for the packages' help page.
While usually packages have at least two alias, `pkg` and one `pkgname-package`, there are two packages that do not have this, grid and methods packages.  
These packages lack the form of `pkgname` to find the package help file.

## Cross references

Having explored the topics on the help pages, it is time to move to the cross references on R.

We can check where these help pages link to, starting by knowing how many are directly to the alias or those that use an anchor:

```{r}
#| label: xrefsBase
#| fig-cap: 
xrefs2df <- function(x) {
  rdxrefsDF <- do.call(rbind, x)
  rdxrefsDF <- as.data.frame(rdxrefsDF)
  rdxrefsDF$Package <- rep(names(x), vapply(x, NROW, numeric(1L)))  
  rownames(rdxrefsDF) <- NULL
  rdxrefsDF[, c("Source", "Target", "Anchor", "Package"), drop = FALSE]
  rdxrefsDF
}
xrefBase <- xrefs2df(xrefs_base)
```


```{r}
#| label: links-base
#| fig-cap: "Packages with links using alias per help page. Each point is a help pages with different number of links. Some packages directly link a lot while others are more mixed."
library("ggplot2")

xref_summary <- xrefBase |> 
  group_by(Package, Source) |> 
  count(Change_name = nzchar(Anchor)) |> 
  ungroup() |> 
  arrange(Package, Source, Change_name) |> 
  mutate(.by = c(Package), help_pages = n_distinct(Source)) |> 
  mutate(.by = c(Package, Source), ratio_source = sum(n[Change_name])/sum(n)) |> 
  summarise(.by = c(Package, Change_name, n, help_pages, ratio_source), size = n()) |> 
  mutate(.by = c(Package), ratio_pkg = sum(n[Change_name])/sum(n)) 

library("forcats")
xref_summary |> 
  ggplot() +
  geom_jitter(aes(ratio_source, fct_reorder(Package, help_pages, .fun = max), 
                  size = n), width = 0) +
  scale_size(limits = c(1, NA), breaks = c(1, seq(from = 10, to = 100, by = 10))) +
  scale_x_continuous(labels = scales::label_percent(), expand = expansion(add = 0.025, mult = 0)) +
  labs(x = "Links using alias", y = element_blank(), 
       size = "Links",
       title = "Links with alias per help page",
       subtitle = "Packages sorted by number of help pages") +
  theme_minimal() +
  theme(plot.title.position = "plot")
```


We can further explore to where they link if we find out the package and topic, instead of just the topic used.

```{r}
#| label: base-split-dest
xrefBase <- cbind(xrefBase, 
                   strcapture("([[:alnum:].]*{2,})?[:=]?(.*)", 
                              x = xrefBase$Anchor, 
                              proto = data.frame(to_pkg = character(), 
                                                 to_target = character())))
```

```{r}
#| label: base-fill-xref
fill_xref <- function(z, alias, duplicate_alias) {

  # Anchors with packages
  missing_target <- nzchar(z$Anchor) & !startsWith(z$Anchor, "=") & !grepl(":", z$Anchor, fixed = TRUE)
  z$to_target[missing_target] <- z$Target[missing_target]
  
  # Anchors with = to base packages with no duplicates
  anchors_nodup <- startsWith(z$Anchor, "=") & !z$Anchor %in% duplicate_alias
  match_target <- match(z$to_target[anchors_nodup], alias$Target)
  z$to_pkg[anchors_nodup] <- alias$Package[match_target]
  
  # Anchors with = to packages with duplicates
  anchors_dup <- startsWith(z$Anchor, "=") & z$Anchor %in% duplicate_alias
  z$to_pkg[anchors_dup] <- z$Package[anchors_dup]
  
  # No anchors
  no_anchor <- !nzchar(z$Anchor)
  match_target <- match(z$Target[no_anchor], alias$Target)
  z$to_pkg[no_anchor] <- alias$Package[match_target]
  z$to_target[no_anchor] <- alias$Target[match_target]
  z
}

xrefBase2 <- fill_xref(xrefBase, aliasesBase, dtBase$Alias)

# We can then add the aliases to know to which help page do they link.
xrefBase3 <- merge(xrefBase2, aliasesBase,
            by.x = c("to_pkg", "to_target"), 
            by.y = c("Package", "Target"), 
            all.x = TRUE, sort = FALSE) |> 
  select(Rd_origin = Source.x, Anchor, Target, Package, to_pkg, to_target, Rd_destiny = Source.y)
# As expected, xref to other packages not in base R are not included. 
```


```{r}
#| label: self-xref
self_xref <- xrefBase3 |> 
  filter(Rd_origin == Rd_destiny, Package == to_pkg) |> 
  select(Rd_origin, Target, Package) |> 
  distinct(Rd_origin, Package)

```

We can check for pages that link to themselves, there are `r nrow(self_xref)`:

```{r}
#| label: self-references
#| tab.cap: "**Help pages on base with links to themselves**"
sx <- self_xref[, c("Package", "Rd_origin")]
colnames(sx)[2] <- c("Help page")
sort_by(sx, ~Package)
```

The cross-reference system doesn't provide a way to link to a specific section (yet?). 
When users follow those links they are redirected to the top of the same help page,
while probably they were expecting to go to another help page.
At least this is something I have felt in the past. 
In my opinion these links could be deleted and rewritten to explain which section should the user read. 

Finally there are help files with references to other help pages: 

```{r}
#| label: plot-base-package-links
#| fig-cap: "Links between R help pages are predominantly to their own package. 
#|  Base is more connected than any other package. Some packages recieve links from very few packages."
# Explore the information
cross_ref_base <- summarise(xrefBase3, 
          .by = c(Package, to_pkg), 
          n = n(), 
          ndocs = n_distinct(Rd_origin), 
          nlinks = n_distinct(to_pkg, to_target), 
          ndocs_d = n_distinct(to_target))

cross_ref_base |> 
  filter(!is.na(to_pkg),
         to_pkg %in% names(aliases_base)) |> 
  mutate(Package = fct_reorder2(Package, n, ndocs, .fun = sum),
         Package = factor(Package, levels = rev(levels(Package)))) |> 
  ggplot() +
  geom_point(aes(fct_relevel(to_pkg, rev(levels(Package))),
                 Package, 
                 col = n, size = ndocs)) +
  labs(x = "To", y = "From", col = "Links", size = "Help pages",
       title = "Links between R help pages:") +
  theme_minimal() +
  scale_x_discrete(position = "top", guide = guide_axis(n.dodge = 2))  +
  theme(plot.title.position = "plot", 
        axis.title.y = element_text(hjust = 0, vjust = 1, angle = 0),
        axis.title.x = element_text(hjust = 0),
        )
```

We can clearly see that the packages link more between their own package but there are some heavily connected packages like utils and base or stats and base. 

While so far we focus on links if we compare that with the existing help pages we might find some interesting patterns:

```{r}
#| label: plot-base-links-files
#| eval: !interactive()
#| fig-cap: "Graph of all the links between help pages of base R. There are some help pages that are not connected to any other help page"
rd_origin <- summarise(xrefBase3, 
          .by = c(Rd_origin, Package, to_pkg, Rd_destiny), 
          n = n()) |> 
  mutate(from = paste0(Package, ":", Rd_origin),
         to = paste0(to_pkg, ":", Rd_destiny)) |> 
  mutate(.by = to, n_to = n_distinct(from)) |> 
  mutate(.by = from, n_from = n_distinct(to)) |> 
  filter(to_pkg %in% names(aliases_base))

g <- rd_origin |> 
  select(-Rd_origin, -to_pkg, -Rd_destiny) |> 
  as_tbl_graph()

ai <- as.igraph(g)
aid <- igraph::decompose(ai)

isolated_help <- sapply(aid[-1], igraph::vertex_attr) |> 
  unlist(use.names = FALSE)

library("tidygraph")
library("ggraph")
g |> 
  ggraph() +
  geom_edge_link(aes(col = n), alpha = 0.15) + 
  geom_node_point(alpha = 0.15) +
  theme_graph() +
  labs(title = "Links between R help pages")
```

Apparently there are some help page on base R that don't have links to navigate to the rest of the documentation.
These are `r isolated_help`.
These help pages will be easier to find them if at least they were linked from other help pages from base R.

### To CRAN

We can also ask which packages are linked from R that are not part of R?

```{r}
#| label: cran-database
db <- tools::CRAN_package_db()

xrefBase3 |> 
  filter(!to_pkg %in% names(aliases_base)) |> 
  count(to_pkg, sort = TRUE, name = "links") |> 
  left_join(db[, c("Package", "Priority", "Maintainer")],
            by = join_by(to_pkg == Package),
            multiple = "first") |> 
  mutate(Maintainer = gsub(" <.+", "", Maintainer),
         `Core member` = Maintainer %in% c("Brian Ripley", "Martin Maechler", "Deepayan Sarkar", "R Core Team", "Kurt Hornik")) |> 
  arrange(-links, Priority, -`Core member`, to_pkg) |> 
  rename(Package = to_pkg)
```

We can see that some packages linked from base R are from the recommended packages and many from the R core members or maintained by the whole R core group.
Some other links include popular packages, while others are not so popular. 

# CRAN packages

Now that we have the tools ready and we already started exploring the relation from base R to CRAN we can explore all the CRAN packages and their relation with base R.

## Aliases

As before we start exploring CRAN aliases:

```{r}
#| label: cran-aliases
aliases_CRAN <- tools:::CRAN_aliases_db()
aliasesCRAN <- alias2df(aliases_CRAN)

dup_cran <- aliasesCRAN |> 
  summarise(.by = Target, n = n_distinct(Package), 
            Package = paste(Package, collapse = ", ")) |> 
  arrange(-n) |> 
  filter(n > 1)

aliasesCRAN |> 
  summarise(.by = Package, n = n_distinct(Source)) |> 
  arrange(-n) |> 
  head(10)
```

We can also find those targets that are duplicated on CRAN:

```{r}
#| label: plot-cran-dup-targets
#| fig.cap: "**Top 25 duplicated topics on CRAN.** Common methods and the magrittr pipe dominate the most frequent topics."
dup_cran |> 
  head(25) |> 
  ggplot() +
  geom_col(aes(n, fct_reorder(Target, n))) +
  scale_x_continuous(expand = expansion(mult = c(0, NA), add = c(0, NA)), 
                     limits = c(0, NA)) +
  labs(x = "Packages", y = element_blank(),
       title = "Top 50 topics more repeated on CRAN") +
  theme_minimal() +
  theme(plot.title.position = "plot")
```

The most repeated topic is the magrittr pipe `%>%` followed by reexports (curious as usually the pipe is also re-exported).
We can see some common methods like `plot`, `print`, `show` and some that are related to the dplyr and shiny, mixed between some methods used for statistics. 

```{r}
#| label: plot-dup-targets

# library("tidygraph")
# library("ggraph")
# ab <- aliasesCRAN |>
#   filter(.by = Target, n_distinct(Package) > 1) |>
#   arrange(Target) |>
#   select(Target, Package)


# as_tbl_graph() |> 
# mutate(Popularity = centrality_degree(mode = 'in')) |> 
# ggraph(layout = "kk") +
# geom_edge_fan(aes(alpha = after_stat(index)), show.legend = FALSE) + 
# geom_node_point(aes(size = Popularity)) + 
# theme_graph(fg_text_colour = 'white')
```



```{r}
#| label: plot-cran-help-topics
#| fig-lab: "**Packages with more topics**"

hcran <- aliasesCRAN |> 
  summarise(.by = c(Source, Package), n = n()) |> 
  arrange(-n) |> 
  mutate(.by = Package, help_pages = n())

hcran |> 
  summarise(.by = Package, n = sum(n)) |> 
  arrange(-n) |> 
  head(25) |> 
  ggplot() +
  geom_col(aes(n, fct_reorder(Package, n, .fun = "sum"))) +
  scale_x_continuous(expand = expansion(), 
                     sec.axis = dup_axis(name = element_blank())) +
  labs(title = "CRAN's packages with more topics",
       x = "Topics", y = element_blank(), size = "Packages") +
  theme_minimal() +
  theme(plot.title.position = "plot")
```

These packages have high number of topics, but this is because they have a high number of help pages? Let's check it:

```{r}
#| label: plot-cran-help-pages
#| fig-lab: "**Packages with more help pages**"
hcran |> 
  summarise(.by = Package, n_h = unique(help_pages)) |> 
  arrange(-n_h) |> 
  head(25) |> 
  ggplot() + 
  geom_col(aes(n_h, fct_reorder(Package, n_h, .fun = "sum"))) +
  labs(y = element_blank(), x = "Help pages", title = "Packages' help pages on CRAN") +
  scale_x_continuous(expand = expansion(), 
                     sec.axis = dup_axis(name = element_blank())) +
  theme_minimal() +
  theme(plot.title.position = "plot")
```


It is clear that more help pages might lead to more topic pages but how much?

```{r}
#| label: plot-cran-topics-helps
#| fig-cap: "**Topics per help page is very high on some packages** Other have
#|  huge ammounts of help pages."

vec <- hcran |> 
  summarise(.by = Package, topics = sum(n), hp = n()) |> 
  summarise(t = median(topics), h = median(hp)) |> 
  simplify2array()
  hcran |> 
  summarise(.by = Package, topics = sum(n), hp = n()) |> 
  ggplot() +
  geom_count(aes(hp, topics), alpha = 0.15) +
  scale_x_continuous(expand = expansion(mult = 0.05)) +
  scale_y_continuous(expand = expansion(mult = 0.05)) +
  scale_size(breaks = c(1, 100, 200, 400, 600, 800)) +
  geom_text_repel(aes(hp, topics, label = Package), col = "darkgreen",
                   data = . %>% filter(topics >= 1000 & hp <= 750)) +
  geom_text_repel(aes(hp, topics, label = Package), 
                   data = . %>% filter(hp > 750)) +
  labs(x = "Help pages", y = "Topics", title = "Topis per help page",
       size = "Packages") +
  theme_minimal() +
  theme(plot.title.position = "plot")
```

Surprisingly there are some package that have many help pages and many topics (those in red) and some that the high number of topics comes from the high number of help pages.

However this is not normal, the median number of help pages is `r unames(vec["h"])` and the median number of topics is `r uname(vec["t"])`. 
So usually CRAN packages tend to be much smaller. 

## Cross references

There are lots of packages and lots of cross-references working with them is slow.

Similar to what we did with R cross-references we first start finding links to base packages.
Then we search for links across CRAN packages.

```{r}
#| label: cran-split-dest
rdxrefs_CRAN <- tools:::CRAN_rdxrefs_db()
rdxrefsCRAN <- xrefs2df(rdxrefs_CRAN)
pkg_no_links <- setdiff(names(rdxrefs_CRAN), rdxrefsCRAN$Package)
rdxrefsCRAN <- cbind(rdxrefsCRAN, 
                     strcapture("([[:alnum:].]*{2,})?[:=]?(.*)", 
                                x = rdxrefsCRAN$Anchor, 
                                proto = data.frame(to_pkg = character(), 
                                                   to_target = character())))

# Fill with base references
rdxrefsCRAN <- fill_xref(rdxrefsCRAN, aliasesBase, dtBase$Alias)

# Fill with CRAN targets
fill_xref_cran <- function(z, alias = aliasesCRAN, duplicates = dup_cran[["Target"]]) {
  # No anchors assuming they are from CRAN
  no_anchor <- !nzchar(z$Anchor)
  no_anchor_no_dup <- no_anchor & is.na(z[["to_pkg"]]) & !z[["Target"]] %in% duplicates
  match_target <- match(z[["Target"]][no_anchor_no_dup], alias[["Target"]])
  
  # Adding values
  z[["to_pkg"]][no_anchor_no_dup] <- alias[["Package"]][match_target]
  z[["to_target"]][no_anchor_no_dup] <- alias[["Target"]][match_target]
  
  # Adding the package of those that have anchors 
  missing_but_anchor <- is.na(z[["to_pkg"]]) & startsWith(z[["Anchor"]], "=")
  z[["to_pkg"]][missing_but_anchor] <- z[["Package"]][missing_but_anchor]
  
  # Adding duplicated Targets/topics at other packages but present on the package
  blank_target <- is.na(z$to_target)
  for (pkg in unique(z[["Package"]][blank_target])) {
    z_keep <- z[["Package"]] == pkg & blank_target
    alias_keep <- alias[["Package"]] == pkg
    m <- match(z[["Target"]][z_keep], alias[["Target"]][alias_keep])
    z[["to_target"]][z_keep] <- alias[["Target"]][alias_keep][m]
    z[["to_pkg"]][z_keep] <- alias[["Package"]][alias_keep][m]
  }
  z
}

rdxrefsCRAN2 <- fill_xref_cran(rdxrefsCRAN)
rdxrefsCRAN3 <- merge(rdxrefsCRAN2, rbind(aliasesBase, aliasesCRAN), 
            by.x = c("to_pkg", "to_target"), 
            by.y = c("Package", "Target"), 
            all.x = TRUE, sort = FALSE) |> 
  select(Rd_origin = Source.x, Anchor, Target, Package, to_pkg, 
         to_target, Rd_destiny = Source.y)
```

After that we merge the data with the files they are at.

```{r}
#| label: base-links-cran
missing <- xrefBase3 |> 
  filter(is.na(Rd_destiny)) |> 
  select(-Rd_destiny, Source = Rd_origin)
xrefBase3_m <- merge(missing, aliasesCRAN, 
            by.x = c("to_pkg", "to_target"), 
            by.y = c("Package", "Target"), 
            all.x = TRUE, sort = FALSE) |> 
  select(Rd_origin = Source.x, Anchor, Target, Package, to_pkg, to_target, Rd_destiny = Source.y)

xrefBase4 <- xrefBase3 |> 
  filter(!is.na(Rd_destiny)) |>
  rbind(xrefBase3_m) |> 
  arrange(Package, Rd_origin)
```

Finally, as I realized checking CRAN links[^4] we add links that use the file names instead of topics[^5].

[^4]: I revisited some code after finding issues further down this post.

[^5]: Yes, even though the documentation makes it clear that this shouldn't be done the help system works around this issue. I raised this issue on the [mailing list](https://stat.ethz.ch/pipermail/r-devel/2024-September/083626.html) and later they were fixed on report [18815](https://bugs.r-project.org/show_bug.cgi?id=18815) before publishing this post.



```{r}
#| label: base-links-files
#| eval: false

# This was used to detect those packages linking to a file.
missingBase <- xrefBase4 |> 
  mutate(n = seq_len(n())) |> 
  filter(is.na(Rd_destiny) & !is.na(to_pkg) & !is.na(to_target)) |> 
  select(-Rd_destiny, Source = Rd_origin)

aliasesCRAN$filename <- tools::file_path_sans_ext(aliasesCRAN$Source)
aliasesBase$filename <- tools::file_path_sans_ext(aliasesBase$Source)
all_aliases <- rbind(aliasesCRAN, aliasesBase)
xrefBase4_m <- merge(missingBase, all_aliases[, -2], 
            by.x = c("to_pkg", "to_target"), 
            by.y = c("Package", "filename"), 
            all.x = TRUE, sort = FALSE, all.y = FALSE) |> 
  select(Rd_origin = Source.x, Anchor, Target, Package, to_pkg, to_target,
         Rd_destiny = Source.y, n) |> 
  distinct(.keep_all = TRUE) |> 
  mutate(Rd_destiny = if_else(!is.na(Rd_destiny), paste0("=", Rd_destiny), NA))

xrefBase5 <- xrefBase4 |> 
  mutate(n = seq_len(n())) |> 
  filter(!n %in% xrefBase4_m$n) |>
  rbind(xrefBase4_m) |> 
  mutate(to_file = startsWith(Rd_destiny, "="),
         Rd_destiny = gsub("^=", "", Rd_destiny)) |> 
  arrange(Package, Rd_origin, n) |> 
  select(-n)

stopifnot(sum(startsWith(xrefBase4_m$Rd_destiny, "="), na.rm = TRUE) == sum(xrefBase5$to_file))
stopifnot(nrow(xrefBase4) == nrow(xrefBase5))
```





```{r}
#| label: cran-links-files
# Try to merge those that we don't know the destiny but we know the other things
missingCRAN <- rdxrefsCRAN3 |> 
  mutate(n = seq_len(n())) |> 
  filter(is.na(Rd_destiny) & !is.na(to_pkg) & !is.na(to_target)) |> 
  select(-Rd_destiny, Source = Rd_origin) 

rdxrefsCRAN3_m <- merge(missingCRAN, all_aliases[, -2], 
            by.x = c("to_pkg", "to_target"), 
            by.y = c("Package", "filename"), 
            all.x = TRUE, sort = FALSE) |> 
  distinct(.keep_all = TRUE) |> 
  select(Rd_origin = Source.x, Anchor, Target, Package, to_pkg, to_target,
         Rd_destiny = Source.y, n) |> 
  mutate(Rd_destiny = if_else(!is.na(Rd_destiny), paste0("=", Rd_destiny), NA))

rdxrefsCRAN4 <- rdxrefsCRAN3 |> 
  mutate(n = seq_len(n())) |> 
  filter(!n %in% rdxrefsCRAN3_m$n) |>
  rbind(rdxrefsCRAN3_m) |> 
  mutate(to_file = if_else(startsWith(Rd_destiny, "="), TRUE, FALSE),
         Rd_destiny = gsub("^=", "", Rd_destiny)) |> 
  arrange(Package, Rd_origin, n) |> 
  select(-n)

stopifnot(sum(startsWith(rdxrefsCRAN3_m$Rd_destiny, "="), na.rm = TRUE) == sum(rdxrefsCRAN4$to_file, na.rm = TRUE))
stopifnot(nrow(rdxrefsCRAN3) == nrow(rdxrefsCRAN4))
```

Now that I have corrected many issues I'll check if it matches CRAN's output.

## Comparing links with CRAN's checks

We can compare the data with CRAN's package details of issues found related to cross-references .

```{r}
#| label: cran-details
ccd <- tools::CRAN_check_details()
crans_issues <- ccd |> 
  as.data.frame() |> 
  filter(Check == "Rd cross-references") |> 
  distinct(Package) |> 
  mutate(CRAN_checks = "rd_cross_reference")
```

There are `r nrow(crans_issues)` packages with issues in "Rd cross-references" across all flavors.
These can be found online too, for example for [r-devel in debian](https://cran.r-project.org/web/checks/check_details_r-devel-linux-x86_64-debian-clang.html), (pick your [flavor](https://cran.r-project.org/web/checks/check_summary.html)). 

To compare with CRAN we need to prepare the data collected. 
First thing, CRAN packages can depend on packages in Bioconductor and other additional repositories for which we don't have the data. 
Identifying those packages will help to avoid any false positive. 

```{r}
#| label: cran-summaries-details
# Collect Bioconductor packages
ap_bioc <- available.packages(repos = BiocManager::repositories()[1:5],
                              filters = character())
bioc_packages <- rownames(ap_bioc)

# "Merge" the two repositories (Keeping those windows specific)
db$Repository <- "CRAN"
db$File <- NA_character_
ap_all <- rbind(db[, colnames(ap_bioc)], ap_bioc)

tpd <- tools::package_dependencies(db$Package, 
                            db = ap_all, 
                            which = c("Depends", "Imports", "Suggests", "Enhances"))

in_tpd <- function(pkg) {
  tpd[[unique(pkg)]]
}

additional_repo <- function(x) {
  any(!is.na(db$Additional_repositories[db$Package == x]))
}

summary_CRAN <- rdxrefsCRAN4 |> 
  summarise(.by = Package, 
            to_file = any(to_file),
            additional_repo = additional_repo(unique(Package)),
            to_file_w_eq_anchor = any(to_file & startsWith(Anchor, "=")),
            missing_destination = any(is.na(Rd_destiny)),
            missing_anchors = any(!nzchar(Anchor)),
            linking2base = any(!to_pkg %in% names(aliases_base) & !is.na(to_pkg)),
            linking2cran = any(!is.na(to_pkg) & !to_pkg %in% names(rdxrefs_CRAN)),
            linking2bioc = any(to_pkg %in% bioc_packages),
            missing_dependency = any(!is.na(to_pkg) & !to_pkg %in% c(Package, in_tpd(Package)) & !to_pkg %in% names(aliases_base))
  ) |> 
  mutate(to_file = if_else(is.na(to_file), FALSE, to_file))
```

To further check what is happening we can compare directly with what CRAN reports.


```{r}
#| label: cran-comparison
summary_CRAN2 <- merge(crans_issues, summary_CRAN, all = TRUE)
csc <- count(summary_CRAN2, CRAN_checks, to_file)
```

There are `r csc$n[csc$to_file & !is.na(csc$CRAN_checks)]` packages currently with cross references notes that could be improved by using topics on the cross-references  (instead of file names).
There are `r csc$n[csc$to_file & is.na(csc$CRAN_checks)]` that are currently not detected by CRAN that could be improved the same way.

Now that we found that they mostly agree let's analyse the data.

```{r}
#| label: cran-checks2
#| eval: FALSE
summary_CRAN2 |> 
  # filter(is.na(CRAN_checks)) |>
  summarise(.by = c(missing_anchors, to_file, missing_dependency),
            n = sum(!additional_repo)) |> 
  arrange(-n) |> 
  mutate(ratio = scales::percent(n/sum(n))) 
  

summary_CRAN2|> count(CRAN_checks, to_file_w_eq_anchor, missing_destination, missing_dependency, missing_anchors)

```


# Further analysis



## Links to base

First we can check the number of links to base R:

```{r}
#| label: plot-base-links
#| fig.cap: "**Base is the most linked package** Packages sorted by number of
#|   links they receive from CRAN's packages"
to_base_links <- rdxrefsCRAN4 |> 
  filter(to_pkg %in% names(aliases_base))
base_links <- count(to_base_links, Package, to_pkg, sort = TRUE) 

base_links |> 
  ggplot() +
  geom_col(aes(n, fct_reorder(to_pkg, n, .fun = "sum"))) +
  scale_x_continuous(expand = expansion()) +
  guides(col = "none", fill = "none") +
  labs(x = "Links", y = element_blank(), title = "Links to base packages") +
  theme_minimal()
```

From the above plot I find surprising there are these few links to the utils package. 
Perhaps looking at which files are linked could help.

```{r}
#| label: base-links-packages
#| fig.cap: "**Links to base from different pacakges**"

base_links_sources <- to_base_links |> 
  summarise(.by = c(to_pkg, Rd_destiny), 
            n = n(), 
            packages = n_distinct(Package), 
            targets = n_distinct(to_target)) |> 
  arrange(to_pkg, -n, -packages)

base_links_sources |> 
  ggplot() +
  geom_jitter(aes(n, fct_reorder(to_pkg, n, .fun = "sum"), col = to_pkg, size = packages), width = 0) +
  scale_x_continuous(expand = expansion(), trans = "log10") +
  guides(col = "none", fill = "none") +
  labs(x = "Links", y = element_blank(), title = "Links to base packages") +
  theme_minimal()
```

As previously reported there are many links to different help pages across packages.
Many of the most popular pages are linked from different many different packages too.
Surprisingly all links to the compiler package are to a single page: `help("compile", package = "compiler")`

```{r}
#| label: base-links-sources
#| fig.cap: "**Most linked pages by packages** Some data types are the most 
#|  linked as well as some plot parameters and methods for model fitting or 
#|   representation."
base_links_sources |> 
  ggplot(aes(packages, n, col = fct_reorder(to_pkg, -n, .fun = "sum"))) +
  geom_point(aes(size = targets)) +
  geom_text_repel(aes(label = Rd_destiny), data = . %>% filter(n > 1000 | packages > 400), show.legend = c(col = FALSE) ) +
  labs(col = "Package", title = "Packages with links to different base R documents", x = "Links from different packages", y = "Links") + 
  theme_minimal()
```


```{r}
#| label: missing-anchors
missing_anchors <- rdxrefsCRAN3 |> 
  summarise(.by = Package, missing = sum(is.na(to_target)), n = n()) |> 
  arrange(-missing)

help_pages <- summarise(aliasesCRAN, .by = Package, 
                        `Help pages` = n_distinct(Source),
                        Topics = n_distinct(Target))
help_pages2 <- aliasesCRAN |> 
  summarise(.by = c(Package, Source),
                        Topics = n_distinct(Target)) |> 
  summarise(.by = Package, across(Topics, 
                                  .fns = c(mean, median, min, max),
                                  .names = c("mean", "median", "min", "max")))
hp <- merge(help_pages, help_pages2, sort = FALSE)

ggplot(hp) +
  geom_count(aes(`Help pages`, Topics, col = median)) +
  geom_hline(yintercept = median(hp$Topics), linetype = "dashed", col = "black") +
  geom_vline(xintercept = median(hp$`Help pages`), linetype = "dashed", col = "black") +
  annotate("text", label = "More help pages and topics\n than the median", 
           x = 400, y = 40) +
  scale_size_binned_area(limits = c(1, NA), name = "Packages") +
  scale_color_continuous(name = "Median topics per page") +
  labs(title = "CRAN packages documentation practice per package") +
  scale_x_log10() +
  scale_y_log10()
```


```{r}
#| label: links
pkg_missing <- data.frame(Package = pkg_no_links, 
                          links = 0, diff_source = 0, diff_target = 0,
                          Links_mean = 0, Links_median = 0, Links_min = 0,
                          Links_max = 0)
pkg_target <- rdxrefsCRAN4 |> 
  summarise(.by = Package, 
            links = n(), 
            diff_source = n_distinct(Rd_origin),
            diff_target = n_distinct(to_pkg, Rd_destiny),
            to_file = any(to_file),
            additional_repo = additional_repo(unique(Package)),
            to_file_w_eq_anchor = any(to_file & startsWith(Anchor, "=")),
            missing_destination = any(is.na(Rd_destiny)),
            missing_anchors = any(!nzchar(Anchor)),
            linking2base = any(!to_pkg %in% names(aliases_base) & !is.na(to_pkg)),
            linking2cran = any(!is.na(to_pkg) & to_pkg != rdxrefs_CRAN),
            linking2bioc = any(to_pkg %in% bioc_packages),
            missing_dependency = any(!is.na(to_pkg) & !to_pkg %in% c(Package, in_tpd(Package)) & !to_pkg %in% names(aliases_base))
  ) |> 
  mutate(to_file = if_else(is.na(to_file), FALSE, to_file))
pkg_target2 <- rdxrefsCRAN4 |> 
  summarise(.by = c(Package, Rd_origin),
            Links = n_distinct(to_pkg, Rd_destiny)) |> 
  summarise(.by = Package, across(Links, 
                                  .fns = c(mean = mean, median = median, 
                                           min = min, max = max)))
pt <- merge(pkg_target, pkg_target2, sort = FALSE) |> 
  full_join(pkg_missing, by = join_by("Package"), fill = 0)

ggplot(pt) +
  geom_abline(slope = 1, intercept = 0) +
  geom_count(aes(diff_source, diff_target, col = links)) +
  geom_hline(yintercept = median(pt$diff_source), linetype = "dashed", col = "black") +
  geom_vline(xintercept = median(pt$diff_target), linetype = "dashed", col = "black") +
  scale_size_binned_area(limits = c(1, NA), name = "Packages",
                         breaks = c(5, 10, 50, 100, 500, 1000),
                         trans = "log10") +
  scale_color_continuous(name = "Links", trans = "log10") +
  labs(title = "CRAN packages' links", 
       x = "Links from different help pages",
       y = "Links to different help pages") +
  scale_x_continuous(trans = scales::pseudo_log_trans(base = 10), 
                     breaks = c(0, 1, 10, 50, 100, 200, 400, 800)) +
  scale_y_continuous(trans = scales::pseudo_log_trans(base = 10), 
                     breaks = c(0, 1, 10, 100, 1000, 10000))
```

Surprisingly there are more than 7000 packages with no links to other documentation pages, while the median is to link to `r median(pt$diff_target)` different help pages, from `r median(pt$diff_source)` different help pages.

We can see some packages that have many help pages but only link to the same help page and some other packages that have few help pages that link to several pages.


## CRAN packages

Now it is time to check for CRAN packages.

There are `r sum(summary_CRAN2$missing_destination)` packages that lack destination of the R document. 
There are also `r sum(summary_CRAN2$missing_anchors)` that miss anchors, the majority of these are to their own R documentation help pages. 
There are `r sum(summary_CRAN2$missing_dependency)` packages that link to a packages that is not on the dependency, while these could be packages using "LinkingTo" this could also be packages linking to archived packages.
There are `r sum(summary_CRAN2$to_file)` packages that link to a file instead of a topic. While this is not forbidden (yet)


```{r}
cran_to <- rdxrefsCRAN4 |> 
  summarise(.by = to_pkg, hp = n_distinct(Rd_destiny), pkges = n_distinct(Package), n = n()) |> 
  arrange(-n, -pkges, -hp)

cran_to |> 
  ggplot() +
  geom_point(aes(pkges, hp, size = n)) +
  geom_text_repel(aes(pkges, hp, label = to_pkg), data = . %>% filter(pkges > 1000 | hp > 750)) +
  scale_x_continuous(expand = expansion(add = 1)) +
  scale_y_continuous(expand = expansion(add = 1)) +
  theme_minimal()
```

```{r}
cran_to |> 
  filter(pkges < 1000, hp < 750) |> 
  ggplot() +
  geom_point(aes(pkges, hp, size = n)) +
  geom_text_repel(aes(pkges, hp, label = to_pkg), data = . %>% filter(pkges > 200 | hp > 250  & pkges > 3)) +
  scale_x_continuous(expand = expansion(add = 1)) +
  scale_y_continuous(expand = expansion(add = 1)) +
  theme_minimal()
```



```{r}
missing_anchors |> 
  filter(missing >= 1) |> 
  left_join(hp, by = join_by(Package)) |> 
  mutate(rate = missing/`Help pages`) |> 
  ggplot() +
  geom_point(aes(`Help pages`, n, size = rate, col = Topics)) +
  geom_hline(yintercept = median(hp$Topics), linetype = "dashed", col = "black") +
  geom_vline(xintercept = median(hp$`Help pages`), linetype = "dashed", col = "black") +
  annotate("text", label = "More help pages and topics\n than the median", 
           x = 40, y = 10000) +
  scale_size_binned_area(labels = scales::label_percent(), limits = c(0, 1)) +
  labs(y = "Links", title = "Packages with missing links", 
       size = "Missing rate") +
  scale_x_log10() +
  scale_y_log10()
```


```{r}
#| label: source-alias
sources_alias <- aliasesCRAN |> 
  summarise(.by = packages, 
            n_sources = n_distinct(Source), 
            n_alias = n_distinct(Target)) |> 
  arrange(-n_sources, -n_alias) 

sources_alias |> 
  count(n_sources, n_alias) |> 
  ggplot() + 
  geom_point(aes(n_sources, n_alias, size = n))
```


<details>
<summary>Missing anchors in your package but not a note from CRAN?</summary>

While checking this I realized packages I maintain don't have anchors and I explored why.

The reasons is I use [roxygen2](https://cran.r-project.org/package=roxygen2) for documentation, so for all links and cross-references. 
However, roxygen2 cross-references in `@family tag` do not follow the same pattern as other cross-references. See [the issue](https://github.com/r-lib/roxygen2/issues/1650) for more information.
</details>


```{r eval=FALSE}
#| label: kurt-hornik-code
rdxrefs <- tools:::CRAN_rdxrefs_db()
## Note that as of 2024-06-11, 7139 CRAN packages have no Rd xrefs:
(n1 <- sum(lengths(rdxrefs) == 0L))

xdf <- tools:::.Rd_rdxrefs_db_to_data_frame(c(tools:::base_rdxrefs_db(),
                                              rdxrefs))

## The ones with package anchors:
ind <- with(xdf, nzchar(Anchor) & !startsWith(Anchor, "="))

targets_with_no_package_anchor <- function(x) {
    target <- x$Target
    anchor <- x$Anchor
    source <- x$Source
    names(target) <- source
    names(anchor) <- source
    c(target[!nzchar(anchor)],
      substring(anchor[startsWith(anchor, "=")], 2L))
}

## The ones without package anchors, split according to package:
rdxrefs <- tapply(xdf[!ind, ],
                  sub("::.*", "", xdf$Source[!ind]),
                  targets_with_no_package_anchor,
                  simplify = FALSE)

## As of 2024-06-11, 12304 CRAN packages have Rd xrefs without package
## anchors:
(n2 <- length(rdxrefs))

aliases_in_base <- tools:::base_aliases_db()
aliases_in_CRAN <- tools:::CRAN_aliases_db()

aliases_for_level_0 <-
    lapply(c(aliases_in_base, aliases_in_CRAN),
           unlist, use.names = FALSE)
aliases_for_level_1 <-
    unique(unlist(aliases_in_base, use.names = FALSE))

targets_needing_anchors <-
    Map(function(x, p) {
        x[is.na(match(x,
                      c(aliases_for_level_0[[p]],
                        aliases_for_level_1)))]
    },
    rdxrefs,
    names(rdxrefs))

## As of 2024-06-11, 9860 CRAN packages have Rd xrefs without package
## anchors but do not need any (all non-package-anchored targets are to
## level 0 and new-level-1 base-package only aliases).
(n3 <- sum(lengths(targets_needing_anchors) == 0L))

## This leaves 2444 CRAN packages which need to add package anchors to
## their Rd xrefs:
(n4 <- length(rdxrefs) - sum(lengths(targets_needing_anchors) == 0L))
## Numbers of Rd xrefs needing package anchors:
(tab <- table(lengths(targets_needing_anchors)))
```


# Conclusions

As a summary of the analysis:

 - Topic more repeated (Advancement: %>%)
 - Help page more linked to
 - Help page with more links
 - Base package more linked to
 - Base page more linked to: character
 - Base page more linked from different packages: par.Rd
 - Base package with more links: base
 - Package with more help pages: 
 - Package linked more
 - Packages with more help pages
 
 
Wrong assumptions I had about cross-references:

 - All packages will have one link to other help pages.
 - One cannot link to a file. 
 - Base packages wouldn't link to packages.
 - 

### Reproducibility

<details>
<summary>Session info</summary>
```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info(info = c("platform", "packages"))
```
</details>
