---
title: "Exploring CRAN's files: part 3"
subtitle: 'References: links between pages, packages and base R'
author: Lluís Revilla Sancho
date: '2024-09-23'
slug: cran-files-3
categories:
  - CRAN
  - r
tags:
  - r
  - packages
  - cran-files
authors:
  - admin
description: ''
draft: no
editor_options:
  chunk_output_type: console
featured: yes
image:
  caption: ''
  focal_point: ''
summary: 'Exploring references'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.retina = 2, cache = TRUE, include = TRUE, echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, eval = TRUE)
Sys.setlocale("LC_ALL", "C")
Sys.setenv(R_CRAN_WEB = "https://cloud.r-project.org")
```


In the blog post of CRAN files, I [initially explored the database](https://llrs.dev/post/2022/07/23/cran-files-1/) (mostly around `CRAN_package_db()`). On the [second post](https://llrs.dev/post/2022/07/28/cran-files-2/) I moved to analyze the archive around `CRAN_archive_db()`.

Recently at useR!2024, Kurt Hornik offered to export data from CRAN, after an email [several new functions were exported](https://developer.r-project.org/blosxom.cgi/R-devel/NEWS/2024/08/22#n2024-08-22) and introduced to base R.

In that presentation and in a previous exchange with Kurt, he explained his project about providing the HTML manual pages of all CRAN packages.
One of the problems with this projects is providing links to the right pages of the packages (others are making it accessible for all users).

In R-devel now there are test that ensure links to other pages are in the right format [^1].
In this post we we'll explore links between documentation pages, what do manuals write about it and what is the state of base R and the packages.

[^1]: By setting `_R_CHECK_XREFS_NOTE_MISSING_PACKAGE_ANCHORS_=true` as environmental variable.

# Introduction

Help pages are defined in R documentation files, commonly with the `.Rd` extension.
Each documentation file can have multiple topics.
A topic is defined in R documentation format with `\alias{topic}`, so I might use interchangeably alias and topics.
Links, or cross-references (xref for short) [should be to an alias](https://cran.r-project.org/doc/manuals/r-devel/R-exts.html#Cross_002dreferences), not to files (allowing to move topics between help pages). 

One can define links with `\link{alias}` or `\link[=alias]{name}` if the link is to a different place than the name [^2]. 
In the second form the content within the square bracket is known as anchor.

There are two other forms accepted: `\link[pkg]{alias}` and `\link[pkg:alias]{name}` which do not use the = inside the square brackets [^3]. 
These are only used in the HTML format. 
Packages referred to on these square brackets should be declared in the DESCRIPTION file, in the ‘Depends’, ‘Imports’, ‘Suggests’ or ‘Enhances’ fields.

One important thing is that links are case sensitive: see that `?Complex` leads to a different help page than `?complex`. 

[^2]: There is also the `\linkS4class{abc}` that expands to `\link[=abc-class]{abc}` for S4 classes.

[^3]: So there are three names for (almost) the same thing: topic, alias and anchor. Anchors do not need to be = but should be resolvable.

With this in mind we can check to which packages have more manual pages, have more cross-references, or have an overview help page (those with *pkgname*-package.Rd).
We will also check which documentation pages have more links, are more linked to, or needs its links fixed.

# Base R

Previously there wasn't a way to know the links and aliases in base R since recently (~2024/08/20) there are a new way to get that. 

```{r base}
aliases_base <- tools::base_aliases_db()
xrefs_base <- tools:::base_rdxrefs_db()
```


## Aliases

```{r}
#| label: alias2df
alias2df <- function(x){
  l <- lapply(x, function(x) {
    data.frame(Source = rep(names(x), lengths(x)),
               Target = unlist(x, FALSE, FALSE))
  })
  aliasesDF <- do.call(rbind, l)
  aliasesDF$Package <- rep(names(l), vapply(l, NROW, numeric(1L)))
  rownames(aliasesDF) <- NULL
  aliasesDF[, c("Source", "Target", "Package"), drop = FALSE]
}

aliasesBase <- alias2df(aliases_base)
library("dplyr")
```

There are `r dplyr::n_distinct(aliasesBase$Source)` help pages with `r dplyr::n_distinct(aliasesBase$Target)` topics/aliases.

```{r}
#| label: base-dup-files
dup_files <- aliasesBase |> 
  select(-Target) |> 
  distinct() |> 
  summarise(.by = Source, 
            dup_pages = n(),
            packages = paste(Package, collapse = ", ")) |> 
  arrange(-dup_pages) |> 
  filter(dup_pages > 1L) |> 
  head()

```

There are some help files with the same name in base R that can be accessed via their targets (via `?Target`). 
Those files are `r dup_targets$Source`.

To distinguish between those files one need to check the title(s) and content of the pages.

Similarly there are multiple help pages of different packages with the same topics:

```{r}
#| label: base-dup-targets
dtBase <- aliasesBase |> 
  summarise(.by = Target, n = n(), Package = paste0(Package, collapse = ", ")) |> 
  filter(n > 1) |> 
  arrange(Target, Package) |> 
  select(Alias = Target, Packages = Package)
dtBase
```

Most of them are in the base and other packages, from the methods package to tools, grDevices, graphics, and utils.


```{r}
#| label: base-dup-pages
dup_pages <- aliasesBase |> 
  filter(.by = Source, Target %in% dup_targets$Alias) |> 
  filter(.by = c(Source, Package), n() == 1) |> 
  select(Package, Target, Source) |> 
  arrange(Target, Package, Source)
```

From these there are `r nrow(dup_pages)` help pages that do not have any other alias.
To access them directly one needs to use `help()`: `help("clipboard", package = "utils")` otherwise one will need to choose them from the menu it will pop up from using `?clipboard`.
This works to access any other help page too.

Last, we can check the topics for the packages' help page.
While usually packages have two alias `pkg` and one `pkgname-package` there are two packages that do not have this, the grid package and the methods package:


```{r}
#| label: base-package.rd
aliasesBase |> 
  filter(grepl("-package.Rd", Source)) |> 
  filter(.by = Package, sum(Target == Package, 
                            grepl(unique(paste0(Package, "-package")), Target)) != 2) |> 
  select(Package, Source, Alias = Target)
```

These packages lack the form of `pkgname` to find the package help file.

## Cross references

We can check to where do these help pages link to, starting by knowin how many are directly to the alias or those that use a name:

```{r}
#| label: xrefsBase
#| fig-cap: 
xrefs2df <- function(x) {
  rdxrefsDF <- do.call(rbind, x)
  rdxrefsDF <- as.data.frame(rdxrefsDF)
  rdxrefsDF$Package <- rep(names(x), vapply(x, NROW, numeric(1L)))  
  rownames(rdxrefsDF) <- NULL
  rdxrefsDF[, c("Source", "Target", "Anchor", "Package"), drop = FALSE]
  rdxrefsDF
}
xrefBase <- xrefs2df(xrefs_base)
```


```{r}
#| label: links-base
#| fig-cap: "Packages with links using alias per help page."
library("ggplot2")

xref_summary <- xrefBase |> 
  group_by(Package, Source) |> 
  count(Change_name = nzchar(Anchor)) |> 
  ungroup() |> 
  arrange(Package, Source, Change_name) |> 
  mutate(.by = c(Package), help_pages = n_distinct(Source)) |> 
  mutate(.by = c(Package, Source), ratio_source = sum(n[Change_name])/sum(n)) |> 
  summarise(.by = c(Package, Change_name, n, help_pages, ratio_source), size = n()) |> 
  mutate(.by = c(Package), ratio_pkg = sum(n[Change_name])/sum(n)) 

xref_summary |> 
  ggplot() +
  geom_jitter(aes(ratio_source, fct_reorder(Package, help_pages, .fun = max), 
                  size = n), width = 0) +
  scale_size(limits = c(1, NA), breaks = c(1, seq(from = 10, to = 100, by = 10))) +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(x = "Links using alias", y = element_blank(), 
       size = "Links",
       title = "Links with alias per help page",
       subtitle = "Packages sorted by number of help pages") +
  theme_minimal() +
  theme(plot.title.position = "plot")
```


We can further explore to where do they link if we split the inside of the square brackets into the package and topic.

```{r}
#| label: base-split-dest
xrefBase <- cbind(xrefBase, 
                   strcapture("([[:alnum:].]*{2,})?[:=]?(.*)", 
                              x = xrefBase$Anchor, 
                              proto = data.frame(to_pkg = character(), 
                                                 to_target = character())))
```

We need to fill those that do not have anchors to which targets do they link to.


```{r}
#| label: base-fill-xref
fill_xref <- function(z, alias, duplicate_alias) {

  # Anchors with packages
  missing_target <- nzchar(z$Anchor) & !startsWith(z$Anchor, "=") & !grepl(":", z$Anchor, fixed = TRUE)
  z$to_target[missing_target] <- z$Target[missing_target]
  
  # Anchors with = to base packages with no duplicates
  anchors_nodup <- startsWith(z$Anchor, "=") & !z$Anchor %in% duplicate_alias
  match_target <- match(z$to_target[anchors_nodup], alias$Target)
  z$to_pkg[anchors_nodup] <- alias$Package[match_target]
  
  # Anchors with = to packages with duplicates
  anchors_dup <- startsWith(z$Anchor, "=") & z$Anchor %in% duplicate_alias
  z$to_pkg[anchors_dup] <- z$Package[anchors_dup]
  
  # No anchors
  no_anchor <- !nzchar(z$Anchor)
  match_target <- match(z$Target[no_anchor], alias$Target)
  z$to_pkg[no_anchor] <- alias$Package[match_target]
  z$to_target[no_anchor] <- alias$Target[match_target]
  z
}

xrefBase2 <- fill_xref(xrefBase, aliasesBase, dtBase$Alias)

# We can then add the aliases to know to which help page do they link.
xrefBase3 <- merge(xrefBase2, aliasesBase,
            by.x = c("to_pkg", "to_target"), 
            by.y = c("Package", "Target"), 
            all.x = TRUE, sort = FALSE) |> 
  select(Rd_origin = Source.x, Anchor, Target, Package, to_pkg, to_target, Rd_destiny = Source.y)
# As expected, xref to other packages not in base R are not included. 

# Links to packages outside the original package but no anchor provided (taking advantage that this is Base R)
no_match_Base <- xrefBase3 |> 
  filter(Anchor == "") |> 
  count(no_match = Package != to_pkg) 

xrefBase3 |> 
  filter(!to_pkg %in% names(aliases_base)) |> 
  View()
```

```{r}
#| label: base-missing-links
missing_links_base <- xrefBase3 |> 
  filter(is.na(Rd_destiny), to_pkg %in% names(aliases_base)) |> 
  select(Rd_origin, Package, Anchor)
missing_links_base
```

As way to validate we can check that base R packages do not have missing links.
Surprisingly there are `r nrow(missing_links_base)` links that do not exists: `help(package = "tools", topic = "print.via.format")` and `help(package = "methods", topic = "RMethodUtils")`.
The first one seems a typo as topic `.print.via.format` does resolve while the second one it seems like a file name was used instead of a topic from the RMethodUtils.Rd file (or simply using `isRematched` as topic). 
We'll return to this later. 


```{r}
#| label: self-xref
self_xref <- xrefBase3 |> 
  filter(Rd_origin == Rd_destiny, Package == to_pkg) |> 
  select(Rd_origin, Target, Package)
```

Looking for other weird links we can check for pages that link to themselves.
The cross-reference system doesn't provide a way to link to a specific section  (yet?). 

When users follow those links they are redirected to the top of the same help page,
while probably they were expecting to go to another help page.
At least this is something I have felt in the past. 

There are currently `r nrow(self_xref)` help files that link to themselves!
In my opinion these links could be deleted. 

There are the help files with more references to other documentation. 

```{r}
#| label: plot-base-package-links
# Explore the information
cross_ref_base <- summarise(xrefBase3, 
          .by = c(Package, to_pkg), 
          n = n(), 
          ndocs = n_distinct(Rd_origin), 
          nlinks = n_distinct(to_pkg, to_target), 
          ndocs_d = n_distinct(to_target))

cross_ref_base |> 
  filter(!is.na(to_pkg),
         to_pkg %in% names(aliases_base)) |> 
  mutate(Package = fct_reorder2(Package, n, ndocs, .fun = sum),
         Package = factor(Package, levels = rev(levels(Package)))) |> 
  ggplot() +
  geom_point(aes(fct_relevel(to_pkg, rev(levels(Package))),
                 Package, 
                 col = n, size = ndocs)) +
  labs(x = "To", y = "From", col = "Links", size = "Help pages",
       title = "Links between R help pages:") +
  theme_minimal() +
  scale_x_discrete(position = "top", guide = guide_axis(n.dodge = 2))  +
  theme(plot.title.position = "plot", 
        axis.title.y = element_text(hjust = 0, vjust = 1, angle = 0),
        axis.title.x = element_text(hjust = 0),
        )
```

We can clearly see that the packages link more between their own package but there are some heavily connected packages like utils and base or stats and base. 

```{r}
#| label: plot-base-links-files
rd_origin <- summarise(xrefBase3, 
          .by = c(Rd_origin, Package, to_pkg, Rd_destiny), 
          n = n()) |> 
  mutate(from = paste0(Package, ":", Rd_origin),
         to = paste0(to_pkg, ":", Rd_destiny)) |> 
  mutate(.by = to, n_to = n_distinct(from)) |> 
  mutate(.by = from, n_from = n_distinct(to)) |> 
  filter(to_pkg %in% names(aliases_base))
library("tidygraph")
library("ggraph")
rd_origin |> 
  select(-Rd_origin, -to_pkg, -Rd_destiny) |> 
  as_tbl_graph() |> 
  ggraph() +
  geom_edge_link(aes(col = n)) + 
  geom_node_point() +
  theme_graph() +
  labs(col = "Links",
       title = "Links between R help pages")
```

I am not plotting this because it is nice, but to show that apparently there are some help page on base R that don't have links to navigate to the rest of the documentation.

### To CRAN

We can also ask which packages are linked from R that are not part of R?

```{r}
#| label: cran-database
db <- tools::CRAN_package_db()

xrefBase3 |> 
  filter(!to_pkg %in% names(aliases_base)) |> 
  count(to_pkg, sort = TRUE, name = "links") |> 
  left_join(db[, c("Package", "Priority", "Maintainer")],
            by = join_by(to_pkg == Package),
            multiple = "first") |> 
  mutate(Maintainer = gsub(" <.+", "", Maintainer),
         `Core member` = Maintainer %in% c("Brian Ripley", "Martin Maechler", "Deepayan Sarkar", "R Core Team", "Kurt Hornik")) |> 
  arrange(-links, Priority, -`Core member`, to_pkg) |> 
  rename(Package = to_pkg)
```

We can see that some packages linked from base R are from the recommended packages and many from the R core members.
Some other links include popular packages. 

# CRAN packages

Now that we have the tools ready we can explore all the CRAN packages.

## Aliases

As before we start exploring CRAN aliases:

```{r}
#| label: cran-aliases
aliases_CRAN <- tools:::CRAN_aliases_db()
aliasesCRAN <- alias2df(aliases_CRAN)

dup_cran <- aliasesCRAN |> 
  summarise(.by = Target, n = n_distinct(Package), 
            Package = paste(Package, collapse = ", ")) |> 
  arrange(-n) |> 
  filter(n > 1)

aliasesCRAN |> 
  summarise(.by = Package, n = n_distinct(Source)) |> 
  arrange(-n) |> 
  head(10)
```

We can also find those targets that are duplicated on CRAN:

```{r}
#| label: plot-cran-dup-targets
dup_cran |> 
  head(25) |> 
  ggplot() +
  geom_col(aes(n, fct_reorder(Target, n))) +
  scale_x_continuous(expand = expansion(mult = c(0, NA), add = c(0, NA)), 
                     limits = c(0, NA)) +
  labs(x = "Packages", y = element_blank(),
       title = "Top 50 topics more repeated on CRAN") +
  theme_minimal() +
  theme(plot.title.position = "plot")
```

The most repeated topic is the magrittr pipe `%>%` followed by reexports (curious as usually the pipe is also re-exported).
We can see some common methods like plot, print, show and some that are related to the dplyr and shiny, mixed between some methods used for statistics. 

```{r}
#| label: plot-dup-targets
# library("tidygraph")
# library("ggraph")
# ab <- aliasesCRAN |> 
#   filter(.by = Target, n_distinct(Package) > 1) |> 
#   arrange(Target) |> 
#   select(Target, Package) |> 
#   mutate(f = 1) |> 
#   tidyr::pivot_wider(id_cols = Target, names_from = Package, values_from = f, values_fill = 0)
  
  # as_tbl_graph() |> 
  # mutate(Popularity = centrality_degree(mode = 'in')) |> 
  # ggraph(layout = "kk") +
  # geom_edge_fan(aes(alpha = after_stat(index)), show.legend = FALSE) + 
  # geom_node_point(aes(size = Popularity)) + 
  # theme_graph(fg_text_colour = 'white')
```

## Cross references

There are lots of packages and lots of cross-references working with them is slow.

Similar to what we did with R cross-references we first start finding links to base packages.
Then we search for links across CRAN packages.

```{r}
#| label: cran-split-dest
rdxrefs_CRAN <- tools:::CRAN_rdxrefs_db()
rdxrefsCRAN <- xrefs2df(rdxrefs_CRAN)

rdxrefsCRAN <- cbind(rdxrefsCRAN, 
                     strcapture("([[:alnum:].]*{2,})?[:=]?(.*)", 
                                x = rdxrefsCRAN$Anchor, 
                                proto = data.frame(to_pkg = character(), 
                                                   to_target = character())))

# Fill with base references
rdxrefsCRAN <- fill_xref(rdxrefsCRAN, aliasesBase, dtBase$Alias)

# Fill with CRAN targets
fill_xref_cran <- function(z, alias = aliasesCRAN, duplicates = dup_cran[["Target"]]) {
  # No anchors assuming they are from CRAN
  no_anchor <- !nzchar(z$Anchor)
  no_anchor_no_dup <- no_anchor & is.na(z[["to_pkg"]]) & !z[["Target"]] %in% duplicates
  match_target <- match(z[["Target"]][no_anchor_no_dup], alias[["Target"]])
  
  # Adding values
  z[["to_pkg"]][no_anchor_no_dup] <- alias[["Package"]][match_target]
  z[["to_target"]][no_anchor_no_dup] <- alias[["Target"]][match_target]
  
  # Adding the package of those that have anchors 
  missing_but_anchor <- is.na(z[["to_pkg"]]) & startsWith(z[["Anchor"]], "=")
  z[["to_pkg"]][missing_but_anchor] <- z[["Package"]][missing_but_anchor]
  
  # Adding duplicated Targets/topics at other packages but present on the package
  blank_target <- is.na(z$to_target)
  for (pkg in unique(z[["Package"]][blank_target])) {
    z_keep <- z[["Package"]] == pkg & blank_target
    alias_keep <- alias[["Package"]] == pkg
    m <- match(z[["Target"]][z_keep], alias[["Target"]][alias_keep])
    z[["to_target"]][z_keep] <- alias[["Target"]][alias_keep][m]
    z[["to_pkg"]][z_keep] <- alias[["Package"]][alias_keep][m]
  }
  z
}

rdxrefsCRAN2 <- fill_xref_cran(rdxrefsCRAN)
rdxrefsCRAN3 <- merge(rdxrefsCRAN2, rbind(aliasesBase, aliasesCRAN), 
            by.x = c("to_pkg", "to_target"), 
            by.y = c("Package", "Target"), 
            all.x = TRUE, sort = FALSE) |> 
  select(Rd_origin = Source.x, Anchor, Target, Package, to_pkg, 
         to_target, Rd_destiny = Source.y)
```

After that we merge the data with the files they are at.

```{r}
#| label: base-links-cran
missing <- xrefBase3 |> 
  filter(is.na(Rd_destiny)) |> 
  select(-Rd_destiny, Source = Rd_origin)
xrefBase3_m <- merge(missing, aliasesCRAN, 
            by.x = c("to_pkg", "to_target"), 
            by.y = c("Package", "Target"), 
            all.x = TRUE, sort = FALSE) |> 
  select(Rd_origin = Source.x, Anchor, Target, Package, to_pkg, to_target, Rd_destiny = Source.y)

xrefBase4 <- xrefBase3 |> 
  filter(!is.na(Rd_destiny)) |>
  rbind(xrefBase3_m) |> 
  arrange(Package, Rd_origin)
```

Finally, as I realized checking CRAN links[^4] we add links that use the file names instead of topics[^5].

[^4]: I revisited some code after finding issues further down this post.

[^5]: Yes, even though the documentation makes it clear that this shouldn't be done the help system works around this issue. I raised this issue on the [mailing list](https://stat.ethz.ch/pipermail/r-devel/2024-September/083626.html)



```{r}
#| label: base-links-files
# TODO: fix why we lose those that are matched by file probably related to the b column added to avoid losing some packages/rows.
missingBase <- xrefBase4 |> 
  mutate(n = seq_len(n())) |> 
  filter(is.na(Rd_destiny) & !is.na(to_pkg) & !is.na(to_target)) |> 
  select(-Rd_destiny, Source = Rd_origin)

aliasesCRAN$filename <- tools::file_path_sans_ext(aliasesCRAN$Source)
aliasesBase$filename <- tools::file_path_sans_ext(aliasesBase$Source)
all_aliases <- rbind(aliasesCRAN, aliasesBase)
xrefBase4_m <- merge(missingBase, all_aliases[, -2], 
            by.x = c("to_pkg", "to_target"), 
            by.y = c("Package", "filename"), 
            all.x = TRUE, sort = FALSE, all.y = FALSE) |> 
  select(Rd_origin = Source.x, Anchor, Target, Package, to_pkg, to_target,
         Rd_destiny = Source.y, n) |> 
  distinct(.keep_all = TRUE) |> 
  mutate(Rd_destiny = if_else(!is.na(Rd_destiny), paste0("=", Rd_destiny), FALSE))

xrefBase5 <- xrefBase4 |> 
  mutate(n = seq_len(n())) |> 
  filter(!n %in% xrefBase4_m$n) |>
  rbind(xrefBase4_m) |> 
  mutate(to_file = startsWith(Rd_destiny, "="),
         Rd_destiny = gsub("^=", "", Rd_destiny)) |> 
  arrange(Package, Rd_origin)

stopifnot(sum(startsWith(xrefBase4_m$Rd_destiny, "="), na.rm = TRUE) == sum(xrefBase5$to_file))
stopifnot(nrow(xrefBase4) == nrow(xrefBase5))
```



```{r}
#| label: cran-links-files
# Try to merge those that we don't know the destiny but we know the other things
missingCRAN <- rdxrefsCRAN3 |> 
  mutate(n = seq_len(n())) |> 
  filter(is.na(Rd_destiny) & !is.na(to_pkg) & !is.na(to_target)) |> 
  select(-Rd_destiny, Source = Rd_origin) 

rdxrefsCRAN3_m <- merge(missingCRAN, all_aliases[, -2], 
            by.x = c("to_pkg", "to_target"), 
            by.y = c("Package", "filename"), 
            all.x = TRUE, sort = FALSE) |> 
  distinct(.keep_all = TRUE) |> 
  select(Rd_origin = Source.x, Anchor, Target, Package, to_pkg, to_target,
         Rd_destiny = Source.y, n) |> 
  mutate(Rd_destiny = if_else(!is.na(Rd_destiny), paste0("=", Rd_destiny), NA))

rdxrefsCRAN4 <- rdxrefsCRAN3 |> 
  mutate(n = seq_len(n())) |> 
  filter(!n %in% rdxrefsCRAN3_m$n) |>
  rbind(rdxrefsCRAN3_m) |> 
  mutate(to_file = if_else(startsWith(Rd_destiny, "="), TRUE, FALSE),
         Rd_destiny = gsub("^=", "", Rd_destiny)) |> 
  arrange(Package, Rd_origin) 
  filter(!is.na(to_file)) |> 
  select(-n)
stopifnot(sum(startsWith(rdxrefsCRAN3_m$Rd_destiny, "="), na.rm = TRUE) == sum(rdxrefsCRAN4$to_file))
stopifnot(nrow(rdxrefsCRAN3) == nrow(rdxrefsCRAN4))
```

Now that I have corrected many issues I'll check if it matches CRAN's output.

## Checking them

We can compare the data with CRAN's package details of issues found related to cross-references .

```{r}
#| label: cran-details
ccd <- tools::CRAN_check_details()
crans_issues <- ccd |> 
  as.data.frame() |> 
  filter(Check == "Rd cross-references") |> 
  distinct(Package) |> 
  mutate(CRAN_checks = "rd_cross_reference")
```

There are `r nrow(crans_issues)` packages with issues in "Rd cross-references" across all flavors.
These can be found online too, for example for [r-devel in debian](https://cran.r-project.org/web/checks/check_details_r-devel-linux-x86_64-debian-clang.html), (pick your [flavor](https://cran.r-project.org/web/checks/check_summary.html)). 

To compare with CRAN we need to prepare the data collected. 
First thing, CRAN packages can depend on packages in Bioconductor and other additional repositories for which we don't have the data. 
Identifying those packages will help to avoid any false positive. 

```{r}
#| label: cran-summaries-details
# Collect Bioconductor packages
ap_bioc <- available.packages(repos = BiocManager::repositories()[1:5],
                              filters = character())
bioc_packages <- rownames(ap_bioc)

# "Merge" the two repositories (Keeping those windows specific)
db$Repository <- "CRAN"
db$File <- NA_character_
ap_all <- rbind(db[, colnames(ap_bioc)], ap_bioc)

tpd <- tools::package_dependencies(db$Package, 
                            db = ap_all, 
                            which = c("Depends", "Imports", "Suggests", "Enhances"))

in_tpd <- function(pkg) {
  tpd[[unique(pkg)]]
}

additional_repo <- function(x) {
  any(!is.na(db$Additional_repositories[db$Package == x]))
}

summary_CRAN <- rdxrefsCRAN4 |> 
  summarise(.by = Package, 
            to_file = any(to_file),
            additional_repo = additional_repo(unique(Package)),
            to_file_w_eq_anchor = any(to_file & startsWith(Anchor, "=")),
            missing_destination = any(is.na(Rd_destiny)),
            missing_anchors = any(!nzchar(Anchor)),
            linking2base = any(!to_pkg %in% names(aliases_base) & !is.na(to_pkg)),
            linking2cran = any(!is.na(to_pkg) & to_pkg != Package),
            linking2bioc = any(to_pkg %in% bioc_packages),
            missing_dependency = any(!is.na(to_pkg) & !to_pkg %in% c(Package, in_tpd(Package)) & !to_pkg %in% names(aliases_base))
  ) |> 
  mutate(to_file = if_else(is.na(to_file), FALSE, to_file))
```

To further check what is happening we can compare directly with what CRAN reports.


```{r}
#| label: cran-comparison
summary_CRAN2 <- merge(crans_issues, summary_CRAN, all = TRUE) |> 
  mutate(to_file = if_else(is.na(to_file), FALSE, to_file))
csc <- count(summary_CRAN2, CRAN_checks, to_file)
pkgs <- summary_CRAN2 |> 
  filter(is.na(missing_anchors)) |> 
  pull(Package)
setdiff(pkgs, names(rdxrefs_CRAN))
rdxrefsCRAN4 |> 
  filter(Package %in% pkgs) |> 
  View("problematic")
n_distinct(rdxrefsCRAN4$Package) == n_distinct(rdxrefsCRAN3$Package)

```

There are some few packages (`r csc$n[is.na(csc$missing_anchors)]`) that I don't detect as targets missing package anchors but CRAN does.


```{r}
#| label: cran-checks2
summary_CRAN2 |> 
  filter(is.na(CRAN_checks)) |> 
  count(missing_anchors, 
linking2base, linking2cran, linking2bioc, missing_dependency, sort = TRUE) |> 
  arrange(n)

m |> count(CRAN_checks, to_file_w_eq_anchor, missing_destination, missing_dependency, missing_anchors)

# TODO find out why those packages do not match the data
# MetaNLP has a link to glmnet that is detected as problematic for CRAN in select_features.Rd but it links correctly to glmnet and glmnet has a target for glmnet.
# # TODODO
pkgs <- m |> 
  filter(!is.na(CRAN_checks)) |> 
  filter(!to_file_w_eq_anchor, !missing_destination, !missing_dependency, !missing_anchors) |>
  pull(Package)

m |> 
  filter(is.na(CRAN_checks)) |> 
  filter(missing_destiny, !missing_anchors, !not_base, missing_other, missing_dependency, !to_bioc, !missing_package)


sdm <- setdiff(mis2$Package, mis$Package[mis$missing_anchors])
sdm2 <- setdiff(mis$Package[mis$missing_anchors], mis2$Package)

  count()
```


# Analysis

```{r}
#| label: missing-anchors
missing_anchors <- rdxrefsCRAN3 |> 
  summarise(.by = Package, missing = sum(is.na(to_target)), n = n()) |> 
  arrange(-missing)

missing_anchors |> 
  mutate(rate = missing/n) |> 
  left_join(count(aliasesCRAN, Package, name = "Help pages")) |> 
  ggplot() +
  geom_point(aes(`Help pages`, n, size = rate))
```


```{r}
#| label: source-alias
sources_alias <- aliasesCRAN |> 
  summarise(.by = packages, 
            n_sources = n_distinct(Source), 
            n_alias = n_distinct(Target)) |> 
  arrange(-n_sources, -n_alias) 

sources_alias |> 
  count(n_sources, n_alias) |> 
  ggplot() + 
  geom_point(aes(n_sources, n_alias, size = n))
```


<details>
<summary>Missing anchors in your package but not a note from CRAN?</summary>

While checking this I realized packages I maintain don't have anchors and I explored why.

The reasons is I use [roxygen2](https://cran.r-project.org/package=roxygen2) for documentation, so for all links and cross-references. 
However, roxygen2 cross-references in `@family tag` do not follow the same pattern as other cross-references. See [the issue](https://github.com/r-lib/roxygen2/issues/1650) for more information.
</details>


```{r eval=FALSE}
#| label: kurt-hornik-code
rdxrefs <- tools:::CRAN_rdxrefs_db()
## Note that as of 2024-06-11, 7139 CRAN packages have no Rd xrefs:
(n1 <- sum(lengths(rdxrefs) == 0L))

xdf <- tools:::.Rd_rdxrefs_db_to_data_frame(c(tools:::base_rdxrefs_db(),
                                              rdxrefs))

## The ones with package anchors:
ind <- with(xdf, nzchar(Anchor) & !startsWith(Anchor, "="))

targets_with_no_package_anchor <- function(x) {
    target <- x$Target
    anchor <- x$Anchor
    source <- x$Source
    names(target) <- source
    names(anchor) <- source
    c(target[!nzchar(anchor)],
      substring(anchor[startsWith(anchor, "=")], 2L))
}

## The ones without package anchors, split according to package:
rdxrefs <- tapply(xdf[!ind, ],
                  sub("::.*", "", xdf$Source[!ind]),
                  targets_with_no_package_anchor,
                  simplify = FALSE)

## As of 2024-06-11, 12304 CRAN packages have Rd xrefs without package
## anchors:
(n2 <- length(rdxrefs))

aliases_in_base <- tools:::base_aliases_db()
aliases_in_CRAN <- tools:::CRAN_aliases_db()

aliases_for_level_0 <-
    lapply(c(aliases_in_base, aliases_in_CRAN),
           unlist, use.names = FALSE)
aliases_for_level_1 <-
    unique(unlist(aliases_in_base, use.names = FALSE))

targets_needing_anchors <-
    Map(function(x, p) {
        x[is.na(match(x,
                      c(aliases_for_level_0[[p]],
                        aliases_for_level_1)))]
    },
    rdxrefs,
    names(rdxrefs))

## As of 2024-06-11, 9860 CRAN packages have Rd xrefs without package
## anchors but do not need any (all non-package-anchored targets are to
## level 0 and new-level-1 base-package only aliases).
(n3 <- sum(lengths(targets_needing_anchors) == 0L))

## This leaves 2444 CRAN packages which need to add package anchors to
## their Rd xrefs:
(n4 <- length(rdxrefs) - sum(lengths(targets_needing_anchors) == 0L))
## Numbers of Rd xrefs needing package anchors:
(tab <- table(lengths(targets_needing_anchors)))
```


# Bioconductor packages

One of the problems of providing HTML manual pages is providing links to the Bioconductor manual pages.
While I have been pressing for having the same file available at the time of writing it is not yet possible to check on them. 

As this is is a recent introduction to R repositories and not compulsory, or mandated by the R-admin section about [setting up repositories](https://cran.r-project.org/doc/manuals/r-release/R-admin.html#Setting-up-a-package-repository) it is not provided by Bioconductor or R-universe or other systems to setup repositories. 

The code in CRAN goes beyond the definition of repository set and Bioconductor hasn't provided this detail of information (yet?).

# Conclusions

### Reproducibility

<details>
<summary>Session info</summary>
```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info(info = c("platform", "packages"))
```
</details>
