---
title: Forest fires in Mexico
author: Lluís Revilla Sancho
date: '2019-05-24'
slug: fires-mexico
categories:
  - r
tags:
  - r
  - tidyverse
  - maps
  - gganimate
image:
  caption: ''
  focal_point: ''
editor_options: 
  chunk_output_type: console
---

## Evolucion de los incendios forestales en México

I saw a map on twitter and I wanted to see if I could change a bit and improve it:

```{r tweet, echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
blogdown::shortcode('tweet', '1129429109212213248')
```

I wanted to make so that the fires change the size with time to appear and disappear instead of moving around Mexico. 


I got the [code from Andrian](https://github.com/bautista-adrian/Incendios_Mx/blob/master/Incendios.R), I ran the code and stored it:
```{r setup, collapse=TRUE}
library("here")
library("ggplot2")
library("gganimate")
library("ggthemes")
# Maps from mexico
# devtools::install_github('diegovalle/mxmaps')
library("mxmaps")
plottable <- readRDS(here("static", "fires-mexico", "fires_mexico.RDS"))
head(plottable)
```

## Plot

My main objective was to remove the points moving above the country from year to year.
The problem is that the aesthetic was grouped ad the very definition by group.
```{r map}
data(mxstate.map)

m <- ggplot() + 
  geom_polygon(data = mxstate.map, aes(x = long, y = lat, group = group), 
               fill = "grey", color = "black") +
  geom_point(data = plottable, 
             aes(x = Longitud, y = Latitud, group = Año), alpha = .2, color = "red") +
  theme(plot.subtitle = element_text(size = 32, color = "grey")) +
  scale_size_discrete()  +
  theme_map()  +
  coord_map()
m
```

## Animation

Now that we have the plot and the fires we can animate it by each year:

```{r animation}
gganimation <- m +
  transition_states(Año, transition_length = 11, state_length = 1, wrap = FALSE) +
    enter_fade() + 
    exit_fade() + 
    labs(title = "Incendios forestales registrados en México",
         subtitle = "Año {closest_state}", 
         x = element_blank(),
         y = element_blank(), 
         caption = "Source: CONAFOR / Elaboracion: @4drian.bautista & @Lluís_Rev")

animate(gganimation, fps = 25, duration = 10)
```
