---
title: "Advanced use of BaseSet"
author: 'Lluís Revilla Sancho @Lluis_Revilla'
subtitle: "Exploring Bioconductor packages"
date: "`r Sys.Date()`"
slug: BaseSet-advanced
tags: [r]
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightLines: true
      countIncrementalSlides: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
library("tidyverse")
library("BaseSet")
```

## BaseSet

A [package](https://github.com/llrs/BaseSet) to work with sets. 

```{r show, eval=FALSE, include=TRUE}
BiocManager::install("llrs/BaseSet")
```

Some features:

- Works with the tidyverse verbs
- [Interacts easily](https://llrs.github.io/BaseSet_slides/) with existing *[GSEABase](https://bioconductor.org/packages/GSEABase)* classes.
- Enables to use probabilities in the association between elements and sets.

---

## Bioconductor packages

We can quickly access the dependencies and maintainers of Bioconductor' packages:

```{r download}
views <- "https://bioconductor.org/packages/devel/bioc/VIEWS"
dcf <- read.dcf(url(views), c("Package", "Maintainer", 
                  "dependencyCount", "Depends", "Imports"))
tbl <- as_tibble(dcf) %>%
   mutate(
     Maintainer = gsub("[[:space:]]+", " ", Maintainer),
     dependencyCount = as.integer(dependencyCount),
     Core = grepl("Bioconductor Package Maintainer", Maintainer),
     Maintainer = stringr::str_split(Maintainer, ",\\s*|( and )"),
     Depend = stringr::str_split(Depends, ",\\s+"), 
     Import = stringr::str_split(Imports, ",\\s+"),
     )
```

Modified code from Martin Morgan

---

The content:

```{r tbl, echo=FALSE, message=TRUE, warning=FALSE}
head(tbl)
```

As you can see the Maintainers field is a list and also the Depends and Import.

---

## Maintainers

For each package we can have several maintainers. We must indicate what are our sets and our elements:
```{r maintainers}
prepare <- unnest(tbl[, c("Maintainer", "Package")])
colnames(prepare) <- c("sets", "elements")
maintainers <- tidySet(prepare)
```

We can see with `nElements(maintainers)` that it has less than `r nElements(maintainers)` maintainers. Because some appear duplicated:

```{r carey}
grep("Carey", name_elements(maintainers), value = TRUE)
```

---

Now we are ready for some plots:

```{r maintainers1, message=FALSE, warning=FALSE, fig.height=4, dev='svg'}
maintainers %>% 
  set_size() %>% 
  ggplot() +
  geom_histogram(aes(size)) + 
  scale_y_log10() +
  theme_minimal() +
  labs(title = "Maintainers by package")
```

---

```{r maintainers2, message=FALSE, warning=FALSE, fig.height=4, dev='svg'}
maintainers %>% 
  filter_element(!grepl("Bioconductor Package Maintainer", 
                        as.character(elements))) %>% 
  element_size() %>% 
  ggplot() +
  geom_histogram(aes(size)) + 
  scale_y_log10() +
  theme_minimal() +
  labs(title = "Packages maintained by person") 
```

---

## Bioconductor packages

We can have a look at the packages maintained by Bioconductor:

```{r Bioc_packages}
(Bioconductor_packages <- maintainers %>% 
  filter_element(grepl("Bioconductor", elements)) %>% 
  droplevels() %>% 
  pull_set(sets))
```

```{r rcore, include=FALSE}
rcore <- c("base", "compiler", "datasets", "grDevices", "graphics", "grid", 
           "methods", "parallel", "splines", "stats", "stats4", "tcltk", "tools", 
           "utils", "KernSmooth", "Mass", "Matrix", "boot", "class", "cluster", 
           "codetools", "foreign", "lattice", "mgcv", "nlme", "nnet", "rpart", 
           "spatial", "survival")
```

And we can see which are installed by default or recommended at the [FAQs](https://cran.r-project.org/doc/FAQ/R-FAQ.html#Which-add_002don-packages-exist-for-R_003f):
in total `r length(rcore)`.

---

## Dependencies

Now let's work on the dependencies.
First some clean up of the package names:

```{r dependencies}
l <- apply(tbl[, c("Depend", "Import")], 1, function(x){
  y <- unlist(x, use.names = FALSE)
  y <- gsub(" ?\\(.*", "", y)
  y <- y[!grepl("\\)", y)]
  y <- y[y != ""]
  unique(y[!is.na(y)])
})
names(l) <- tbl[, "Package", drop = TRUE]
head(l, 2)
```

---

Now we are ready to transform to a TidySet:

```{r dependencies0, message=FALSE, warning=FALSE}
packages <- tidySet(l) %>% 
  filter_element(elements != "R") %>% 
  mutate_element(Origin = case_when(
    elements  %in%  rcore ~ "R core",
    elements  %in% Bioconductor_packages ~ "Bioconductor Core",
    elements  %in% name_sets(maintainers) ~  "Bioconductor",
    TRUE ~ "Other")) %>% 
    activate("sets") %>% 
    mutate(Type = case_when(
    sets  %in% rcore ~ "R core",
    sets  %in% Bioconductor_packages ~ "Bioconductor Core",
    sets  %in% name_sets(maintainers) ~  "Bioconductor",
    TRUE ~ "Other"))

sets(packages) <- inner_join(sets(packages),
                             tbl[, c("Package", "dependencyCount")], 
                             by = c("sets" = "Package"))
is_nested(packages)
```

In this case we can see that some packages appear as dependencies of others as expected.

---

We can plot the direct dependencies:

```{r dependensies1, message=FALSE, warning=FALSE, fig.height=4, dev='svg'}
packages %>% 
  set_size() %>% 
  ggplot() +
  geom_histogram(aes(size), bins = 50) +
  scale_y_log10() +
  theme_minimal() +
  labs(title = "Packages dependencies")
```

---

```{r dependensies2, message=FALSE, warning=FALSE, fig.height=4, dev='svg'}
p <- packages %>% 
  element_size() %>% 
  left_join(elements(packages))
p %>% 
  ggplot() +
  geom_histogram(aes(size)) + 
  scale_y_log10() +
  theme_minimal() +
  labs(title = "Packages depended")
```


```{r dependensies4, message=FALSE, warning=FALSE, fig.height=4, dev='svg'}
p %>% 
  ggplot(aes(Origin, size)) +
  geom_violin() +
  geom_point(position = position_jitter()) +
  theme_minimal() +
  scale_y_log10()
table(p$Origin)
```
 
---

## Fuzzy sets

The relationship between the element and the set is a probability: it can be or it can't be.

For example from [https://docs.abcam.com/pdf/immunology/t_cells_the_usual_subsets.pdf](t_cells_the_usual_subsets.pdf)

```{r fuzzy1}

signature <- list(Th1 = c("CD3" = 1, "CD4" = 1, "IL-12R" = 0.75, "CXCR3" = 0.2),
                  Th2 = c("CD3" = 1, "CD4" = 1, "IL-4R" = 1, "CCR4" = 0.8, 
                          "CRTH2" = 0.2),
                  Th9 = c("CD3" = 1, "CD4" = 1, "IL-9" = 0.2, "IL-10" = 0.4),
                  Th17 = c("CD3" = 1, "CD4" = 1, "IL-23R" = 0.9, "CCR6" = 0.2, 
                           "IL-1R" = 0.8, "CD161" = 1),
                  Tm = c("CCR7" = 0.5, "CD44" = 1, "CD62L" = 0.2, "CD3" = 0.57, 
                         "IL-7R" = 0.89, "CD127" = 0.32, "IL-15R" = 0.83),
                  NKT = c("NK1.1" = 1, "SLAMF1" = 0.9, "SLAMF6" = 0.7, 
                          "PLZF" = 0.4)
                  )

cells <- tidySet(signature)
```

---

We can check that the most probable 

```{r }
set_size(cells)
```

---

## Thanks!!

Thanks for all the feedback from:
 
 - Kayla Morall
 - Kevin Rue
 - Martin Morgan
 - Ludwig Geistlinger

And for your feedback!
