---
title: 'New rtweet release: 2.0.0'
author: Lluís Revilla Sancho
date: '2024-02-16'
slug: new-rtweet-release-2-0-0
categories:
  - r
  - CRAN
tags:
  - rtweet
authors:
  - admin
description: ''
draft: no
editor_options:
  chunk_output_type: console
featured: no
image:
  caption: ''
  focal_point: ''
subtitle: ''
summary: ''
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.retina = 2)
```


This is a brief announcement of rtweet version 2.0.0.
This major version changes signals the move from the API v1.1 to the API v2.

There haven't been many changes since 1.2.1 but this is to signal that the API v1.1 is deprecated.

The previous release was a bit of a rush to meet with the requirements of CRAN maintainers to fix an error and it wasn't polished. 
Some users complained that it was difficult to find what worked.
In this release I focused mostly to make life easier for users:

Now there is a document the deprecated functions from API v1.1 to API v2: see `help("rtweet-deprecated", "rtweet")`.
I also made it easier for the rtweet to work with API v2: the release of httr2 1.0.0 version helped to avoid some workarounds with the authentication process.

I also focused on updating the vignettes to the most up to date recommendations.
I am not sure the streaming vignettes is up to date (but keep reading why I left it as is).

Last, following CRAN policy: if users create rtweet data they can now delete it with  `client_clean()` and `auth_clean()`.

# Future releases

For the last year I [asked the community](https://github.com/ropensci/rtweet/issues/763) for a co-maintainer with interest in the package.
Unfortunately, people that showed some interest at the end didn't commit to it.

At the same time I [also asked](https://llrs.dev/post/2023/02/16/rtweet-future/) for  [donations](https://www.buymeacoffee.com/llrs) to support an API access.
It currently costs 100€ to access most endpoints which is needed to test and develop the package. 
However, this is more than half of what I spend in groceries last month.  
Other packages like [academictwitteR](https://cran.r-project.org/package=academictwitteR) are also stopping development/support. 
Although not archived from CRAN, it has a note in the README:

> Note this repo is now ARCHVIED due to changes to the Twitter API. The paid API means open-source development of this package is no longer feasible.

Similarly without financial help and community interest I won't invest more time on it.  
This is the last version that I release.
I have other interests and I would like to focus on other projects.
My focus will be on updating and releasing some packages I have.
I also want to focus more on my own company to help the R community (and beyond). 
I will write about the company shortly.

There have been some discussions on social media how to signal deprecation of packages.
The only method available on CRAN that I know is to declare a package ORPHANATED.
I have requested to CRAN to declared the package ORPHANATED.

### Reproducibility

<details>
```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info()
```
</details>
