---
title: Writing a thesis with bookdown
author: Lluís Revilla Sancho
date: '2022-05-08'
slug: writing-thesis-bookdown
categories:
  - news
tags:
  - R
  - bookdown
authors: []
description: ''
draft: no
editor_options:
  chunk_output_type: console
featured: no
image:
  caption: ''
  focal_point: ''
subtitle: ''
summary: ''
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.retina = 2)
```

## Starting writing

On this post I am documenting the experiences I had writing  my [PhD thesis](https://thesis.llrs.dev) with bookdown. 
I made the thesis in web and pdf format (and epub) to make more available the thesis.
Most of the experiences and advise I'll share here are based on my experiences to improve the pdf format. 
It is the most important format as ultimately is what I'm going to use for printing.  

First of all you should know there is a package [thesisdown](https://github.com/ismayc/thesisdown) with a few templates for some universities. 
If yours is there, or if you want to learn how are they you can have a look at the files.  
On my case I didn't have any template and the university guidelines are not long, (have two compulsory pages at the beginning and have 5 sections).
That's why I tweaked the default format inspired by the recent thesis defended on my group. 

Well, without further delay let's dive in things I learned 

## Things I should have known/used

ref text.

short caption

caption placement


## Nice litte tricks

Band colors

TOC, with TOC

TOF, TOT

Acronyms

Limit figures to subsection

Github actions to render them

# Conclusion

Stablished and well documented 

gitbook vs bs4 format


### Reproducibility

<details>
```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info()
```
</details>
