---
title: Upgrading rtweet to 1.0.0
author: Lluís Revilla Sancho
date: '2022-07-04'
slug: rtweet-1-0-0
categories:
  - Bioconductor
tags:
  - r
  - rtweet
  - packages
authors:
  - admin
description: ''
draft: no
editor_options:
  chunk_output_type: console
featured: no
image:
  caption: ''
  focal_point: ''
subtitle: ''
summary: ''
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.retina = 2)
```

In this post I will provide some examples of what has changed between rtweet 0.7.0 and rtweet 1.0.0. 
I hope both the changes and this guide will help all users.
I won't provide a full list of changes but you can consult it on the [NEWS](https://docs.ropensci.org/rtweet/news/index.html).

## Big breaking change

::: {#hello .rare-block}
Hello **world**!!
:::


### More consistent output

This is probably what will affect more users: All functions that return data about tweets return the same columns.

```r
> tweets <- search_tweets("weather")
> colnames(tweets)
 [1] "user_id"                 "status_id"              
 [3] "created_at"              "screen_name"            
 [5] "text"                    "source"
 [7] "display_text_width"      "reply_to_status_id"     
 [9] "reply_to_user_id"        "reply_to_screen_name"   
[11] "is_quote"                "is_retweet"             
[13] "favorite_count"          "retweet_count"          
[15] "quote_count"             "reply_count"            
[17] "hashtags"                "symbols"                
[19] "urls_url"                "urls_t.co"              
[21] "urls_expanded_url"       "media_url"              
[23] "media_t.co"              "media_expanded_url"     
[25] "media_type"              "ext_media_url"          
[27] "ext_media_t.co"          "ext_media_expanded_url" 
[29] "ext_media_type"          "mentions_user_id"       
[31] "mentions_screen_name"    "lang"                   
[33] "quoted_status_id"        "quoted_text"            
[35] "quoted_created_at"       "quoted_source"          
[37] "quoted_favorite_count"   "quoted_retweet_count"   
[39] "quoted_user_id"          "quoted_screen_name"     
[41] "quoted_name"             "quoted_followers_count" 
[43] "quoted_friends_count"    "quoted_statuses_count"  
[45] "quoted_location"         "quoted_description"     
[47] "quoted_verified"         "retweet_status_id"      
[49] "retweet_text"            "retweet_created_at"     
[51] "retweet_source"          "retweet_favorite_count" 
[53] "retweet_retweet_count"   "retweet_user_id"        
[55] "retweet_screen_name"     "retweet_name"           
[57] "retweet_followers_count" "retweet_friends_count"  
[59] "retweet_statuses_count"  "retweet_location"       
[61] "retweet_description"     "retweet_verified"       
[63] "place_url"               "place_name"             
[65] "place_full_name"         "place_type"             
[67] "country"                 "country_code"           
[69] "geo_coords"              "coords_coords"          
[71] "bbox_coords"             "status_url"             
[73] "name"                    "location"               
[75] "description"             "url"                    
[77] "protected"               "followers_count"        
[79] "friends_count"           "listed_count"           
[81] "statuses_count"          "favourites_count"       
[83] "account_created_at"      "verified"               
[85] "profile_url"             "profile_expanded_url"   
[87] "account_lang"            "profile_banner_url"     
[89] "profile_background_url"  "profile_image_url"   
```

::: {#hello2 .r  style="background: green;"}

hi

:::

```r
> tweets <- search_tweets("weather")
> colnames(tweets)
 [1] "created_at"                    "id"                           
 [3] "id_str"                        "full_text"                    
 [5] "truncated"                     "display_text_range"           
 [7] "entities"                      "metadata"                     
 [9] "source"                        "in_reply_to_status_id"        
[11] "in_reply_to_status_id_str"     "in_reply_to_user_id"          
[13] "in_reply_to_user_id_str"       "in_reply_to_screen_name"      
[15] "geo"                           "coordinates"                  
[17] "place"                         "contributors"                 
[19] "is_quote_status"               "retweet_count"                
[21] "favorite_count"                "favorited"                    
[23] "retweeted"                     "lang"                         
[25] "quoted_status_id"              "quoted_status_id_str"         
[27] "quoted_status"                 "possibly_sensitive"           
[29] "retweeted_status"              "text"                         
[31] "favorited_by"                  "scopes"                       
[33] "display_text_width"            "quoted_status_permalink"      
[35] "quote_count"                   "timestamp_ms"                 
[37] "reply_count"                   "filter_level"                 
[39] "query"                         "withheld_scope"               
[41] "withheld_copyright"            "withheld_in_countries"        
[43] "possibly_sensitive_appealable"
```

rtweet now minimizes the processing of tweets and only returns the same data as provided.
However, due to the nested nature of the data returned some fields are now nested inside other or not provided by default.
At the same time it provides with new information fields about each tweet like the `withheld_*` columns.

If you scanned through the text column fields you might have noticed that there aren't `"user_id"` or `"screen_name"` columns returned: users data is now provided different.
To access users data you need to use `users_data()` (and if you look up for a user and want to know about their tweets you can use `tweets_data()`).

```r
> colnames(users_data(tweets))
 [1] "id"                      "id_str"                 
 [3] "name"                    "screen_name"            
 [5] "location"                "description"            
 [7] "url"                     "protected"              
 [9] "followers_count"         "friends_count"          
[11] "listed_count"            "created_at"             
[13] "favourites_count"        "verified"               
[15] "statuses_count"          "profile_image_url_https"
[17] "profile_banner_url"      "default_profile"        
[19] "default_profile_image"   "withheld_in_countries"  
[21] "derived"                 "withheld_scope"         
[23] "entities" 
```

Even then there is more nested data in the entities column (which is present in both tweets and users!).

 ```r
> names(tweets$entities[[1]])
[1] "hashtags"      "symbols"       "user_mentions" "urls"         
[5] "media" 
 ```

### More consistent interface

All paginated functions that don't return tweets now use a consistent pagination interface (except the premium endpoints). 
They all store the "next cursor" in an `rtweet_cursor` attribute, which will be automatically retrieved when you use the `cursor` argument.
They all support `max_id` and `since_id` to find earlier and later tweets respectively, as well as `retryonratelimit` to wait as long as needed when rate limited

`get_friends()` and `get_followers()`

### Saving data

An unexpected consequence of returning more data as in the API is that it is now harder to save it in a tabular format.
For instance one tweet might have one media, two user mention and three hasthags.
There isn't a simple way to saved it in a single row uniformly for all tweets.

This resulted in deprecating `save_as_csv`, `read_twitter_csv` and related functions. 
They will be removed in later versions but at the moment it is not possible to use `save_as_csv` with the new data structure and it won't be possible to load it from a csv. 
There are other packages that provide similar functionality such as [{tidytags}](https://docs.ropensci.org/tidytags/) and users should decide how to use it.

## Other breaking changes

If you tweet via `post_tweet()` and add an image, gif or video you'll need to provide the media alternative text. 

`tweet_shot()` has been hard deprecated as it did no longer work correctly. 
There might be possible to bring it back, but I don't understand the code provided and can't maintain it (see the long discussion about it [here](https://github.com/ropensci/rtweet/issues/458)). 
  
emojis, langs and stopwordslangs data provided with rtweet is no longer available.
I don't think it is the right place to provide this kind of data and I don't want to be responsible to update them. 


The functions like `suggested_*()` have been removed as they didn't work since 2019.

## Easier authentication

If you just want to test the package use the default authentication `auth_setup_default()` that comes with rtweet.

If you want to use the package more than 1 day, set up your own token via `rtweet_user()`.
It will open a window to authenticate via the account you have opened in your default browser.
This authentication won't allow you to do everything but for most use cases it will work. 

If you want to make a heavier usage of the package I recommend to register yourself as developer and use one of the following two mechanisms for different use cases:

- Collect data and analyze: `rtweet_app()`.
- Set up a bot: `rtweet_bot()`

Find more information on the [Authentication with rtweet vignette](https://docs.ropensci.org/rtweet/articles/auth.html).

Previously rtweet saved each token created, if you don't use the default token you'll need to save them manually via `save_as(token, "my_app")`. 
Bonus, if you name your token as default (`save_as(token, "default")`) it will be used since loading the library.

To check which credentials you have stored use `auth_list()` and load them via `auth_as("my_app")`.
All the rtweet functions will use the latest token loaded with `auth_as` (unless you manually specify one when calling it).
If you are not sure which token you are using you can use `auth_get()` it will return the token in use, list them or ask you to authenticate.

If you have previous tokens saved or problems identifying which token is which use `auth_sitrep()`.
This will provides clues to which tokens might be duplicated or misconfigurated but it won't check if they work. 

## Future

Twitter API v2 is being released and soon it will replace API v1. 
rtweet uses API v1 so it will need to adapt to the new endpoints and new data returned.

First will be the streaming endpoints in November, so expect more breaking changes around those dates.

I would also like to provide some classes to make it easier for dependencies and the package itself to handle the outputs.
Specifically I would like to provide functions to make it easier to reply to previous tweets, 
provide shortcuts to find out some nested information and make it easier to handle.


## Conclusions

While I made many breaking changes I hope this helps both users and the maintainers.

Feel free to ask on the [rOpenSci community](https://discuss.ropensci.org/tag/rtweet) if you have questions about the transition or find something amiss. 
Probably I overlooked something, let me know and I'll try to fix it or answer.

In case you find a bug, check the open issues and if not reported previously open an [issue on github](https://github.com/ropensci/rtweet/issue/). 
Don't forget to make a [reprex](https://cran.r-project.org/web/packages/reprex/readme/README.html) and if possible provide the id of the tweets you are having trouble or when I came to look at it I might not be able to find them.

Happy tweeting!


### Reproducibility

<details>
```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info()
```
</details>
