---
title: Creating an RSS feed for r-bloggers for my blogdown
author: Lluís Revilla Sancho
date: '2020-11-15'
slug: r-blogger-blogdown
categories:
  - r
tags:
  - R
authors: []
description: ''
editor_options:
  chunk_output_type: console
featured: no
image:
  caption: ''
  focal_point: ''
subtitle: ''
summary: ''
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

I am following [Yongfu's blog](https://yongfu.name/2018/12/13/hugo_rss.html):
I copy here the content in case it is moved:

>     Copy the default RSS template

>    Change <description>{{ .Summary | html }}</description> to <description>{{ .Content | html }}</description>

>    Create subdirectories in layouts/ and save the RSS template files (name them rss.xml) in them (one for each subdirectory).



I was first confused when I wanted to copy the rss.xml file mentioned and didn't know where to go.
I finally found the [section](https://gohugo.io/templates/rss/#the-embedded-rssxml) from which I could go to the file I was supposed to copy. 

So far so good, so without further testing I assume this worked. 
Following [the instructions](https://gohugo.io/templates/rss/), I moved the rss.xml file to layouts/posts/ to only report post of the blog, not new projects or other leafs from Hugo.

Next step is filtering and further configuring the feed.
Filtering because I cannot feed R-bloggers with other languages than English, and configure because I don't want to submit/provide everything, just the latest posts.

So, a quick search reported [several](https://benjamincongdon.me/blog/2020/01/14/Tips-for-Customizing-Hugo-RSS-Feeds/) [leads](https://discourse.gohugo.io/t/how-to-generate-pages-excluding-specific-rss/4026):

So I set a limit of 20 post on the config.toml file.



### References

```{r bibsetup, echo=FALSE, message=FALSE, warning=FALSE}
## Load knitcitations with a clean bibliography
library('knitcitations')
cleanbib()
cite_options(hyperlink = 'to.doc', citation_format = 'text', style = 'html')
pi <- sessioninfo::package_info()
packages <- c(pi$package[pi$attached], 'knitcitations')
l <- sapply(packages, function(x){citation(x)[1]}, simplify = FALSE)
bib <- c(l, 'blogdown' = citation('blogdown')[2])
```

```{r results = 'asis', echo = FALSE, cache = FALSE}
bibliography(style = 'html')
```

### Reproducibility

<details>
```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info()
```
</details>
