---
title: Creating an RSS feed for r-bloggers for my blogdown
author: Lluís Revilla Sancho
date: '2020-11-15'
slug: r-blogger-blogdown
draft: true
categories:
  - r
tags:
  - R
authors: []
description: ''
editor_options:
  chunk_output_type: console
featured: no
image:
  caption: ''
  focal_point: ''
subtitle: ''
summary: ''
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

I am following [Yongfu's blog](https://yongfu.name/2018/12/13/hugo_rss.html):
I copy here the content in case it is moved:

>     Copy the default RSS template

>    Change <description>{{ .Summary | html }}</description> to <description>{{ .Content | html }}</description>

>    Create subdirectories in layouts/ and save the RSS template files (name them rss.xml) in them (one for each subdirectory).

I was first confused when I wanted to copy the rss.xml file mentioned and didn't know where to go.
I finally found the [section](https://gohugo.io/templates/rss/#the-embedded-rssxml) from which I could go to the file I was supposed to copy. 

So far so good, so without further testing I assume this worked. 
Following [the instructions](https://gohugo.io/templates/rss/), I moved the rss.xml file to `layouts/post/` to only report posts of the blog, not new projects or other leafs from Hugo.

I also created a rss.xml on `layouts/categories/`. 
As the templates provided by gohugo.io didn't work I copied and pasted the [files from Yongfu's Blog](https://github.com/liao961120/Hugo-RSS/tree/master/layouts).

To check that it worked I went to [`llr.dev/categories/cran/index.xml`](https://llr.dev/categories/cran/index.xml).
Surprisingly it worked for some categories but not for others ([r](https://llr.dev/categories/r/index.xml)), so I needed to tweak the layout.

I set a limit of 20 post on the config.toml file `rssLimit = 20`. 

### Reproducibility

<details>
```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info()
```
</details>
