---
title: NSE in base R
author: Lluís Revilla Sancho
date: '2019-07-05'
slug: NSE-base
categories:
  - r
tags:
  - r
  - tidyverse
  - base
  - NSE
image:
  caption: ''
  focal_point: ''
---

After some discussions on the internet (and some breaks I got in a pipe using [tidyr](https://cran.r-project.org/package=tidyr)) I'll decided to try to emulate [dplyr](https://cran.r-project.org/package=dplyr) and other tidyr functions with base functions.

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```


## Filter

The most important tipical function is filter, which basically does subsetting:

```{r filter1}
library("dplyr")
library("bench")
mtcars_filter <- filter(mtcars, am == "1")
mtcars_subset <- subset(mtcars, am == "1")
all.equal(mtcars_filter, mtcars_subset)
```

The first different are the rownames that are changed. 
So we can make a function to make the result equal:

```{r filter2}
filter_base <- function(x, ...) {
  y <- subset(x, ...)
  rownames(y) <- NULL
  y
}
mtcars_subset <- filter_base(mtcars, am == "1")
# From the previous subset
all.equal(mtcars_filter, mtcars_subset)
m <- mark(filter(mtcars, am == "1"), 
          filter_base(mtcars, am == "1"),
          iterations = 1000)
plot(m) + ggplot2::theme_classic()
```

```{r filter3}
mtcars_filter <- filter(mtcars, am == "1", mpg > 25)
mtcars_subset <- filter_base(mtcars, am == "1", mpg > 25)
all.equal(mtcars_filter, mtcars_subset)
```
The problem here is that we used a `,` to concatenate the different conditions.
To concatenate a logical condition it is typical to use `&`. If we go this way:
```{r filter4}
mtcars_subset <- filter_base(mtcars, am == "1" & mpg > 25)
# mtcars_filter is the same
all.equal(mtcars_filter, mtcars_subset)

m <- mark(filter(mtcars, am == "1", mpg > 25),
          filter_base(mtcars, am == "1" & mpg > 25),
          iterations = 1000)
plot(m) + ggplot2::theme_classic()
```

So we've show what is the equivalent of filter for some easy cases.

## Grouping

One of the functions I use most when using dplyr is `group_by`. 
But what is the equivalent of `group_by` on base? 
I don't know, so let's explore what does it do to try to find it's equivalent. 
Let's look into it:

```{r group_by1}
(mtcars_grouped <- group_by(mtcars, vs))
class(mtcars_grouped)
```

First thing we note is that it converts it to a tibble. Second that it has a 
`Groups: Species [3]` at the top of the printed result. To look what is this and where it comes from let's look at the structure with `str`:

```{r group_by2}
str(mtcars_grouped)
``` 
We can see that it has a new attribute named "groups":

```{r group_by3}
groups_dplyr <- attr(mtcars_grouped, "groups")
groups_dplyr$vs
groups_dplyr$.rows
```
This attribute is first a vector of the names of the group and a list of the indicies of each group.

So what is the equivalent? I don't know, but we can store the names and indicies with:

```{r group_by4}
(groups_base <- split(seq_along(mtcars$vs), mtcars$vs))
```

Now let's try to complicate it more:

```{r group_by5}
mtcars_grouped <- group_by(mtcars, vs, am)
group <- attr(mtcars_grouped, "group")
```

What would be our equivalent?

```{r group_by6}
(groups_base <- split(seq_along(mtcars$vs), as.factor(paste(mtcars$vs, mtcars$am))))
```
We create a "hash" of the two factors added to create the index of each combination.

## Summarise

The base function for summarise is `aggregate`, we'll look into it:

```{r summarise1}
mtcars_summarised <- mtcars %>% 
  group_by(vs, am) %>% 
  summarise(hp = mean(hp))

mtcars_aggregated <- aggregate(hp ~ vs + am,
                               data = mtcars, 
                               FUN = mean, simplify = FALSE)
mtcars_summarised
mtcars_aggregated
```

We can see that we get the same result but with different order and class. 
We need to order the data.frame and remove the rows in order to get them equal 
(and get back to a `data.frame` in the case of `summarise`).

```{r summarise2}
mtcars_aggregated$hp <- unlist(mtcars_aggregated$hp) # To unlist the summary
mtcars_aggregated <- mtcars_aggregated[order(mtcars_aggregated$hp, 
                                             decreasing = TRUE), ]
rownames(mtcars_aggregated) <- NULL
mtcars_aggregated
mtcars_summarised
all.equal(mtcars_aggregated, mtcars_summarised)
```

Now we create some functions:

```{r summarise3}
# In just one long line to avoid calls to %>% 
mtcars_summarise <- function(){ 
  as.data.frame(summarise(group_by(mtcars, vs, am), hp = mean(hp)))
}
mtcars_summarise_usual <- function(){ 
  mtcars %>% 
    group_by(vs, am) %>% 
    summarise(hp = mean(hp))
}

mtcars_aggregate <- function(){ 
  mtcars_aggregated <- aggregate(hp ~ vs + am,
                               data = mtcars, 
                               FUN = mean)
  mtcars_aggregated$hp <- unlist(mtcars_aggregated$hp) # To unlist the summary
  mtcars_aggregated <- mtcars_aggregated[order(mtcars_aggregated$hp, 
                                               decreasing = TRUE), ]
  rownames(mtcars_aggregated) <- NULL
  mtcars_aggregated
}

mtcars_aggregate_base <- function() {
   aggregate(hp ~ vs + am,
                               data = mtcars, 
                               FUN = mean)
}
m <- mark(mtcars_aggregate(), 
          mtcars_summarise(),
          mtcars_summarise_usual(), 
          mtcars_aggregate_base(),
          iterations = 1000, check = FALSE)
plot(m) + ggplot2::theme_classic()
```


## References

After writing this I found this [other summary](https://tavareshugo.github.io/data_carpentry_extras/base-r_tidyverse_equivalents/base-r_tidyverse_equivalents.html) which covers the same
