---
title: Christmas tree
author: Lluís Revilla Sancho
date: '2019-12-18'
slug: christmas-tree
categories:
  - r
tags:
  - r
  - christmas
  - plot
authors: []
description: ''
editor_options:
  chunk_output_type: console
featured: no
image:
  caption: ''
  focal_point: ''
subtitle: ''
summary: ''
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

So it is Christmas and I want a tree. I found this new package experimental [ggfx](https://github.com/thomasp85/ggfx) and I thought it would be nice to use it for the lights. 
Then some time later I found (through twitter) this [other repo](https://github.com/r-link/christmas_tree_challenge) about building trees.

So here I am building my own christmas tree!

# The very basics

We start with installations:

```{r install, eval=FALSE}
devtools::install_github('thomasp85/ggfx')
```

We'll build our tree with triangles:

```{r triangle}
library("ggplot2")
ggplot(data.frame(x = c(-5, 5, 0), y = c(0, 0, 5))) +
  geom_polygon(aes(x, y), fill = "green") +
  theme_void()
```

To make it easier we use some functions for the triangles:

```{r tree_function}
iso_triangle <- function(wide, height, xvertix, yvertix) {
  y <- c(rep(yvertix-height, 2), yvertix)
  x <- c(xvertix+wide/2, xvertix-wide/2, xvertix)
  data.frame(x, y)
}

square <- function(length, x_bottom_right, y_bottom_right) {
  x <- rep(c(x_bottom_right-length, x_bottom_right), 2)
  y <- rep(c(y_bottom_right, y_bottom_right+length), each = 2)
  data.frame(x, y)[c(1, 3, 4, 2), ]
}
```

And off to build a tree:

```{r tree}
tree <- ggplot() +
  aes(x, y) +
  geom_polygon(data = square(0.5, 0.25, 0.25), fill = "brown") +
  geom_polygon(data = iso_triangle(10, 4, 0, 4.5), fill = "darkgreen") + 
  geom_polygon(data = iso_triangle(9, 4, 0, 4.5), fill = "lightgreen") + 
  geom_polygon(data = iso_triangle(8, 3, 0, 6), fill = "green") +
  geom_polygon(data = iso_triangle(7, 3, 0, 6), fill = "darkgreen") +
  geom_polygon(data = iso_triangle(6, 2.5, 0, 7), fill = "lightgreen") + 
  geom_polygon(data = iso_triangle(5, 2.5, 0, 7), fill = "green") + 
  geom_polygon(data = iso_triangle(4, 1.5, 0, 7.5), fill = "darkgreen") +
  geom_polygon(data = iso_triangle(3, 1.5, 0, 7.5), fill = "lightgreen") +
  geom_polygon(data = iso_triangle(2, 1.5, 0, 8), fill = "green")
tree
```

The tree looks good enough, but if you disagree or want some more realistic trees see [this challenge](https://github.com/r-link/christmas_tree_challenge).
Very funny to play with it (but I didn't manage to build my tree with it :/). 
Anyway, now we have the tree and we need to turn them to a glowing tree!!

# Decorating the tree

Now it is time to decorate and turn on the lights with the help of ggfx:

```{r decorating}
library("ggfx")
tree + 
  with_blur(geom_point(data =data.frame(x = 0, y = 8), size = 3, col = "yellow"), sigma = 2, stack = TRUE) + 
  with_blur(geom_point(data =data.frame(x = c(2, -3), y = c(1.5, 1)), size = 3, col = "darkgreen"), sigma = 2, stack = TRUE) + 
  with_blur(geom_point(data =data.frame(x = c(2, -3), y = 2.5+c(1, 1.2)), size = 3, col = "purple"), sigma = 2, stack = TRUE) + 
  with_blur(geom_point(data =data.frame(x = -5, y = 0.5), size = 3, col = "red"), sigma = 2, stack = TRUE) + 
  with_blur(geom_point(data =data.frame(x = 5, y = 0.5), size = 3, col = "orange"), sigma = 2, stack = TRUE) + 
  with_blur(geom_point(data =data.frame(x = -2.5, y = 4), size = 3, col = "red"), sigma = 2, stack = TRUE) + 
  with_blur(geom_point(data =data.frame(x = 3, y = 3.5), size = 3, col = "black"), sigma = 2, stack = TRUE) + 
  with_blur(geom_point(data =data.frame(x = c(-1, 1.25), y = c(5, 5.5)), size = 3, col = "orange"), sigma = 2, stack = TRUE) + 
  theme_void() +
  theme(plot.background = element_rect(fill = "black")) +
  NULL
```

And now we have our Christmas tree decorated!

Happy Christmas!!

### Reproducibility

<details>
```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info()
```
<details>
