---
title: Twitter bot
author: Lluís Revilla Sancho
date: '2019-08-13'
slug: twitter-bot
categories:
  - r
tags:
  - twitter
image:
  caption: ''
  focal_point: ''
editor_options: 
  chunk_output_type: console
---


I was talking with a friend about social networks when he mentioned that it 
wasn't worth his time to invest on podcasts. 
He said that I looked up his twitter account, that that's more useful for him. 
This reminded me that I haven't used these wonderful tools about twitter nor had I the motivation for analyzing time serie data.

This blogpost is my attempt to find how this user uses some kind of automated mechanism to publish.


```{r include=FALSE}
user <- "josemariasiota"
```

```{r}
library("rtweet")
user_tweets <- get_timeline(user, n = 180000, type = "mixed", 
                            include_rts = TRUE)
```

Now that we have the tweets we can look if he is a bot:
```{r}
library("tweetbotornot") # from mkearney/tweetbotornot
# you might need to install this specific version of textfeatures:
# devtools::install_version('textfeatures', version='0.2.0')
botornot(user_tweets)
```
It gives a very high probability.

We can visualize them with:
```{r}
library("ggplot2")
ts_plot(user_tweets, "weeks") +
  theme_bw() +
  labs(title = "Tweets by @josemariasiota",
       subtitle = "Grouped by week", x = NULL, y = "tweets")
```

We can group the tweets by the source of them, if interactive or using some other service:

```{r count}
library("dplyr")
count(user_tweets, source, sort = TRUE)
user <- user_tweets %>% 
  mutate(source = case_when(
    grepl(" for | on | Web ", source) ~ "direct",
    TRUE ~ source
  ))

user %>% 
  count(source, sort = TRUE)


user <- user %>% 
  mutate(reply = case_when(
    is.na(reply_to_status_id) ~  "content?",
    TRUE ~ "reply"))
user %>% 
  count(reply, source, sort = TRUE)
library("stringr")
user <- user %>% 
  mutate(link = str_extract(text, "https?://.+\\b"),
         n_link = str_count(text, "https?://"),
         n_users = str_count(text, "@[:alnum:]+\\b"),
         n_hashtags = str_count(text, "#[:alnum:]+\\b"),
         via = str_count(text, "\\bvia\\b"))
user %>% count(n_link, reply, sort = TRUE)

user %>% 
  group_by(lang, source) %>% 
  summarise(n = n(), n_link = sum(n_link), n_users = sum(n_users), n_hashtags = sum(n_hashtags)) %>% 
  arrange(-n) %>% 
  ggplot() +
  geom_point(aes(lang, source, size = n)) +
  theme_bw()
```

We can see that depending on the service there are some languages that are not used.

We can visualize the tweets as they happen with:
```{r}
user %>% 
  mutate(hms = hms::as_hms(created_at),
         d = as.Date(created_at)) %>% 
  ggplot(aes(d, hms, col = source, shape = reply)) +
  geom_point() +
  theme_bw() +
  labs(y = "Hour", x = "Date", title = "Tweets") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y", 
               expand = c(0.01, 0)) +
  scale_y_time(labels = function(x) strftime(x, "%H"),
               breaks = hms::hms(seq(0, 24, 1)*60*60), expand = c(0.01, 0))
```

We can clearly see a change on the end of 2016, I will focus on that point forward.

A package that got my attention on twitter was [`anomalize`](https://cran.r-project.org/package=anomalize) which search for anomalies on time series of data. I hope that using this algorithm it will find when the data is not automated

```{r}
library("anomalize")
```

The excellent guide at their [website](https://business-science.github.io/anomalize/) is easy to understand and follow


```{r}
user <- user %>% 
  filter(created_at > as.Date("2016-11-01")) %>% 
  arrange(created_at) %>% 
  time_decompose(created_at, method = "stl", merge = TRUE, message = TRUE) 
user %>% 
  filter(created_at > as.Date("2016-11-01")) %>% 
  anomalize(remainder, method = "iqr") %>%
  time_recompose() %>%
  # Anomaly Visualization
  plot_anomalies(time_recomposed = TRUE, ncol = 3, alpha_dots = 0.25) +
  labs(title = "User anomalies", 
       subtitle = "STL + IQR Methods", 
       x = "Time") 
user %>% 
  filter(created_at > as.Date("2016-11-01")) %>% 
  anomalize(remainder, method = "iqr") %>%
  plot_anomaly_decomposition() +
  labs(title = "Decomposition of Anomalized Lubridate Downloads")
```

We can clearly see some tendencies on the tweeting so it is automated, since then. We can further check it with:


```{r}
user %>% 
  filter(created_at > as.Date("2016-11-01")) %>% 
  botornot()
```

