---
title: CRAN dependencies
author: Lluís Revilla Sancho
date: '2020-01-25'
slug: cran-dependencies
categories:
  - r
tags:
  - r
  - CRAN
authors: []
description: ''
editor_options:
  chunk_output_type: console
featured: no
image:
  caption: ''
  focal_point: ''
subtitle: ''
summary: ''
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

# New policy in CRAN

With the new policy of a maximum of 20 packages in Imports. 
Let's see how many dependencies has each package on CRAN and Bioconductor:

# CRAN

```{r cran}
ap <- available.packages()
dp <- tools::package_dependencies(rownames(ap), db = ap, which = "Imports", 
                                  recursive = FALSE)
dp_n <- lengths(dp)
tb_dp <- sort(table(dp_n), decreasing = TRUE)
barplot(tb_dp)
```


Most of the packages has 0 dependencies and just `r sum(dp_n >= 20)` has 20 dependencies or more:

```{r names}
names(dp_n)[dp_n >= 20]
```

# Bioconductor

The interesting part is discovering how to add the repository. 
The trick is to make us of `BiocManager::repositories()`:

```{r BioC}
BioC_software <- BiocManager::repositories()["BioCsoft"]
bp <- available.packages(contriburl = contrib.url(BioC_software))
dp_BioC <- tools::package_dependencies(rownames(bp), db = bp, which = "Imports", 
                                  recursive = FALSE)
dp_BioC_n <- lengths(dp_BioC)
tb_dp_BioC <- sort(table(dp_BioC_n), decreasing = TRUE)
barplot(tb_dp_BioC)
```

It seems that software packages on Bioconductor tend to have more dependencies than on CRAN. 
If this policy would be implemented on Bioconductor it would affect `r sum(dp_BioC_n >= 20)` packages:

```{r names2}
names(dp_BioC_n)[dp_BioC_n >= 20]
```

### Reproducibility

<details>
```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info()
```
</details>
