---
title: R weekly new editor
author: Lluís Revilla Sancho
date: '2020-02-13'
slug: r-weekly-new-editor
categories:
  - r
tags:
  - r
  - community
  - Rweekly
  - GitHub
  - twitter
authors: []
description: ''
editor_options:
  chunk_output_type: console
featured: no
image:
  caption: ''
  focal_point: ''
subtitle: ''
summary: 'Checking who can be Rweekly.org editor with R'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

[Rweekly](https://rweekly.org) is looking for [new editors](https://docs.google.com/forms/d/e/1FAIpQLSet2Tq_mWWOVsKWxGOSoUg8DzCPlW2-nxIFOSkkRvlUFxQFLw/viewform). But they need to have submitted "at least 6 PRs on R Weekly". If you submitted something through [the webpage](https://rweekly.org/submit) you also can apply. But I'll look at how many people has submitted pull requests (PR) through GitHub at the repo rweekly/rewekly. 

# GH

So the GH package is good for this, but we need to know the API of Github. After a quick search I found the end point of the API:

```{r}
library("gh")
PR <- gh("GET /search/issues?q=repo:rweekly/rweekly.org+is:pr+is:merged&per_page=100") # Copied from https://developer.github.com/v3/pulls/
PR$total_count
```

We know that there have been 552, we'll need `r ceiling(PR$total_count/100)` calls to the appy, because it only return 30 values on each call. 

This time we'll use copy and paste for a quick solution:

```{r pr_calls}
PR2 <- gh("GET /search/issues?q=repo:rweekly/rweekly.org+is:pr+is:merged&per_page=100&page=2")
PR3 <- gh("GET /search/issues?q=repo:rweekly/rweekly.org+is:pr+is:merged&per_page=100&page=3")
PR4 <- gh("GET /search/issues?q=repo:rweekly/rweekly.org+is:pr+is:merged&per_page=100&page=4")
PR5 <- gh("GET /search/issues?q=repo:rweekly/rweekly.org+is:pr+is:merged&per_page=100&page=5")
PR6 <- gh("GET /search/issues?q=repo:rweekly/rweekly.org+is:pr+is:merged&per_page=100&page=6")
```
Now that we have the data we need to retrive the user names:

```{r users}
data <- list(PR, PR2, PR3, PR4, PR5, PR6)

users <- lapply(data, function(x) {
  vapply(x$items, function(y) {y$user$login}, character(1L))
})
users <- sort(unlist(users))
```

We know now that `r length(unique(users))` has contributed through PR.
Which of them are done by at the same people? 

```{r barplot}
ts <- sort(table(users), decreasing = TRUE)
par(mar=c(8,3,3,0))
barplot(ts, las = 2, border = "gray",
        main = "Contributors to Rweekly.org")
```

So we have `r sum(ts >=6)` contributors which are ellegible, less if we remove the current editors:

```{r ellegible_users}
names(ts)[ts >= 6]
```

### Reproducibility

<details>
```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info()
```
</details>
