---
title: 'Bioconductor submissions: reviews'
author: Lluís Revilla Sancho
date: '2020-07-09'
slug: bioconductor-submissions-reviews
categories:
  - Bioconductor
  - r
tags:
  - Bioconductor
  - r
authors: []
description: ''
editor_options:
  chunk_output_type: console
featured: no
draft: true
image:
  caption: ''
  focal_point: ''
subtitle: ''
summary: ''
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

First post  is on [Bioconductor submissions](https://llrs.dev/2020/06/bioconductor-submissions/) raised some questions comments but at the time I didn't have a good way to answer them:

- issues get closed after they got assigned a reviewer and before the reviewer actually gets a chance to start the review.
- issues assigned to multiple people or reviews that switched reviewers

The common problem is that I didn't have the information available. To retrieve the information from github I developed a [new package](https://llrs.dev/2020/06/social-github/) which . 

### Reproducibility

<details>
```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info()
```
</details>
