---
title: Fast correlations
author: Lluís Revilla Sancho
date: '2020-01-08'
slug: fast-correlations
categories:
  - r
tags:
  - r
authors: []
description: ''
editor_options:
  chunk_output_type: console
featured: no
image:
  caption: ''
  focal_point: ''
subtitle: ''
summary: ''
draft: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```


## Correlations

One of the few methods that are commonly done are correlations. 
I found several implementations of Pearson correlations, 
and I was curious to know if which is the fastest one.

## Implementations

So far the implementations I found are:

```{r eval = FALSE}
stats::cor
WGCNA::cor
miRcomb::cor
coop::pcor
HiClimR::fastCor
```

Most of them are on CRAN and one is at github.

# Dependencies

If we are only interested on the correlation most of the dependencies that a package would bring aren't rellevant.
So how many dependencies have each of those packages?

```{r dep}
ip <- installed.packages()
dp <- tools::package_dependencies(c("stats", "WGCNA", "miRcomb", "coop", "HiClimR"), 
                            db = ip, which = "Imports")
sort(lengths(dp), decreasing = TRUE)
```


# Speed

```{r bench}
library("bench")
x <- runif(50)
y <- runif(50)
stats_cor <- function() {
  stats::cor(x, y, method = "pearson")
}

WGCNA_cor <- function() {
  WGCNA::cor(x, y, method = "pearson")[1, 1]
}

coop_cor <- function() {
  coop::pcor(x, y)
}
HiClimR_cor <- function() {
  HiClimR::fastCor(matrix(c(x, y), nrow = 50, ncol = 2), upperTri = FALSE, verbose = FALSE)[2, 1]
}

bm <- mark(stats_cor(), WGCNA_cor(), coop_cor(), HiClimR_cor(), iterations = 10000)
bm
```

```{r plot}
plot(bm) + ggplot2::theme_bw()
```

So we can see that for this basic comparison stats is king followed by coop, WGCNA, and HiClimR.

It might be that these other functions are optimized for matrices so lets see it again


```{r mat}
x <- matrix(runif(100), ncol = 10, nrow = 10)
y <- matrix(runif(100), ncol = 10, nrow = 10)
stats_cor2 <- function() {
  stats::cor(x, x, method = "pearson")
}

WGCNA_cor2 <- function() {
  WGCNA::cor(x, x, method = "pearson")
}

coop_cor2 <- function() {
  coop::pcor(x)
}
HiClimR_cor2 <- function() {
  HiClimR::fastCor(x, upperTri = FALSE, verbose = FALSE)
}

bm2 <- mark(stats_cor2(), WGCNA_cor2(), coop_cor2(), HiClimR_cor2(), iterations = 10000)
bm2
```


```{r plot2}
plot(bm2) + ggplot2::theme_bw()
```

Here we can see that coop takes the lead and stats is the second fastest followed by WGCNA and HiClimR.

### References

```{r bibsetup, echo=FALSE, message=FALSE, warning=FALSE}
## Load knitcitations with a clean bibliography
library('knitcitations')
cleanbib()
cite_options(hyperlink = 'to.doc', citation_format = 'text', style = 'html')
pi <- sessioninfo::package_info()
packages <- c(pi$package[pi$attached], 'knitcitations')
l <- sapply(packages, function(x){citation(x)[1]}, simplify = FALSE)
bib <- c(l, 'blogdown' = citation('blogdown')[2])
```

```{r results = 'asis', echo = FALSE, cache = FALSE}
bibliography(style = 'html')
```

### Reproducibility

<details>
```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info()
```
</details>
