---
title: 'Bioconductor submissions: reviews'
author: Lluís Revilla Sancho
date: '2020-07-09'
slug: bioconductor-submissions-reviews
categories:
  - Bioconductor
  - r
tags:
  - Bioconductor
  - r
authors: []
description: 'Looking in detail to Bioconductor sumbissions: interactions between bots, reviewers and the community.'
editor_options:
  chunk_output_type: console
featured: no
draft: true
image:
  caption: ''
  focal_point: ''
subtitle: 'The second part of Bioconductor submissions'
summary: 'Looking in detail to Bioconductor sumbissions: interactions between bots, reviewers and the community.'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

First post  is on [Bioconductor submissions](https://llrs.dev/2020/06/bioconductor-submissions/) raised some questions comments but at the time I didn't have a good way to answer them:

- issues get closed after they got assigned a reviewer and before the reviewer actually gets a chance to start the review.
- issues assigned to multiple people or issues that switched reviewers

The common problem is that I didn't have the information available. 
On the previous post I only gathered the information of the state of the issues at that moment. 
This excluded label changes, reviewer changes, renaming the issues, who commented on the issues and many more things.
To retrieve these information from github I developed a [new package](https://llrs.dev/2020/06/social-github/) which downloads it from [Github](https://github.com) to make the analysis possible.

```{r}
library("socialGH")
repo <- "Bioconductor/Contributions"
gi <- get_issues(repo)
i <- unique(gi$id)
gt <- lapply(i, get_timelines, repository = repo)
gt2 <- do.call(rbind, gt)
saveRDS(gt, "static/20200711_timelines_Bioconductor_contributions.RDS")
```

```{r}
library("tidyverse")
theme_set(theme_minimal())
gi2 <- gi %>% 
  select(-n_comments) %>% 
  mutate(actor = poster, event = "created")

g <- rbind(gi2[, colnames(gt2)], gt2)
g2 <- g %>% 
  arrange(id, created) %>% 
  group_by(id) %>% 
  mutate(event_n = 1:n(),
         event = unlist(event, FALSE, FALSE),
         state = ifelse(event_n == 1, "opened", state))
```

```{r}
g2 %>% 
  group_by(id) %>% 
  filter(event_n == max(event_n)) %>% 
  ungroup() %>% 
  ggplot() +
  geom_histogram(aes(event_n))
g2 %>% 
  ggplot() +
  geom_point(aes(event_n, id, col = event))
g2 %>% 
  group_by(id) %>% 
  summarise(open_time = max(created) - min(created),
            n = max(event_n), id = unique(id)) %>% 
  ungroup() %>% 
  ggplot() +
  geom_point(aes(n, open_time, col = id)) +
  guides(col = FALSE)
g2 %>% 
  mutate(assigned = map(assignee, function(x) {
    ifelse(!is.null(unlist(x, FALSE, FALSE)), 
                  getElement(x, "user"), 
                  NA_character_)
  })
           ) %>% 
  group_by(id) %>% 
  filter(any(!is.na(assigned))) %>% 
  head()

k <- g2 %>% mutate(k == map2(actor, poster, function(x, y) {
  print(is(x))v
  print(is(y))
  nrow(x) == nrow(y)
})) %>% pull(k)

```


### Reproducibility

<details>
```{r reproducibility, echo = FALSE}
## Reproducibility info
options(width = 120)
sessioninfo::session_info()
```
</details>
